---
title: "Mar 7 2018"
output: html_notebook
---


# Multivariate ANOVA

##  Multivariate analysis of variance

  
  -  Standard ANOVA has just one response variable.
  -  What if you have more than one response?
  -  Try an ANOVA on each response separately.
  -  But might miss some kinds of interesting dependence between the responses that distinguish the groups.
  
  


##  Packages
  
```{r }
library(car)
library(tidyverse)
```   
  


##  Small example

  
  -  Measure yield and seed weight of plants grown under 2 conditions: low and high amounts of fertilizer.
  -  Data (fertilizer, yield, seed weight):

 
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/manova1.txt"
hilo=read_delim(my_url," ")
```

  -  2 responses, yield and seed weight.
  
  


##  The data
  
```{r }
hilo
```   
  


##  Boxplot for yield for each fertilizer group

 
```{r ferto}
ggplot(hilo,aes(x=fertilizer,y=yield))+geom_boxplot()
```
  
Yields overlap for fertilizer groups.
  


##  Boxplot for weight for each fertilizer group

 
```{r }
ggplot(hilo,aes(x=fertilizer,y=weight))+geom_boxplot()
``` 

Weights overlap for fertilizer groups.
  


##  ANOVAs for yield and weight


```{r }
hilo.y=aov(yield~fertilizer,data=hilo)
summary(hilo.y)
hilo.w=aov(weight~fertilizer,data=hilo)
summary(hilo.w)
```

Neither response depends significantly on fertilizer. But...
  


##  Plotting both responses at once

Have two response variables (not more), so can plot the
response variables against *each other*, labelling points by
which fertilizer group they're from.

```{r }
g=ggplot(hilo,aes(x=yield,y=weight,
    colour=fertilizer))+geom_point(size=5)
g
``` 

Want line through points $(31,14)$ and $(38,10)$ (why? Later):

```{r}
d=tribble(
    ~line_x, ~line_y,
    31, 14,
    38, 10)
d
g=g+geom_line(data=d,aes(x=line_x,y=line_y,
  colour=NULL))
``` 


Fitting regression line through points in `d`. 
Adding to previous `ggplot`, so `geom\_smooth` 
inherits `colour` from first one. This data frame has
no `fertilizer` (previous `colour`), so have to unset.
  

  


##  The plot
  
 
```{r }
g
```
  
  


##  MANOVA
  



  
  -  High-fertilizer plants have both yield and weight high.
  -  True even though no sig difference in yield or weight individually.
  -  Drew line separating highs from lows on plot.

 



##  MANOVA finds multivariate differences
  
  
  -  Is difference found by diagonal line significant? MANOVA finds out.


```{r size="footnotesize"}
response=with(hilo,cbind(yield,weight))
response
hilo.1=manova(response~fertilizer,data=hilo)
summary(hilo.1)
```
    

-  Yes! Difference between groups is *diagonally*, not just up/down
(weight) or left-right (yield). The *yield-weight combination* matters.

  



##  Strategy


-  Create new response variable by gluing together columns of
  responses, using `cbind`.
-  Use `manova` with new response, looks like `lm` otherwise.
-  With more than 2 responses, cannot draw graph. What then?
-  If MANOVA test significant, cannot use Tukey. What then?
-  Use **discriminant analysis** (of which more later).




##  Another way to do MANOVA

  
Install (once) and load package `car`:
  
```{r eval=F}
library(car)
```


  
  


## Another way...

```{r size="footnotesize"}
hilo.2.lm=lm(response~fertilizer,data=hilo)
hilo.2=Manova(hilo.2.lm)
hilo.2
```   

    
-  Same result as small-m `manova`.
-  `Manova` will also do *repeated measures*, coming up
  later.
  



## Another example: peanuts

  
  -   Three different varieties
of peanuts (mysteriously, 5, 6 and 8) planted in two different
locations.
-  Three response variables: `y`, `smk` and
`w`.
  

 
```{r message=F,size="small"}
peanuts.orig=read_delim("peanuts.txt"," ")
peanuts.orig
```
    
    


## Setup for analysis

```{r }
peanuts = peanuts.orig %>%
  mutate(location=factor(location),
         variety=factor(variety)) 
response=with(peanuts,cbind(y,smk,w))
response
``` 

  


## Analysis (using `Manova`)

```{r size="scriptsize"}
peanuts.1=lm(response~location*variety,data=peanuts)
peanuts.2=Manova(peanuts.1)
peanuts.2
``` 


-  Interaction not quite significant.
-  Weak dependence of `(y,smk,w)` on the location-variety *combination.*
-  Understanding that dependence beyond our scope right now.
- Come back to this with discriminant analysis.



  

# Repeated measures by profile analysis


## Repeated measures by profile analysis

  
  -  More than one response *measurement* for each subject. Might be
    
    -  measurements of the same thing at different times
    -  measurements of different but related things
    
  -  Generalization of matched pairs (``matched triples'', etc.).
  -  Variation: each subject does several different treatments at different times (called *crossover design*).
  -  Expect measurements on same subject to be correlated, so
    assumptions of independence will fail.
  -  Called *repeated measures*. Different approaches, but 
 *profile analysis* uses `Manova` (set up right way).
  -  Another approach uses *mixed models* (random effects).
  


## Packages
  
```{r }
library(car)
library(tidyverse)
```   
  



## Example: histamine in dogs
  
  
  -  8 dogs take part in experiment.
  -  Dogs randomized to one of 2 different drugs.
  -  Response: log of blood concentration of histamine 0, 1, 3 and 5 minutes after taking drug. (Repeated measures.)
  -  Data in `dogs.txt`, column-aligned.
  



## Read in data
  
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/dogs.txt"
dogs=read_table(my_url)
```   
  


## Setting things up

 
```{r}
dogs
```

```{r}
response=with(dogs,cbind(lh0,lh1,lh3,lh5))
response
dogs.lm=lm(response~drug,data=dogs)
```
  
    


## The repeated measures MANOVA

Get list of response variable names; we call them `times`. Save
in data frame.


 
```{r }
class(response)
times=colnames(response)
times
```

```{r}
times.df=data.frame(times)
times.df
dogs.manova=Manova(dogs.lm,idata=times.df,
     idesign=~times)
dogs.manova
```




Interaction significant. Pattern of response over time different
for the two drugs.


## Wide and long format

  
  -  Want to investigate interaction.
  -  But data frame has several observations per line (``wide format''):
 
```{r}
dogs
```
    
  -  Plotting works with data in ``long format'':
    one response per line.
  -  The responses are log-histamine at different times, labelled
    `lh`-something. Call them all `lh` and put them in
    one column, with the time they belong to labelled.
  
  



## Running `gather`, try 1
  

```{r size="footnotesize"}
dogs %>% gather(time,lh,lh0:lh5) 
```   
  


## Getting the times
  
Not quite right: for the times, we want just the numbers, not the
letters `lh` every time. Want new variable
containing just number in `time`:
`parse\_number`. 

```{r }
dogs %>% gather(timex,lh,lh0:lh5) %>% 
    mutate(time=parse_number(timex))

``` 



## What I did differently}
  
  
  -  I realized that `gather` was going to produce something
    like `lh1`, which I needed to do something further with, so
    this time I gave it a temporary name `timex`.
  -  This enabled me to use the name `time` for the actual
    numeric time.
  -  This works now, so next save into a new data frame `dogs.long`.
  
  


## Saving the results
  

```{r }
dogs.long = dogs %>% gather(timex,lh,lh0:lh5) %>% 
    mutate(time=parse_number(timex))
``` 

This says:


-  Take data frame dogs, and then:
-  Combine the columns `lh0` through `lh5` into one
  column called `lh`, with the column that each `lh`
  value originally came from labelled by `timex`, and then:
-  Pull out numeric values in `timex`, saving in `time` and then:
-  save the result in a data frame `dogs.long`.

## Spaghetti plot   

```{r}
ggplot(dogs.long,aes(x=time,y=lh,colour=drug,group=dog))+
  geom_point()+geom_line()
```



## Interaction plot
  
```{r }
ggplot(dogs.long,aes(x=time,y=lh,
                     colour=drug,group=drug))+
  stat_summary(fun.y=mean,geom="point")+
  stat_summary(fun.y=mean,geom="line")
```   
  



## Comments
  


-  Plot mean `lh` value at each time, joining points on same
  drug by lines.
-  drugs same at time 0
-  after that, Trimethaphan higher than Morphine.
-  Effect of drug not consistent over time: significant interaction.






## Take out time zero

  
  -  Lines on interaction plot would then be parallel, and so interaction should
no longer be significant.
-  Go back to original ``wide'' `dogs` data frame.
  
  

 
```{r }
response=with(dogs,cbind(lh1,lh3,lh5)) # excluding time zero
dogs.lm=lm(response~drug,data=dogs)
times=colnames(response)
times.df=data.frame(times)
dogs.manova=Manova(dogs.lm,idata=times.df,
                   idesign=~times)
``` 




## Results and comments

 
```{r}
dogs.manova
``` 


-  Correct: interaction no longer significant.
-  Significant effect of time.
-  Drug effect not quite significant (some variety among dogs
  within drug).

  


## Is the non-significant drug effect reasonable?
  
  
  -  Plot *actual data*: `lh` against `days`,
    labelling observations by drug: ``spaghetti plot''.
  -  Uses long data frame (confusing, yes I know):
 

-  Plot (time,lh) points coloured  by drug
-  and connecting measurements for each *dog* by lines.

  
-  This time, we want `group=dog` (want the measurements for each
*dog* joined by lines), but `colour=drug`:
  
```{r}
g=ggplot(dogs.long,aes(x=time,y=lh,
    colour=drug,group=dog)) +
  geom_point()+geom_line()
```   

  


## The spaghetti plot
  
```{r}
g
```   
  


## Comments
  
  
  -  For each dog over time, there is a strong increase and gradual
    decrease in log-histamine. This
    explains the significant time effect.
  -  The pattern is more or less the same for each dog, regardless
    of drug. This explains the non-significant interaction.
  -  Most of the trimethaphan dogs (blue) have higher log-histamine
    throughout (time 1 and after), and some of the morphine dogs have
    lower.
  -  *But* two of the morphine dogs have log-histamine
    profiles like the trimethaphan dogs. This ambiguity is probably
    why the `drug` effect is not quite significant.
  
  
## Mixed models

- Another way to fit repeated measures
- Subjects (on whom repeated measures taken) are *random sample of all possible subjects* (random effects)
- Times and treatments are *the only ones we care about* (fixed effects)
- Use package `lme4` function `lmer` (like `lm` in some ways)
- Uses long-format "tidy" data

## Fitting the model

```{r}
library(lme4)
dogs.long
dogs.2=lmer(lh~drug*time+(1|dog),data=dogs.long)
```

- note specification of random effect: each dog has "random intercept" that moves log-histamine up or down for that dog over all times


## What can we drop?

- using `drop1`:

```{r}
drop1(dogs.2,test="Chisq")
```

- Interaction again not significant, but P-value smaller than before

## Re-fit without interaction

```{r}
dogs.3=update(dogs.2,.~.-drug:time)
drop1(dogs.3,test="Chisq")
```

- This time neither drug nor (surprisingly) time is significant.
- MANOVA and `lmer` methods won't agree, but both valid ways to approach problem.
 
## The exercise data
  
  
  -  30 people took part in an exercise study.
  -  Each subject was
    randomly assigned to one of two diets (``low fat'' or ``non-low
    fat'') and to one of three exercise programs (``at rest'',
    ``walking'', ``running'').
  -  There are $2\times3 = 6$ experimental treatments, and thus
    each one is replicated $30/6=5$ times.
  -  Nothing unusual so far.
  -  However, each subject had their pulse rate measured at three
    different times (1, 15 and 30 minutes after starting their
    exercise), so have repeated measures.
  
  


## Reading the data
  
  Separated by *tabs*:  
  
```{r }
my_url="http://www.utsc.utoronto.ca/~butler/d29/exercise.txt"
exercise.long=read_tsv(my_url)
```   
  
  


## The data

```{r}
exercise.long
```   


-  This is ``long format'', which is usually what we want.
-  But for repeated measures analysis, we want *wide* format!
-  ``undo'' gather: `spread`.

  


## Making wide format
  
  
  -  Spread needs three things: a data frame, a column that is
    going to be split, and the column to make the values out of:
    
```{r size="footnotesize"}
exercise.wide=exercise.long %>% spread(time,pulse)
exercise.wide 
```
-  See how we would normally `gather` \texttt{min01, min15,
    min30} into one column called `pulse` labelled by the
  number of minutes? But `Manova` needs it the other way.
  
  


## Setting up the repeated-measures analysis
  
  
  -  Make a response variable consisting of `min01, min15, min30`:
```{r }
response=with(exercise.wide,
           cbind(min01, min15, min30))
```
-  Predict that from `diet` and `exertype` and
  interaction using `lm`:
```{r }
exercise.1=lm(response~diet*exertype,
  data=exercise.wide)
response
```   

-  Run this through `Manova`:
```{r }
times=colnames(response)
times.df=data.frame(times)
exercise.2=Manova(exercise.1,idata=times.df,
                  idesign=~times)
```   
  
  


## Results
  
  

```{r }
exercise.2
```       




-  Three-way interaction significant, so cannot remove anything.
-  Pulse rate depends on diet and exercise type *combination*,
  and *that* is different for each time.

  


## Making some graphs
  
  
  -  Three-way  interactions are difficult to understand. To make
    an attempt, look at some graphs.
  -  Plot time trace of pulse rates for each individual, joined by
    lines, and make *separate* plots for each
    `diet-exertype` combo.
  -  `ggplot` again. Using *long* data frame:

```{r}
exercise.long
```

Try 1

Use exercise type as colour and diet as shape, grouping by subject id

```{r}
ggplot(exercise.long,aes(x=time,y=pulse,colour=exertype,shape=diet,group=id))+
  geom_point()+geom_line()
```

Confusing. How about separate *facets* for diet and exercise type?


```{r }
g=ggplot(exercise.long,aes(x=time,y=pulse,
  group=id))+geom_point()+geom_line()+
  facet_grid(diet~exertype)
``` 

-  `facet_grid(diet~exertype)`: do a separate plot for each
  combination of diet and exercise type, with diets going down the
  page and exercise types going across. (Graphs are usually landscape,
  so have the factor `exertype` with more levels going across.)


  


 
## The graph(s)
  
```{r}
g
```   
  
  


## Comments on graphs
  
  
  -  For subjects who were at rest, no change in pulse rate over
    time, for both diet groups.
  -  For walking subjects, not much change in pulse rates over
    time. Maybe a small increase on average between 1 and 15 minutes.
  -  For both running groups, an overall increase in pulse rate
    over time, but the increase is stronger for the `lowfat`
    group.
  -  No consistent effect of diet over all exercise groups.
  -  No consistent effect of exercise type over both diet groups.
  -  No consistent effect of time over all diet-exercise type combos.
  
  


## ``Simple effects'' of diet for the subjects who ran
  
  
  -  Looks as if there is only any substantial time effect for the
    runners. For them, does diet have an effect?
  -  Pull out only the runners from the wide data:
```{r }
runners.wide = exercise.wide %>%
  filter(exertype=="running")
```     
-  Create response variable and do MANOVA. Some of this looks like
  before, but I have different data now:
  
```{r size="footnotesize"}
response=with(runners.wide,cbind(min01,min15,min30))
runners.1=lm(response~diet,data=runners.wide)
times=colnames(response)
times.df=data.frame(times)
runners.2=Manova(runners.1,idata=times.df,
                 idesign=~times)
```   
  
  


## Results`
  
```{r }
runners.2
```   

  
  
  -  The `diet` by `time` interaction is still
    significant (at $\alpha=0.05$): the effect of time on pulse rates is different for
    the two diets.
  -  At $\alpha=0.01$, the interaction is not significant, and then
    we have only two (very) significant main effects of `diet`
    and `time`. 
  
  


## How is the effect of diet different over time?
  
  
  -  Table of means. Only I need long data for this, so make it (in
    a pipeline):
    
```{r }
summ = runners.wide %>%
  gather(time,pulse,min01:min30) %>%
  group_by(time,diet) %>%
  summarize(mean_pulse=mean(pulse))
``` 

  
  


## Understanding diet-time interaction

  
    -  The summary:
```{r }
summ
```   
  -  Pulse rates at any given time higher for `lowfat` (diet
  effect), 
  -  Pulse rates increase over time of exercise (time effect),
    
  -  but the *amount by which pulse rate higher* for a diet depends on
  time: `diet` by `time` interaction.

  
  



## Interaction plot

-  We went to trouble of finding means by group, so making
  interaction plot is now mainly easy:
  
```{r }
ggplot(summ,aes(x=time,y=mean_pulse,colour=diet,
  group=diet))+geom_point()+geom_line()
```   

-  The lines are not parallel, so there is interaction between diet
  and time: the difference in mean pulse rate is getting bigger.



