---
title: "R Notebook"
output: html_notebook
---


# Discriminant analysis




## Discriminant analysis

  
  - ANOVA and MANOVA: predict a (counted/measured) response from group membership.
  - Discriminant analysis: predict group membership based on counted/measured variables.
  - Covers same ground as logistic regression (and its variations), but emphasis on classifying observed data into correct groups.
  - Does so by searching for linear combination of original variables that best separates data into groups (canonical variables).
  - Assumption here that groups are known (for data we have). If trying to "best separate" data into unknown groups, see *cluster analysis*.
  - Examples: revisit seed yield and weight data, peanut data,
    professions/activities data; remote-sensing data.
  




## Packages

  
```{r}
library(MASS)
library(tidyverse)
library(ggrepel)
```   

`ggrepel` allows labelling points on a plot so they don't
overwrite each other.


## About `select`
  
  
  - Both `dplyr` (in `tidyverse`) and `MASS`
    have a function called `select`, and *they do
      different things*.
  - How do you know which `select` is going to get called? 
  - With `library`, the one loaded *last* is visible,
    and others are not.
  - Thus we can access the `select` in `dplyr` but
    not the one in `MASS`. If we wanted that one, we'd have to
    say `MASS::select`.
  - This is why I loaded `MASS` before
    `tidyverse`. If I had done it the other way around, the
    `tidyverse` `select`, which I want to use, would have
    been the invisible one.  
  


## Example 1: seed yields and weights


```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/manova1.txt"
hilo=read_delim(my_url," ")
g=ggplot(hilo,aes(x=yield,y=weight,
  colour=fertilizer))+geom_point(size=4)
``` 


```{r}
g
```   


  
  
  Recall data from MANOVA: needed a multivariate analysis to find
  difference in seed yield and weight based on whether they were high
  or low fertilizer.


  




## Basic discriminant analysis

```{r}
hilo.1=lda(fertilizer~yield+weight,data=hilo)
``` 


- Uses `lda` from package MASS.
- "Predicting" group membership from measured variables.




## Output

  
```{r}
hilo.1
``` 



## Things to take from output
  
  
  - Group means: high-fertilizer plants have (slightly) higher
    mean yield and weight than low-fertilizer plants.
  
  - "Coefficients of linear discriminants": `LD1,
      LD2,`... are scores constructed from observed variables that
    best separate the groups.
    
    - Understand by pretending all variables standardized (mean 0,
      $+$ above mean, $-$ below mean). If yield and weight high (above
      average), contribute a $+$ to LD1 score, so LD1
      *negative*. If yield and weight low (think $-$), LD1 score
      *positive*.
    - High-fertilizer plants have higher yield and weight, thus
      negative LD1 score. Low-fertilizer plants have low yield and
      weight, thus positive LD1 score.
    - One LD1 score for each observation. Plot with actual groups.
    
  
  


## How many linear discriminants?
    
    
  - Number of variables
  - Number of groups *minus 1*
  - Smaller of these
  - Seed yield and weight: 2 variables, 2 groups,
      $\min(2,2-1)=1$. 
    
  


## Getting LD scores
  
Feed output from LDA into `predict`:

```{r}
hilo.pred=predict(hilo.1)
names(hilo.pred)
``` 

Component $x$ contains LD score(s), here in descending order:

```{r}
d = cbind(hilo,hilo.pred$x) %>% arrange(desc(LD1))
d
``` 

High fertilizer have yield and weight high, negative LD1 scores.
  


## Plotting LD1 scores
  
  With one LD score, plot against (true) groups, eg. boxplot:
  
```{r}
ggplot(d,aes(x=fertilizer,y=LD1))+geom_boxplot()
```   
  


## Potentially misleading
  
  
  - These are like regression slopes:
    
```{r}
hilo.1$scaling
```     

- Reflect change in LD1 score for 1-unit change in variables.
- But one-unit change in variables might not be comparable:
  
```{r}
summary(hilo)
```   

- Here, IQRs *identical*, so 1-unit change in each variable
  means same thing.
  
  


## What else is in `hilo.pred?`
  
```{r}
names(hilo.pred)
```


- `class`: predicted fertilizer level (based on values of
  `yield` and `weight`).
- `posterior`: predicted probability of being low or high
  fertilizer given `yield` and `weight`.
 
    


## Predictions and predicted groups
  
... based on `yield` and `weight`:

```{r}
cbind(hilo,predicted=hilo.pred$class)
table(obs=hilo$fertilizer,pred=hilo.pred$class)
``` 

Or:

```{r}
d2=tibble(obs=hilo$fertilizer,pred=hilo.pred$class)
d2
d2 %>% count(obs,pred)
```

or even

```{r}
d2=tibble(obs=hilo$fertilizer,pred=hilo.pred$class)
d2 %>% count(obs,pred) %>% 
  mutate(stat=ifelse(obs==pred,"correct","wrong"))

```




## Understanding the predicted groups
  
  
  - Each predicted fertilizer level is exactly same as observed
    one (perfect prediction).
  - Table shows no errors: all values on top-left to bottom-right
    diagonal. 
  


## Posterior probabilities
  
  show how clear-cut the classification decisions were:
  
```{r}
pp=round(hilo.pred$posterior,4)
d=cbind(hilo,hilo.pred$x,pp)
d
``` 

Only obs. 7 has any doubt: `yield` low for a high-fertilizer,
but high `weight` makes up for it.
  






## Example 2: the peanuts

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/peanuts.txt"
peanuts=read_delim(my_url," ")
peanuts 
``` 

Recall: `location` and `variety` both significant in
MANOVA. Make combo of them (over):

  


## Location-variety combos
  
```{r}
peanuts.combo = peanuts %>% unite(combo,variety,location) 
peanuts.combo
```  

  


## Discriminant analysis
  
```{r}
peanuts.1=lda(combo~y+smk+w,data=peanuts.combo)
peanuts.1
```   

Or, in pieces:

```{r}
peanuts.1$scaling
```

and

```{r}
peanuts.1$svd
```


- Now 3 LDs (3 variables, 6 groups, $\min(3,6-1)=3$).
- First: relationship of LDs to original variables. Look for
  coeffs far from zero: here,
  
  -   high `LD1` mainly high `w`
  or low `y`.
- high `LD2` mainly high `w`.
  
- `svd` values show relative importance of LDs:
  `LD1` much more important than `LD2`.



## Group means by variable
  
```{r}
peanuts.1$means
```   


- `5_2` clearly smallest on `y`, `smk`, near
  smallest on `w`
- `8_2` clearly biggest on `smk`, `w`, also
  largest on `y`
- `8_1` large on `w`, small on `y`.
- `scaling` links LDs with original variables,
  `means` links original variables with groups.
- Implies: link between groups and LDs.

  



## The predictions and misclassification
  
```{r}
peanuts.pred=predict(peanuts.1)
table(obs=peanuts.combo$combo,
      pred=peanuts.pred$class)
```   

Actually classified very well. Only one `6_2` classified as a
`5_1`, rest all correct.

Or: 


```{r}
dd=tibble(obs=peanuts.combo$combo,pred=peanuts.pred$class)
dd
dd %>% count(obs,pred)
```

or even

```{r}
dd %>% count(obs,pred) %>% 
  mutate(stat=ifelse(obs==pred,"correct","wrong"))
```



## Posterior probabilities

```{r}
pp=round(peanuts.pred$posterior,2)
peanuts.combo %>% select(-c(y,smk,w)) %>%
   cbind(.,pred=peanuts.pred$class,pp)
```   

*Some* doubt about which combo each plant belongs in, but not too
much. The one misclassified plant was a close call.



## Discriminant scores, again
  
  
  - How are discriminant scores related to original variables?
  - Construct data frame with original data and discriminant
    scores side by side:
```{r}
peanuts.1$scaling
lds=round(peanuts.pred$x,2)
mm=with(peanuts.combo,
  data.frame(combo,y,smk,w,lds))
```   
- LD1 positive if `w` large and/or `y` small.
- LD2 positive if `w` large.    
  - But, what if `y, smk, w` differ in spread?
    
  



## Discriminant scores for data

```{r}
mm
```   
  
  
- Obs. 5 and 6 have most negative `LD1`: large `y`,
  small `w`.
- Obs. 4 has most negative `LD2`: small `w`.

Plot:
```{r,fig.width=10}
ggplot(mm,aes(x=LD1,y=LD2,colour=combo,shape=combo))+geom_point()
  ```




## Predict typical LD1 scores
  
  First and third quartiles for three response variables (reading down):
  
```{r}
quartiles = peanuts %>% select(y:w) %>%
  map_df(quantile, c(0.25,0.75))
quartiles
new=with(quartiles,crossing(y,smk,w))
```   
  


## The combinations
  
```{r}
new
pp=predict(peanuts.1,new)
```   
  


## Predicted typical LD1 scores
  
```{r}
cbind(new,pp$x) %>% arrange(LD1)
```   


- Very negative LD1 score with large `y` and small
  `w`
- `smk` doesn't contribute much to LD1
- Very positive LD1 score with small `y` and large
  `w`.
- Same as we saw from Coefficients of Linear Discriminants.

  



  
## Plot LD1 vs. LD2, labelling by combo
  
```{r}
g=ggplot(mm,aes(x=LD1,y=LD2,colour=combo,
  label=combo))+geom_point()+
  geom_text_repel()+guides(colour=F) ; g
```   
  


## "Bi-plot" from `ggbiplot`
  
See installation notes below.  
  
```{r}
ggbiplot::ggbiplot(peanuts.1,
  groups=factor(peanuts.combo$combo))
``` 

## Installing `ggbiplot`
  
  
  - `ggbiplot` not on CRAN, so usual
    `install.packages` will not work.
  - Install package `devtools` first (once):
    
```{r, eval=F}
install.packages("devtools")
```     
  - Then install `ggbiplot` (once):
```{r, eval=FALSE}
library(devtools)
install_github("vqv/ggbiplot")
```     
  
I already have these installed, so I don't need to install them again.
For you, take the `eval=FALSE` out of the code chunk header (once) to install them (and then put it back).



## Cross-validation
  
  
  - So far, have predicted group membership from same data used to
    form the groups --- dishonest!
  - Better: *cross-validation*: form groups from all
    observations *except one*, then predict group membership for
    that left-out observation.
  - No longer cheating!
  - Illustrate with peanuts data again.
  
  


## Misclassifications
  
  - Fitting and prediction all in one go:
  
```{r}
peanuts.cv=lda(combo~y+smk+w,
  data=peanuts.combo,CV=T)
table(obs=peanuts.combo$combo,
      pred=peanuts.cv$class)
```   

- Some more misclassification this time.
  



## Repeat of LD plot
 
```{r}
g
```   
  


## Posterior probabilities
  
```{r}
pp=round(peanuts.cv$posterior,3)
data.frame(obs=peanuts.combo$combo,
           pred=peanuts.cv$class,pp)
```   
  


## Why more misclassification?
  
  
  - When predicting group membership for one observation, only
    uses the *other one* in that group.
  - So if two in a pair are far apart, or if two groups overlap,
    great potential for misclassification.
  - Groups `5_1` and `6_2` overlap.
  - `5_2` closest to `8_1`s looks more like an
    `8_1` than a `5_2` (other one far away).
  - `8_1`s relatively far apart and close to other things,
    so one appears to be a `5_2` and the other an `8_2`.
  
  



## Example 3: professions and leisure activities

  
  - 15 individuals from three different professions (politicians,
    administrators and belly dancers) each participate in four
    different leisure activities: reading, dancing, TV watching and
    skiing. After each activity they rate it on a 0--10 scale.
  - Some of the data:

```
bellydancer 7 10 6 5
bellydancer 8 9 5 7
bellydancer 5 10 5 8
politician 5 5 5 6
politician 4 5 6 5
admin 4 2 2 5
admin 7 1 2 4
admin 6 3 3 3
```
  - How can we best use the scores on the activities to predict a person's profession?
  - Or, what combination(s) of scores best separate data into profession groups?
  



## Discriminant analysis

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/profile.txt"
active=read_delim(my_url," ")
active.1=lda(job~reading+dance+tv+ski,data=active)
active.1$svd
active.1$scaling
```   


- Two discriminants, first fair bit more important than second.
- `LD1` depends (negatively) most on `dance`, a bit
  on `tv`.
- `LD2` depends mostly on `tv`.






## Misclassification
  
```{r}
active.pred=predict(active.1)
table(obs=active$job,pred=active.pred$class)
```   

Everyone correctly classified. Or:

```{r}
tibble(obs=active$job,pred=active.pred$class) %>% 
  count(obs,pred) %>% 
  mutate(stat=ifelse(obs==pred,"correct","wrong"))
```

  


## Plotting LDs
  
```{r}
mm=data.frame(job=active$job,active.pred$x,person=1:15)
g=ggplot(mm,aes(x=LD1,y=LD2,
    colour=job,label=job))+geom_point()+
    geom_text_repel()+guides(colour=F) ; g
```   
  


## Biplot
  
```{r}
ggbiplot::ggbiplot(active.1,groups=active$job)
```   
  


## Comments on plot
  
  
  - Groups well separated: bellydancers top left, administrators
    top right, politicians lower middle.
  - Bellydancers most negative on `LD1`: like dancing most.
  - Administrators most positive on `LD1`: like dancing least.
  - Politicians most negative on `LD2`: like TV-watching most.
  
  


## Plotting individual `persons`
  
Make `label` be identifier of person. Now need legend:

```{r}
ggplot(mm,aes(x=LD1,y=LD2,
    colour=job,label=person))+geom_point()+
    geom_text_repel()
```   
  
  


## Posterior probabilities

```{r}
pp=round(active.pred$posterior,3)
data.frame(obs=active$job,pred=active.pred$class,pp)
```   


Not much doubt.


## Cross-validating the jobs-activities data
  
Recall: no need for `predict`. Just pull out `class` and
make a table:  
  
```{r}
active.cv=lda(job~reading+dance+tv+ski,
  data=active,CV=T)
table(obs=active$job,pred=active.cv$class)
```   

Or:

```{r}
tibble(obs=active$job,pred=active.cv$class) %>% 
  count(obs,pred) %>% 
  mutate(is_correct=ifelse(obs==pred,"yes","no"))
```



This time one of the bellydancers was classified as a politician.
  


## and look at the posterior probabilities
  
picking out the ones where things are not (absolutely) certain:

```{r}
pp=round(active.cv$posterior,3)
data.frame(obs=active$job,pred=active.cv$class,pp) %>%
  mutate(max=pmax(admin,bellydancer,politician)) %>%
  filter(max<0.9995)
```   


- Bellydancer was "definitely" a politician!
- One of the administrators might have been a politician too.

  



## Why did things get misclassified?

  
```{r}
g
```       
  
  
    
    - Go back to plot of discriminant scores:
    - one bellydancer much closer to the politicians,
    - one administrator a bit closer to the politicians.
    
  
  
  

  



## Example 4: remote-sensing data

  
  - View 38 crops from air, measure 4 variables `x1-x4`.
  - Go back and record what each crop was.
  - Can we use the 4 variables to distinguish crops?
  


## Reading in
  
```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/remote-sensing.txt"
crops=read_table(my_url)
crops
```   
  



## Starting off: number of LDs
  
```{r}
crops.lda=lda(crop~x1+x2+x3+x4,data=crops)
crops.lda
``` 

```{r}
crops.lda$svd
```


- 4 LDs (four variables, six groups).
- 1st one important, maybe 2nd as well.

  



## Connecting original variables and LDs
  
```{r}
crops.lda$means
round(crops.lda$scaling,3)
```   


- Links groups to original variables to LDs.



## `LD1 and \texttt{LD2`}
  
```{r}
round(crops.lda$scaling,3)
``` 
%$

- `LD1` mostly `x1` (minus), so clover low on
  `LD1`, corn high.
- `LD2` `x3` (minus), `x2` (plus), so
  sugarbeets should be high on `LD2`.


  


## Predictions
  
  
  - Thus:
```{r}
crops.pred=predict(crops.lda)
table(obs=crops$crop,pred=crops.pred$class)
```   
- Not very good, eg. only 6 of 11 `Clover` classified correctly.
- Set up for plot:
  
```{r}
mm=data.frame(crop=crops$crop,crops.pred$x)
```   
    
  

  



## Plotting the LDs
  
```{r}
ggplot(mm,aes(x=LD1,y=LD2,colour=crop,shape=crop))+
  geom_point()
```   
  


## Biplot
  
```{r}
ggbiplot::ggbiplot(crops.lda,groups=crops$crop)
```   
  


\begin{frame}[figure]{Comments}
  
  
  - Corn high on LD1 (right).
  - Clover all over the place, but mostly low on LD1 (left).
  - Sugarbeets tend to be high on LD2.
  - Cotton tends to be low on LD2.
  - Very mixed up.
  
  


## Try removing Clover

    
```{r}
crops %>% filter(crop!="Clover") -> crops2
crops2.lda=lda(crop~x1+x2+x3+x4,data=crops2)
```   

- LDs for `crops2` will be different from before.
- Concentrate on plot and posterior probs.

```{r}
crops2.pred=predict(crops2.lda)
mm=data.frame(crop=crops2$crop,crops2.pred$x)
```   
  

  



## `lda output`
  
  
Different from before:

```{r}
crops2.lda
```  
  


## Plot

A bit more clustered:
  
```{r}
ggplot(mm,aes(x=LD1,y=LD2,colour=crop,shape=crop))+
  geom_point()
```   



## Biplot
  
```{r}
ggbiplot::ggbiplot(crops2.lda,groups=crops2$crop)
```   
  


## Quality of classification
  
```{r}
table(obs=crops2$crop,pred=crops2.pred$class)
```   

or:

```{r}
tibble(obs=crops2$crop,pred=crops2.pred$class) %>% 
  count(obs,pred) %>% 
  mutate(is_correct=ifelse(obs==pred,"yes","no"))

```


Better.
  


## Posterior probs, the wrong ones
 

```{r}
post=round(crops2.pred$posterior,3)
data.frame(obs=crops2$crop,pred=crops2.pred$class,post) %>%
  filter(obs!=pred)
```   
}


- These were the misclassified ones, but the posterior probability
  of being correct was not usually too low.
- The correctly-classified ones are sometimes more clear-cut, but not completely convincing:

```{r}
data.frame(obs=crops2$crop,pred=crops2.pred$class,post) %>%
  filter(obs==pred)
```



  


## MANOVA
  
Began discriminant analysis as a followup to MANOVA. Do our variables
significantly separate the crops (excluding Clover)?

```{r}
response=with(crops2,cbind(x1,x2,x3,x4))
crops2.manova=manova(response~crop,data=crops2)
summary(crops2.manova)
``` 

Yes, at least one of the crops differs (in means) from the others. So
it is worth doing this analysis.

Having said that, the P-value is not *very* small, so the separation of crops might not be very strong (as we saw).

We did this the wrong way around, though!
  


## The right way around
  
  
  - *First*, do a MANOVA to see whether any of the groups
    differ significantly on any of the variables.
  - *If the MANOVA is significant*, do a discriminant
    analysis in the hopes of understanding how the groups are different.
  - For remote-sensing data (without Clover):
    
    - LD1 a fair bit more important than LD2 (definitely ignore LD3).
    - LD1 depends mostly on `x1`, on which Cotton was high
      and Corn was low. 
    

  - Discriminant analysis in MANOVA plays the same kind of role
    that Tukey does in ANOVA.
  
  










