---
title: "Feb 7 2018"
output: html_notebook
---

## Packages

```{r}
# install.packages("survminer") from CRAN
```


```{r}
library(tidyverse)
library(survival)
library(survminer)
library(broom)
```

- `survival` does survival model-fitting
- `survminer` makes nice survival plots
- `broom` does tidy model output (computable with) 

## The dancing data

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/d29/dancing.txt"
dance=read_table(my_url)
dance
```

- Number of months observed dancing
- whether or not they quit (1=quit, 0=still dancing), 
- treatment=1: visit to dance competition
- age

## Create response variable for survival analysis

Called `mth` in notes, but I now prefer `y`:

```{r}
y=with(dance,Surv(Months,Quit))
y
```

`+` means "censored": event (quitting) not observed.

```{r}
print.default(y)
```


## Fit a Cox proportional-hazards model

```{r}
dance.1=coxph(y~Treatment+Age,data=dance)
summary(dance.1)
```

Both explanatory variables significant at $\alpha=0.10$ (not much data).

Keep both. (If one truly not significant, remove.)


## Model checking

Make residual plot using "martingale residuals":


```{r}
ggcoxdiagnostics(dance.1)+geom_smooth()
```

Looking for no pattern. Martingale residuals can be very negative, so won't be normal. I call this no pattern. Therefore model satisfactory. 

## Predicted survival

Make combinations of ages 20 and 40, treatment no or yes:

```{r}
treatments=c(0,1)
ages=c(20,40)
dance.new=crossing(Treatment=treatments,Age=ages)
dance.new
```

Make the predictions. `data=` is not needed for predictions, but *is* needed for plot below:

```{r}
s=survfit(dance.1,newdata=dance.new,data=dance)
summary(s)
```

Number of each `survival` column is number of row in `dance.new`:

```{r}
dance.new
```

but this is hard to see; a plot is much nicer:

```{r}
ggsurvplot(s,conf.int=F)
```

Identify "strata" with rows of `dance.new` (one coloured survival curve for each treatment-age combo). Eg. stratum 1 (red) is no treatment, age 20, stratum 4 (purple) is treatment, age 40.

Survival curve up and to the right (the purple one) is associated with the longest survival,
survival curve down and to left (red) associated with shortest survival.

What makes a woman likely to continue dancing longest? Being older, and going to the dance competition. 

Assess effect of age by comparing strata of different age but same treatment, 
- eg. strata 1 and 2, 
- red and green, 
- age 20 and 40. 

Green has better survival, so age 40
women tend to keep dancing for longer.

Assess effect of treatment by comparing strata of same age but different treatment, 
- eg strata 2 and 4, 
- no treatment and treatment, 
- green and purple.

Purple, treatment, associated with keeping dancing for longer.

## The lung cancer data set

```{r}
lung
```

or

```{r}
summary(lung)
```

Remove missings:

```{r}
cc=complete.cases(lung)
cc
lung.complete = lung %>% filter(cc)
```

```{r}
lung.complete
```

```{r}
summary(lung.complete)
```


## Construct response, and model in terms of everything else but `inst`:

```{r}
y=with(lung.complete,Surv(time,status==2))
lung.1=coxph(y~.-inst-time-status,
  data=lung.complete)
```

Summary:

```{r}
summary(lung.1)
```

Using `broom` ideas to pick things out of the output:

```{r}
glance(lung.1)
```

`p.value.log`, `p.value.sc` and `p.value.wald` all test for overall significance of the model. All small, so *something* predicts survival.

```{r}
tidy(lung.1) 
```

Some non-significant things there, with `meal.cal` having highest P-value. 

Can:

- remove one at a time using `summary` or `drop1` to see what to remove next
- remove several and then test (via `anova`)
- use `step` to do whole backward elimination

`step`:

```{r}
step.back=step(lung.1,direction="backward",test="Chisq")
```

Or take out a bunch and test:

```{r}
lung.2=update(lung.1,.~.-age-pat.karno-meal.cal)
tidy(lung.2) %>% select(term,p.value)
lung.3=update(lung.2,.~.-ph.karno-wt.loss)
tidy(lung.3) %>% select(term,estimate,p.value)
anova(lung.3,lung.1)
```

Different results. Go with the model `lung.3` with just `sex` and `ph.ecog`.

## Problems with the model?

```{r}
ggcoxdiagnostics(lung.3)+geom_smooth(se=F)
```

apparently OK.

## Predictions

The usual:

```{r}
sexes=c(1,2)
ph.ecogs=0:3
lung.new=crossing(sex=sexes,ph.ecog=ph.ecogs)
lung.new
s=survfit(lung.3,newdata=lung.new,data=lung.complete)
```

Plot:

```{r}
ggsurvplot(s,conf.int=F)
```

- connect coloured survival curves with strata and thus with rows of `lung.new`.
- longest survival time: stratum 5, females, `ph.ecog` 0
- shortest survival time: stratum 4, males, `ph.ecog` 3
- effect of gender is that females survive longer than males with same `ph.ecog` score (compare strata 1 and 5, 2 and 6, etc)
- effect of `ph.ecog` score is that survival longer if score *lower* (compare eg. strata 5 through 8)

## How can proportional hazards go wrong?

Invent data set where survival with age goes up and then down:

```{r}
age=seq(20,60,5)
survtime=c(10,12,11,21,15,20,8,9,11)
stat=c(1,1,1,1,0,1,1,1,1)
d=tibble(age,survtime,stat)
d
y=with(d,Surv(survtime,stat))
```

Fit model:

```{r}
y.1=coxph(y~age,data=d)
summary(y.1)
```

Check diagnostic plot:

```{r}
ggcoxdiagnostics(y.1)+geom_smooth()
```

Apparently a curve; suggests adding age-squared to model:

```{r}
y.2=coxph(y~age+I(age^2),data=d)
summary(y.2)
```

Age-squared is significant at $\alpha=0.10$. Diagnostics now?

```{r}
ggcoxdiagnostics(y.2)+geom_smooth(se=F)
```

Better.

```{r}
new=tibble(age=c(20,35,50,65))
s=survfit(y.2,new,data=d)
```

```{r}
ggsurvplot(s,conf.int=F)
new
```
