---
title: "2018-01-17"
output: html_notebook
---

## The curvy data

Read in and examine some other data:

```{r 20180117-1}
my_url="http://www.utsc.utoronto.ca/~butler/d29/curvy.txt"
curvy=read_delim(my_url," ")
ggplot(curvy,aes(x=xx,y=yy))+geom_point()
```

Fit a linear regression, and see what goes wrong:

```{r 20180117-2}
curvy.1=lm(yy~xx,data=curvy) ; summary(curvy.1)
ggplot(curvy.1,aes(x=.fitted,y=.resid))+geom_point()
```

Add squared term, either of 2 ways:

```{r 20180117-3}
curvy.2=lm(yy~xx+I(xx^2),data=curvy)
curvy.2a=update(curvy.1,.~.+I(xx^2))
summary(curvy.2)
```

Look at residuals from that:

```{r 20180117-4}
ggplot(curvy.2,aes(x=.fitted,y=.resid))+geom_point()
```

Better, though is that fan-out?


## the made-up data

```{r 20180117-5}
library(MASS)
my_url="http://www.utsc.utoronto.ca/~butler/d29/madeup.csv"
madeup=read_csv(my_url)
madeup
ggplot(madeup,aes(x=x,y=y))+geom_point()+
  geom_smooth()
```

## Box-Cox

```{r 20180117-6}
boxcox(y~x,data=madeup)
```

Use log of `y`:

```{r 20180117-7}
log.y=log(madeup$y) 
ggplot(madeup,aes(x=x,y=log.y))+geom_point()+
  geom_smooth()
```

Get rid of `MASS`:

```{r 20180117-8}
detach("package:MASS",unload=T)
```

## Multiple regression

```{r 20180117-9}
library(tidyverse)
my_url="http://www.utsc.utoronto.ca/~butler/d29/regressx.txt"
visits=read_delim(my_url," ")
visits
visits.1=lm(timedrs~phyheal+menheal+stress,
  data=visits)
summary(visits.1)
```

Another look at results

```{r 20180117-10}
library(broom)
tidy(visits.1)
```

Just `menheal`:

```{r 20180117-11}
visits.2=lm(timedrs~menheal,data=visits) 
summary(visits.2)
```

Why? Correlations:

```{r 20180117-12}
visits %>% select_if(is.numeric) %>% cor()
```

Residual plot (from 1st regression):

```{r 20180117-13}
ggplot(visits.1,aes(x=.fitted,y=.resid))+geom_point()
```

Normal quantile plot of residuals

```{r 20180117-14}
r=resid(visits.1)
qqnorm(r)
qqline(r)
```

Absolute residuals

```{r 20180117-15}
ggplot(visits.1,aes(x=.fitted,y=abs(.resid)))+
  geom_point()+geom_smooth()
```

Try log of visits (plus one, since visits could be zero):

```{r 20180117-16}
lgtime=with(visits,log(timedrs+1))
visits.3=lm(lgtime~phyheal+menheal+stress,
  data=visits)
summary(visits.3)
```

Residual plots from this:

```{r 20180117-17}
ggplot(visits.3,aes(x=.fitted,y=.resid))+
  geom_point()
r=residuals(visits.3); qqnorm(r); qqline(r)
ggplot(visits.3,aes(x=.fitted,y=abs(.resid)))+
  geom_point()+geom_smooth()
```

Box-Cox:

```{r 20180117-18}
tp=with(visits,timedrs+1)
library(MASS)
boxcox(tp~phyheal+menheal+stress,data=visits)
```

Zoom in:

```{r 20180117-19}
my.lambda=seq(-0.3,0.1,0.01)
my.lambda
boxcox(tp~phyheal+menheal+stress,lambda=my.lambda,
  data=visits)
```

Log is definitely good.

