
## Assessing assumptions

- Our $t$-tests assume normality of variable being tested
- but, Central Limit Theorem says that normality matters less if sample is "large"
- in practice "approximate normality" is enough, but how do we assess whether what we have is normal enough?
- so far, use histogram/boxplot and make a call, allowing for sample size.

## What actually has to be normal

- is: **sampling distribution of sample mean**
- the distribution of sample mean over *all possible samples*
- but we only have *one* sample!
- Idea: assume our sample is representative of the population, and draw samples from our sample (!), with replacement.
- This gives an idea of what different samples from the population might look like.
- Called *bootstrap*, after expression "to pull yourself up by your own bootstraps".

## Blue Jays attendances

```{r bootstrap-R-1, echo=FALSE, message=FALSE}
my_url <- url("http://ritsokiguess.site/datafiles/jays15-home.csv")
jays <- read_csv(my_url)
set.seed(457299)
```

```{r bootstrap-R-2}
jays$attendance
```

- A bootstrap sample:

```{r bootstrap-R-3}
s <- sample(jays$attendance, replace = TRUE)
s
```

## Getting mean of bootstrap sample

- A bootstrap sample is same size as original, but contains repeated values (eg. 15062) and missing ones (42917).
- We need the mean of our bootstrap sample:

```{r bootstrap-R-4}
mean(s)
```

- This is a little different from the mean of our actual sample:

```{r bootstrap-R-5}
mean(jays$attendance)
```

- Want a sense of how the sample mean might vary, if we were able to take repeated samples from our population.
- Idea: take lots of *bootstrap* samples, and see how *their* sample means vary.

## Setting up bootstrap sampling

- Begin by setting up a dataframe that contains a row for each bootstrap sample. I usually call this column `sim`. Do just 4 to get the idea:

```{r bootstrap-R-6}
tibble(sim = 1:4)
```

## Drawing the bootstrap samples

- Then set up to work one row at a time, and draw a bootstrap sample of the attendances in each row:

\small
```{r bootstrap-R-7}
tibble(sim = 1:4) %>% 
  rowwise() %>%
  mutate(sample = list(sample(jays$attendance, replace = TRUE))) 
```
\normalsize

- if you want to see all the sample values, add `unnest(sample)` to the end 
- Each row of our dataframe contains *all* of a bootstrap sample of 25 observations drawn with replacement from the attendances.

\normalsize

## Sample means

- Find the mean of each sample:

\footnotesize
```{r bootstrap-R-8}
tibble(sim = 1:4) %>% 
  rowwise() %>% 
  mutate(sample = list(sample(jays$attendance, replace = TRUE))) %>% 
  mutate(my_mean = mean(sample))
```
\normalsize

- These are (four simulated values of) the bootstrapped sampling distribution of the sample mean.

## Make a histogram of them

- rather pointless here, but to get the idea:

\footnotesize
```{r bootstrap-R-9}
tibble(sim = 1:4) %>% 
  rowwise() %>% 
  mutate(sample = list(sample(jays$attendance, replace = TRUE))) %>% 
  mutate(my_mean = mean(sample)) %>% 
  ggplot(aes(x = my_mean)) + geom_histogram(bins = 3) -> g
```
\normalsize

## The (pointless) histogram

```{r bootstrap-R-10}
g
```

## Now do again with a decent number of bootstrap samples

- say 1000, and put a decent number of bins on the histogram also:

\footnotesize
```{r bootstrap-R-11}
tibble(sim = 1:1000) %>% 
  rowwise() %>% 
  mutate(sample = list(sample(jays$attendance, replace = TRUE))) %>% 
  mutate(my_mean = mean(sample)) %>% 
  ggplot(aes(x = my_mean)) + geom_histogram(bins = 10) -> g
```
\normalsize

## The (better) histogram

```{r bootstrap-R-12}
g
```

## Comments

- This is very close to normal
- The bootstrap says that the sampling distribution of the sample mean is close to normal, even though the distribution of the data is not
- A sample size of 25 is big enough to overcome the skewness that we saw
- This is the Central Limit Theorem in practice
- It is surprisingly powerful.
- Thus, the $t$-test is actually perfectly good here.

## Comments on the code

- You might have been wondering about this:

\small
```{r bootstrap-R-13}
tibble(sim = 1:4) %>% 
  rowwise() %>% 
  mutate(sample = list(sample(jays$attendance, replace = TRUE)))
```
\normalsize

- how did we squeeze all 25 sample values into one cell?
  - sample is a so-called "list-column" that can contain anything.
- why did we have to put `list()` around the `sample()`?
  - because `sample` produces a collection of numbers, not just a single one
  - the `list()` signals this: "make a list-column of samples".
  
  
## Two samples

- Assumption: *both* samples are from a normal distribution.
- In practice, each sample is "normal enough" given its sample size, since Central Limit Theorem will help.
- Use bootstrap on each group independently, as above.

## Kids learning to read

```{r bootstrap-R-14, echo=FALSE, message=FALSE}
my_url <- url("http://ritsokiguess.site/datafiles/drp.txt")
kids <- read_delim(my_url," ")
```

```{r bootstrap-R-15}
ggplot(kids, aes(x=group, y=score)) + geom_boxplot()
```


## Getting just the control group 

- Use `filter` to select rows where something is true:

```{r bootstrap-R-16}
kids %>% filter(group=="c") -> controls
controls
```

## Bootstrap these

\footnotesize
```{r bootstrap-R-17}
tibble(sim = 1:1000) %>% 
  rowwise() %>% 
  mutate(sample = list(sample(controls$score, replace = TRUE))) %>% 
  mutate(my_mean = mean(sample)) %>% 
  ggplot(aes(x = my_mean)) + geom_histogram(bins = 10) -> g
```
\normalsize

## Plot

```{r bootstrap-R-18}
g
```

## ... and the treatment group:

\footnotesize
```{r bootstrap-R-19}
kids %>% filter(group=="t") -> treats
tibble(sim = 1:10000) %>% 
  rowwise() %>% 
  mutate(sample = list(sample(treats$score, replace = TRUE))) %>% 
  mutate(my_mean = mean(sample)) %>% 
  ggplot(aes(x = my_mean)) + geom_histogram(bins = 10) -> g
```
\normalsize

## Histogram

```{r bootstrap-R-20}
g
```

## Comments

- sampling distributions of sample means both look pretty normal
- as we thought, no problems with our two-sample $t$ at all.

