
## Survival analysis


* So far, have seen:


   * response variable counted or measured (regression)

   * response variable categorized (logistic regression)

and have predicted response from explanatory variables.

* But what if response is time until event (eg. time of
survival after surgery)?

* Additional complication: event might not have happened at end of study (eg. patient still alive). But knowing that patient has "not died yet" presumably informative. Such data called *censored*. 

* Enter *survival analysis*, in particular the "Cox proportional hazards model". 

* Explanatory variables in this context often called *covariates*.


## Example: still dancing?


* 12 women who have just started taking dancing lessons are
followed for up to a year, to see whether they are still taking
dancing lessons, or have quit. The "event" here is "quit".

* This might depend on:


   * a treatment (visit to a dance competition)

   * woman's age (at start of study).

## Data

\normalsize
```
Months  Quit   Treatment Age
1        1        0      16
2        1        0      24
2        1        0      18
3        0        0      27
4        1        0      25
7        1        1      26
8        1        1      36
10       1        1      38
10       0        1      45
12       1        1      47
```
\normalsize

## About the data


* `months` and `quit` are kind of combined response:


   *  `Months` is number of months a woman was actually observed dancing

   * `quit` is 1 if woman quit, 0 if still dancing at end of study.


* Treatment is 1 if woman went to dance competition, 0 otherwise.

* Fit model and see whether `Age` or `Treatment`
have effect on survival.

* Want to do predictions for probabilities of still dancing as
they depend on whatever is significant, and draw plot.

## Packages (for this section) 

* Install packages `survival` and `survminer` if not done. 

* Load `survival`, `survminer`,
`broom` and `tidyverse`:

```{r bSurvival-1, eval=F}
library(tidyverse)
library(survival)
library(survminer)
library(broom)
```


## Read data 


- Column-aligned: 

\normalsize
```{r bSurvival-2}
url <- "http://ritsokiguess.site/datafiles/dancing.txt"
dance <- read_table(url)
```
\normalsize


## The data

\small
```{r bSurvival-3}
dance
```
\normalsize
   

## Examine response and fit model


* Response variable: 

\small
```{r bSurvival-4}
dance %>% mutate(mth = Surv(Months, Quit)) -> dance
dance
```
\normalsize


  

* Then fit model, predicting `mth` from explanatories:
```{r bSurvival-5 }
dance.1 <- coxph(mth ~ Treatment + Age, data = dance)
```
 


## Output looks a lot like regression

\scriptsize
```{r bSurvival-6}
summary(dance.1)
```
\normalsize

## Conclusions


* Use $\alpha=0.10$ here since not much data.

* Three tests at bottom like global F-test. Consensus that
something predicts survival time (whether or not dancer quit and how
long it took).

* `Age` (definitely), `Treatment` (marginally) both
predict survival time.


## Model checking


* With regression, usually plot residuals against fitted values.

* Not quite same here (nonlinear model), but ``martingale
residuals'' should have no pattern vs.\ "linear predictor".

* `ggcoxdiagnostics` from package `survminer`
makes plot, to which we add smooth. If smooth trend more or less
straight across, model OK. 

* Martingale residuals can go very negative, so won't always
look normal.


## Martingale residual plot for dance data

This looks good (with only 12 points):

```{r bSurvival-7, fig.height=3}
#| warning = FALSE
ggcoxdiagnostics(dance.1) + geom_smooth(se = F)
```

   

## Predicted survival probs 
- The function we use is called
`survfit`, though actually works rather like
`predict`. 
- First create a data frame of values to predict from. We'll do all
combos of ages 20 and 40, treatment and not, using
`crossing` to get all the combos:

\small
```{r bSurvival-8}
treatments <- c(0, 1)
ages <- c(20, 40)
dance.new <- crossing(Treatment = treatments, Age = ages)
dance.new
```
\normalsize 

## The predictions  
One prediction *for each time* for each combo of age and treatment in `dance.new`:

```{r bSurvival-9, echo=F}
options(width = 80)
url <- "http://ritsokiguess.site/datafiles/dancing.txt"
dance <- read_table(url)
dance %>% mutate(mth = Surv(Months, Quit)) -> dance
```



\footnotesize
```{r bSurvival-10}
s <- survfit(dance.1, newdata = dance.new, data = dance)
summary(s)
```
\normalsize



## Conclusions from predicted probs


* Older women more likely to be still dancing than younger women
(compare "profiles" for same treatment group).

* Effect of treatment seems to be to increase prob of still
dancing (compare "profiles" for same age for treatment group
vs.\ not)

* Would be nice to see this on a graph. This is `ggsurvplot` from package `survminer`:




```{r bSurvival-11 }
g <- ggsurvplot(s, conf.int = F)
```

## "Strata" (groups)

- uses "strata" thus (`dance.new`): 

\footnotesize

```{r bSurvival-12, echo=F}
dance.new
```
\normalsize



## Plotting survival probabilities 
```{r survival-plot, fig.height=3.8}
g
```

   
