---
title: "logistic regression"
output: html_notebook
---

## Packages

```{r thing-1}
#library(MASS) # for later
library(tidyverse)
```

## Basic logistic regression

Read in the rat data:

```{r thing-2}
rats=read_delim("rat.txt"," ")
rats
```

These are 6 rats getting different doses of some poison, and what happened to them. For a rat getting larger dose, are they more likely to die?

Response variable categorical:

```{r thing-3}
rats
```

but needs to be `factor`, so make one (can overwrite):

```{r thing-4}
rats = rats %>% mutate(status=factor(status))
rats
```

now fit model:

```{r thing-5}
rats.1=glm(status~dose,family="binomial",data=rats)
summary(rats.1)
```

what did we actually predict?

```{r thing-6}
rats$status
```

First level is baseline (died), second is predicted (lived).

So negative slope means probability of **living** *decreases* as dose increases.

### predictions

Use `predict`. With just a model to predict from, uses original data `dose` values:

```{r thing-7}
p=predict(rats.1,type="response")
cbind(rats,p)
```

To predict for other doses, create a new data frame of them first:

```{r thing-8}
doses=c(2.5,3.5,6)
new=tibble(dose=doses)
new
p=predict(rats.1,new,type="response")
cbind(new,p)
```

## Rats part 2

More data (60 rats), arranged differently:

```{r thing-9}
rat2=read_delim("rat2.txt"," ")
rat2

```

Have *frequencies*: number of rats living or dying at each dose. Each line of data represents many rats (10), not just one.

Now have to *create* response variable (don't just have a "lived" or "died" in a column).

Create two-column response, column of lived, column of died:

```{r thing-10}

## ------------------------------------------------------------------------
response=with(rat2,cbind(lived,died))
response
class(response)
```

then use that on the left of the squiggle:

```{r thing-11}
rat2.1=glm(response~dose,family="binomial",
  data=rat2)
summary(rat2.1)
```

Probability of living decreases with increasing dose, and is now significant.

Using a data frame as response doesn't work:

```{r thing-12}
response = rat2 %>% select(lived,died)
rat2.1a=glm(response~dose,family="binomial",data=rat2)
```

but this does:

```{r thing-13}
response = rat2 %>% select(lived,died) %>% as.matrix()
rat2.1b=glm(response~dose,family="binomial",data=rat2)
summary(rat2.1b)
```

Predictions (using original doses in data set):

```{r thing-14}
p=predict(rat2.1,type="response")
cbind(rat2,p)
```

## Plotting  observed proportions against predictions

With multiple observations per line, can plot predicted probabilities of a rat living at that dose against the actual proportion of rats that lived.

- Add predictions to data frame
- Calculate observed proportion
- Plot observed proportions against dose as points
- Plot predictions joined by lines
- label the y-axis properly:

```{r thing-15}
rat2 %>% bind_cols(pred=p) %>% 
  mutate(obs=lived/(lived+died)) %>% 
  ggplot(aes(x=dose,y=obs))+geom_point()+
    geom_line(aes(y=pred))+
    scale_y_continuous("proportion")
```

## Sepsis data

```{r thing-16}
sepsis=read_delim("sepsis.txt"," ")
sepsis
```

Predict probability of death (`death=1`) from everything else

```{r thing-17}
sepsis.1=glm(death~., family="binomial", data=sepsis)
summary(sepsis.1)
```

Take out `malnut`:

```{r thing-18}
sepsis.2=update(sepsis.1,.~.-malnut)
summary(sepsis.2)
```

All coefficients positive (why?)

Do some predictions, for individuals 4, 1, 2, 11, 32:

```{r thing-19}
my_rows=c(4,1,2,11,32)
new = sepsis %>% slice(my_rows)
p=predict(sepsis.2,new,type="response")
cbind(new,p)
```

Note effect of age (first three rows) and of risk factors (last two) on prob of death.

Diagnostic: plot residuals against anything as usual:

```{r thing-20}
ggplot(sepsis.2,aes(x=.fitted,y=.resid))+geom_point()+geom_smooth()
```

Hard to read: positive residuals are from people who died, negative ones from people who survived. As long as smooth trend more or less straight across, OK.

Also need lack of trend of residuals vs. age. These are in different data frames, so either get them from right place:

```{r thing-21}
ggplot(sepsis.2,aes(x=sepsis$age,y=.resid))+geom_point()
```

or use `augment` from `broom`:

```{r thing-22}
library(broom)
d=augment(sepsis.2)
d
```

and then plot (now in *same* data frame):

```{r thing-23}
ggplot(d,aes(x=age,y=.resid))+geom_point()+geom_smooth()
```

The long string of points is people with no risk factors who survived (`death=0`). Prob of survival decreases with age, so these residuals get gradually more negative.

Apparently no trend on this plot, so relationship with age appears appropriate.

The other explanatory variables are 0-1, so assess their relationship with residuals by boxplots, eg. `alcohol`:

```{r thing-24}
ggplot(d,aes(x=factor(alcohol),y=.resid))+geom_boxplot()
```

Centre of each group should be zero; median for alcoholics is a bit low.

## Something else

Let's create a much simpler model containing just age and alcoholism and do some predictions for it:

```{r thing-25}
sepsis.3=update(sepsis.2,.~.-shock-bowelinf)
summary(sepsis.3)
```

Let's do some predictions for all combinations of `age` 25, 50, 75 and `alcohol` 0 and 1. Here's my procedure for doing that:

```{r thing-26}
ages=c(25,50,75)
alcohols=c(0,1)
new=crossing(age=ages,alcohol=alcohols)
new
```

The column names in `new` must be *exactly* the same as in the original data frame.

Then, feed `new` into `predict` as usual:

```{r thing-27}
p=predict(sepsis.3,new,type="response")
cbind(new,p)
```

- Being alcoholic vs. not greatly increases the probability of death (at any age) eg. compare rows 5 and 6
- Being older somewhat increases the probability of death (whether alcoholic or not) eg. compare rows 1, 3, 5.

Consistent with positive slopes in logistic regression.

