## Packages

Uses my package `mkac` which is on Github. Install with:

```{r time-series-1, eval=F}
library(devtools)
install_github("nxskok/mkac")
```

Plus these. You might need to install some of them first:

```{r time-series-2, eval=F}
library(ggfortify)
library(forecast)
library(tidyverse)
library(mkac) 
```

## Time trends

```{r time-series-3,echo=FALSE}
knitr::opts_chunk$set(fig.cap="")
```


* Assess existence or nature of time trends with:
  * correlation
  * regression ideas.
  * (later) time series analysis
  
## World mean temperatures


Global mean temperature every year since 1880:

\small
```{r time-series-4, message=F, warning=F, fig.height=3.3}
temp=read_csv("temperature.csv")
ggplot(temp, aes(x=year, y=temperature)) + 
  geom_point() + geom_smooth()
```
\normalsize

## Examining trend

* Temperatures increasing on average over time, but pattern very irregular.
* Find (Pearson) correlation with time, and test for significance:

```{r time-series-5}
with(temp, cor.test(temperature,year))
```

## Comments

* Correlation, 0.8695, significantly different from zero. 
* CI shows how far from zero it is.

Tests for *linear* trend with *normal* data. 

## Kendall correlation

Alternative, Kendall (rank) correlation, which just tests for monotone trend (anything upward, anything downward) and is resistant to outliers:

```{r time-series-6}
with(temp, cor.test(temperature,year,method="kendall"))
```

Kendall correlation usually closer to 0 for same data, but here P-values comparable. Trend again strongly significant.

## Mann-Kendall

- Another way is via **Mann-Kendall**: Kendall correlation with time.

- Use my package `mkac`:

\footnotesize
```{r time-series-7}
kendall_Z_adjusted(temp$temperature)
```
\normalsize

## Comments

- Standard Mann-Kendall assumes observations *independent*.
- Observations close together in time often *correlated* with each other.
- Correlation of time series "with itself" called **autocorrelation**.
- Adjusted P-value above is correction for autocorrelation (via Hamed-Rao).

## Examining rate of change

* Having seen that there *is* a change, question is "how fast is it?"
* Examine slopes:

  * regular regression slope, if you believe straight-line regression
  * Theil-Sen slope: resistant to outliers, based on medians
  

## Ordinary regression against time

```{r time-series-8}
lm(temperature~year, data=temp) %>% tidy() -> temp.tidy
temp.tidy
```

- Slope about 0.006 degrees per year 
- about this many degrees over course of data):

```{r time-series-9}
temp.tidy %>% pluck("estimate", 2)*130
```


## Theil-Sen slope

also from `mkac`:

```{r time-series-10}
theil_sen_slope(temp$temperature)
```

## Conclusions

- Slopes:
  * Linear regression: 0.005863
  * Theil-Sen slope: 0.005676
  * Very close.
- Correlations:
  * Pearson 0.8675
  * Kendall 0.6993
  * Kendall correlation smaller, but P-value equally significant (often the case)


## Constant rate of change?

Slope assumes that the rate of change is same over all years, but trend seemed to be accelerating:

```{r time-series-11, fig.height=3, message=F}
ggplot(temp, aes(x=year, y=temperature)) + 
  geom_point() + geom_smooth()
```

## Pre-1970 and post-1970:

```{r time-series-12}
temp %>% 
  mutate(time_period=
           ifelse(year<=1970, "pre-1970", "post-1970")) %>% 
  nest_by(time_period) %>%
  rowwise() %>% 
  mutate(theil_sen = theil_sen_slope(data$temperature))
```

Theil-Sen slope is very nearly *four times* as big since 1970 vs. before.





## Actual time series: the Kings of England

* Age at death of Kings and Queens of England since William the Conqueror (1066):

```{r time-series-13}
kings=read_table("kings.txt", col_names=F)
```

Data in one long column `X1`, so `kings` is data frame with one column. 

## Turn into `ts` time series object

```{r time-series-14}
kings.ts=ts(kings)
kings.ts
```


## Plotting a time series

`autoplot` from `ggfortify` gives time plot:

```{r time-series-15,"Kings-Time-Series", fig.height=3.7}
autoplot(kings.ts)
```

##  Comments

* "Time" here is order of monarch from William the Conqueror (1st) to George VI (last).

* Looks to be slightly increasing trend of age-at-death

* but lots of irregularity.


##  Stationarity

A time series is **stationary** if:

* mean is constant over time
* variability constant over time and not changing with mean.

Kings time series seems to have:

* non-constant mean
* but constant variability
* not stationary.

##  Getting it stationary

- Usual fix for non-stationarity is *differencing*: get new series from original one's values: 2nd - 1st, 3rd - 2nd etc.

In R, `diff`:

```{r time-series-16}
kings.diff.ts=diff(kings.ts)
```

##  Did differencing fix stationarity?

Looks stationary now:

```{r "Differenced-Kings-Series", fig.height=3.9}
autoplot(kings.diff.ts)
```



##  Births per month in New York City

from January 1946 to December 1959:

\small
```{r time-series-17, message=F}
ny=read_table("nybirths.txt",col_names=F)
ny
```
\normalsize

## As a time series

```{r time-series-18, echo=F}
w=getOption("width")
options(width=100)
```

\tiny
```{r time-series-19}
ny.ts=ts(ny,freq=12,start=c(1946,1))
ny.ts
```
\normalsize

```{r time-series-20, echo=F}
options(width=w)
```


## Comments

Note extras on `ts`:

* Time period is 1 year
* 12 observations per year (monthly) in `freq`
* First observation is 1st month of 1946 in `start`

Printing formats nicely.


##  Time plot

* Time plot shows extra pattern:

```{r time-series-21,fig.cap="", fig.height=3.8}
autoplot(ny.ts)
```

##  Comments on time plot

* steady increase (after initial drop)
* repeating pattern each year (seasonal component).
* Not stationary.

##  Differencing the New York births

Does differencing help here? Looks stationary, but some regular spikes:

```{r time-series-22, fig.height=3.6}
ny.diff.ts=diff(ny.ts)
autoplot(ny.diff.ts)
```


##  Decomposing a seasonal time series


A visual (using original data):

```{r time-series-23, fig.height=3.5}
ny.d <- decompose(ny.ts) 
ny.d %>% autoplot()
```

##  Decomposition bits


Shows:

* original series
* a "seasonal" part: something that repeats every year
* just the trend, going steadily up (except at the start)
* random: what is left over ("remainder")

##  The seasonal part

Fitted seasonal part is same every year, births lowest in February and highest in July:

```{r time-series-24, echo=F}
w=getOption("width")
options(width=100)
```


\tiny
```{r time-series-25}
ny.d$seasonal
```
\normalsize

```{r time-series-26, echo=F}
options(width=w)
```


```{r time-series-27, echo=FALSE}
set.seed(457299)
```

##  Time series basics: white noise

Each value independent random normal. Knowing one value tells you nothing about the next. "Random" process.


```{r "White-Noise", fig.height=3.0}
wn=rnorm(100)
wn.ts=ts(wn)
autoplot(wn.ts)
```




##  Lagging a time series

This means moving a time series one (or more) steps back in time:

```{r time-series-28}
x=rnorm(5)
tibble(x) %>% mutate(x_lagged=lag(x)) -> with_lagged
with_lagged
```

Gain a missing because there is nothing before the first observation.

##  Lagging white noise

```{r time-series-29, warning=F, fig.height=4}
tibble(wn) %>% mutate(wn_lagged=lag(wn)) -> wn_with_lagged
ggplot(wn_with_lagged, aes(y=wn, x=wn_lagged))+geom_point()
with(wn_with_lagged, cor.test(wn, wn_lagged, use="c")) # ignore the missing value
```


##  Correlation with lagged series

If you know about white noise at one time point, you know *nothing* about it at the next. This is shown by the scatterplot and the correlation. 


On the other hand, this:

\footnotesize
```{r time-series-30}
tibble(age=kings$X1) %>% 
  mutate(age_lagged=lag(age)) -> kings_with_lagged
with(kings_with_lagged, cor.test(age, age_lagged))
```
\normalsize

## Correlation with next value?

```{r time-series-31, warning=F, fig.height=4}
ggplot(kings_with_lagged, aes(x=age_lagged, y=age)) + 
  geom_point()
```

##  Two steps back:

\small
```{r time-series-32}
kings_with_lagged %>% 
  mutate(age_lag_2=lag(age_lagged)) %>% 
  with(., cor.test(age, age_lag_2))
```
\normalsize

Still a correlation two steps back, but smaller (and no longer significant).

##  Autocorrelation

Correlation of time series with *itself* one, two,... time steps back is useful idea, called **autocorrelation**. Make a plot of it with `acf` and `autoplot`. Here, white noise:

```{r time-series-33, fig.height=3.3}
acf(wn.ts, plot=F) %>% autoplot()
```

No autocorrelations beyond chance, anywhere (except *possibly* at lag 13).

Autocorrelations work best on *stationary* series.

##  Kings, differenced

```{r time-series-34, fig.height=4}
acf(kings.diff.ts, plot=F) %>% autoplot()
```

##  Comments on autocorrelations of kings series

Negative autocorrelation at lag 1, nothing beyond that. 

* If one value of differenced series positive, next one most likely negative.
* If one monarch lives longer than predecessor, next one likely lives shorter.

## NY births, differenced

```{r time-series-35, fig.height=4}
acf(ny.diff.ts, plot=F) %>% autoplot()
```

##  Lots of stuff:

* large positive autocorrelation at 1.0 years (July one year like July last year)
* large negative autocorrelation at 1 month.
* smallish but significant negative autocorrelation at 0.5 year = 6 months.
* Other stuff -- complicated.

##  Souvenir sales

Monthly sales for a beach souvenir shop in Queensland, Australia:

```{r time-series-36, echo=F}
w=getOption("width")
options(width=80)
```


\scriptsize
```{r time-series-37, message=F}
souv=read_table("souvenir.txt", col_names=F)
souv.ts=ts(souv,frequency=12,start=1987)
souv.ts
```
\normalsize

```{r time-series-38, echo=F}
options(width=w)
```


##  Plot of souvenir sales

```{r time-series-39, fig.height=4}
autoplot(souv.ts)
```

##  Several problems:

* Mean goes up over time
* Variability gets larger as mean gets larger
* Not stationary

##  Problem-fixing:

Fix non-constant variability first by taking logs:

```{r time-series-40, fig.height=3.5}
souv.log.ts=log(souv.ts)
autoplot(souv.log.ts)
```

##  Mean still not constant, so try taking differences

```{r time-series-41, fig.height=4}
souv.log.diff.ts=diff(souv.log.ts)
autoplot(souv.log.diff.ts)
```

##  Comments

* Now stationary
* but clear seasonal effect.

##  Decomposing to see the seasonal effect

```{r time-series-42, fig.height=4}
souv.d=decompose(souv.log.diff.ts)
autoplot(souv.d)
```

##  Comments 

**Big** drop in one month's differences. Look at seasonal component to see which:


```{r time-series-43, echo=F}
w=getOption("width")
options(width=90)
```

\tiny
```{r time-series-44}
souv.d$seasonal
```
\normalsize

```{r time-series-45, echo=F}
options(width=w)
```

January.

##  Autocorrelations

```{r time-series-46, fig.height=4}
acf(souv.log.diff.ts, plot=F) %>% autoplot()
```

* Big positive autocorrelation at 1 year (strong seasonal effect)
* Small negative autocorrelation at 1 and 2 months.


##  Moving average

- A particular type of time series called a **moving average** or MA process captures idea of autocorrelations at a few lags but not at others.

- Here's generation of MA(1) process, with autocorrelation at lag 1 but not otherwise:

```{r time-series-47}
beta=1
tibble(e=rnorm(100)) %>% 
  mutate(e_lag=lag(e)) %>% 
  mutate(y=e+beta*e_lag) %>% 
  mutate(y=ifelse(is.na(y), 0, y)) -> ma
```

## The series

\small
```{r time-series-48}
ma
```
\normalsize

##  Comments

* `e` contains independent "random shocks". 
* Start process at 0. 
* Then, each value of the time series has that time's random shock, plus a multiple of the last time's random shock. 
* `y[i]` has shock in common with `y[i-1]`; should be a lag 1 autocorrelation. 
* But `y[i]` has no shock in common with `y[i-2]`, so no lag 2 autocorrelation (or beyond).


##  ACF for MA(1) process

Significant at lag 1, but beyond, just chance:

```{r time-series-49, fig.height=3.5}
acf(ma$y, plot=F, na.rm=T) %>% autoplot()
```



##  AR process

Another kind of time series is AR process, where each value depends on previous one, like this (loop):

```{r time-series-50}
e=rnorm(100)
x=numeric(0)
x[1]=0
alpha=0.7
for (i in 2:100)
{
  x[i]=alpha*x[i-1]+e[i]
}
```

## The series

```{r time-series-51}
x
```


##  Comments

* Each random shock now only used for its own value of `x`
* but `x[i]` also depends on previous value `x[i-1]`
* so correlated with previous value
* *but* `x[i]` also contains multiple of `x[i-2]` and previous x's
* so all x's correlated, but autocorrelation dying away.

##  ACF for AR(1) series

```{r time-series-52, fig.height=4}
acf(x, plot=F) %>% autoplot()
```

##  Partial autocorrelation function

This cuts off for an AR series:

```{r time-series-53, fig.height=3.7}
pacf(x, plot=F) %>% autoplot()
```

The lag-2 autocorrelation should not be significant, and isn't.

##  PACF for an MA series decays slowly

```{r time-series-54, fig.height=4}
pacf(ma$y, plot=F) %>% autoplot()
```

##  The old way of doing time series analysis

Starting from a series with constant variability (eg. transform first to get it, as for souvenirs):

* Assess stationarity.
* If not stationary, take differences as many times as needed until it is.
* Look at ACF, see if it dies off. If it does, you have MA series.
* Look at PACF, see if that dies off. If it does, have AR series.
* If neither dies off, probably have a mixed "ARMA" series.
* Fit coefficients (like regression slopes).
* Do forecasts.


##  The new way of doing time series analysis (in R)

* Transform series if needed to get constant variability
* Use package `forecast`.
* Use function `auto.arima` to estimate what kind of series best fits data.
* Use `forecast` to see what will happen in future.

##  Anatomy of `auto.arima` output

```{r time-series-55}
auto.arima(ma$y)
```

Comments over.

## Comments

* ARIMA part tells you what kind of series you are estimated to have:
  * first number (first 0) is AR (autoregressive) part
  * second number (second 0) is amount of differencing here
  * third number (1) is MA (moving average) part
  
* Below that, coefficients (with SEs)
* AICc is measure of fit (lower better)

##  What other models were possible?

Run `auto.arima` with `trace=T`:

\small
```{r time-series-56}
auto.arima(ma$y,trace=T)
```
\normalsize

Also possible were MA(2) and ARMA(1,1), both with AICc=273.7.


##  Doing it all the new way: white noise

```{r time-series-57}
wn.aa=auto.arima(wn.ts)
wn.aa
```

Best fit *is* white noise (no AR, no MA, no differencing). 

##  Forecasts:

```{r time-series-58, echo=F}
options(width=65)
```


\small
```{r time-series-59}
forecast(wn.aa)
```
\normalsize

Forecasts all 0, since the past doesn't help to predict future.


##  MA(1)

```{r time-series-60}
y.aa=auto.arima(ma$y)
y.aa
y.f=forecast(y.aa)
```

##  Plotting the forecasts for MA(1) 

```{r time-series-61, fig.height=4}
autoplot(y.f)
```


##  AR(1)

```{r time-series-62}
x.aa=auto.arima(x)
x.aa
```

Oops! Thought it was MA(1), not AR(1)!

## Fit right AR(1) model:

```{r time-series-63}
x.arima=arima(x,order=c(1,0,0))
x.arima
```

##  Forecasts for `x`

```{r time-series-64, fig.height=4}
forecast(x.arima) %>% autoplot()
```

## Comparing wrong model:

```{r time-series-65, fig.height=4}
forecast(x.aa) %>% autoplot()
```


##  Kings
  
```{r time-series-66}
kings.aa=auto.arima(kings.ts)
kings.aa
```

##  Kings forecasts:

\small
```{r time-series-67}
kings.f=forecast(kings.aa)
kings.f
```
\normalsize

##  Kings forecasts, plotted

```{r time-series-68, fig.height=4}
autoplot(kings.f) + labs(x="index", y= "age at death")
```




##  NY births

Very complicated:

\small
```{r time-series-69}
ny.aa=auto.arima(ny.ts)
ny.aa
```
\normalsize


##  NY births forecasts

Not *quite* same every year:

\tiny
```{r time-series-70}
ny.f=forecast(ny.aa,h=36)
ny.f
```
\normalsize


##  Plotting the forecasts

```{r time-series-71, fig.height=4}
autoplot(ny.f)+labs(x="time", y="births")
```


##  Log-souvenir sales

```{r time-series-72}
souv.aa=auto.arima(souv.log.ts)
souv.aa
souv.f=forecast(souv.aa,h=27)
```

##  The forecasts

Differenced series showed low value for January (large drop). December highest, Jan and Feb lowest:

\scriptsize
```{r time-series-73}
souv.f
souv.f$mean
exp(souv.f$mean)
print.default(souv.f)
```
\normalsize

##  Plotting the forecasts

```{r time-series-74, fig.height=4}
autoplot(souv.f)
```


##  Global mean temperatures, revisited

```{r time-series-75}
temp.ts=ts(temp$temperature,start=1880)
temp.aa=auto.arima(temp.ts)
temp.aa
```

##  Forecasts

```{r time-series-76, fig.height=4}
temp.f=forecast(temp.aa)
autoplot(temp.f)+labs(x="year", y="temperature")
```

