---
title: "Analysis of Covariance"
---



## Analysis of covariance


* ANOVA: explanatory variables categorical (divide data into groups)

* traditionally, analysis of covariance has categorical $x$'s plus one numerical $x$ ("covariate") to be adjusted for.

* `lm` handles this too.

* Simple example: two treatments (drugs) (`a` and `b`), with before and after scores. 


* Does knowing before score and/or treatment help to predict after score?

* Is after score different by treatment/before score?





## Data

Treatment, before, after: 


\scriptsize
```
a 5 20
a 10 23
a 12 30
a 9 25
a 23 34
a 21 40
a 14 27
a 18 38
a 6 24
a 13 31
b 7 19
b 12 26
b 27 33
b 24 35
b 18 30
b 22 31
b 26 34
b 21 28
b 14 23
b 9 22
```
\normalsize


## Packages

```{r bAncova-1}
library(tidyverse)
library(broom)
library(marginaleffects)
```

the last of these for predictions.

## Read in data

```{r bAncova-2, message=F}
url <- "http://ritsokiguess.site/datafiles/ancova.txt"
prepost <- read_delim(url, " ")
prepost
```



## Making a plot
```{r ancova-plot, fig.height=4.5}
ggplot(prepost, aes(x = before, y = after, colour = drug)) +
  geom_point()
```



## Comments 

* As before score goes up, after score goes up.

* Red points (drug A) generally above blue points (drug B), for
comparable before score.

* Suggests before score effect *and* drug effect.


## The means
```{r bAncova-3 }
prepost %>%
  group_by(drug) %>%
  summarize(
    before_mean = mean(before),
    after_mean = mean(after)
  )
```
 


* Mean "after" score slightly higher for treatment A.

* Mean "before" score much higher for treatment B.

* Greater *improvement* on treatment A. 



## Testing for interaction
```{r bAncova-4 }
prepost.1 <- lm(after ~ before * drug, data = prepost)
anova(prepost.1)
```
 


* Interaction not significant. Will remove later.

## Predictions

Set up values to predict for:

```{r}
summary(prepost)
```

```{r}
new <- datagrid(before = c(9.75, 14, 21.25), 
                drug = c("a", "b"), model = prepost.1)
```

## and then


```{r}
cbind(predictions(prepost.1, newdata = new)) %>% 
  select(drug, before, estimate)
```
\normalsize

## Predictions (with interaction included), plotted

```{r, fig.height=4}
plot_predictions(model = prepost.1, condition = c("before", "drug"))
```

Lines almost parallel, but not quite.



## Taking out interaction
\small
```{r bAncova-8 }
prepost.2 <- update(prepost.1, . ~ . - before:drug)
anova(prepost.2)
```
 
\normalsize


* Take out non-significant interaction.

* `before` and `drug` strongly significant.

* Do predictions again and plot them.

## Predictions

```{r}
cbind(predictions(prepost.2, newdata = new)) %>% 
  select(drug, before, estimate)
```


## Plot of predicted values

```{r, fig.height=4}
plot_predictions(prepost.2, condition = c("before", "drug"))
```

This time the lines are *exactly* parallel. No-interaction model forces them
to have the same slope. 



## Different look at model output


* `anova(prepost.2)` tests for significant effect of
before score and of drug, but doesn't help with interpretation.

* `summary(prepost.2)` views as regression with slopes:

\scriptsize
```{r bAncova-11 }
summary(prepost.2)
```

\normalsize       




## Understanding those slopes

\footnotesize
```{r bAncova-12}
tidy(prepost.2)
```

\normalsize



* `before` ordinary numerical variable; `drug`
categorical. 

* `lm` uses first category `druga` as baseline.

* Intercept is prediction of after score for before score 0 and
*drug A*.

* `before` slope is predicted change in after score when
before score increases by 1 (usual slope)

* Slope for `drugb` is *change* in predicted after
score for being on drug B rather than drug A. Same for *any*
before score (no interaction).



## Summary


* ANCOVA model: fits different regression line for each group,
predicting response from covariate.

* ANCOVA model with interaction between factor and covariate
allows different slopes for each line.

* Sometimes those lines can cross over!

* If interaction not significant, take out. Lines then parallel.

* With parallel lines, groups have consistent effect regardless
of value of covariate.

