---
title: "1-way anova in stan"
output: html_notebook
---

## packages

```{r anova-loo-1}
library(tidyverse)
library(rstan)
library(loo)
library(conflicted)
conflict_prefer("extract", "rstan")
conflict_prefer("filter", "dplyr") 
conflict_prefer("lag", "dplyr")
conflict_prefer("loo", "loo")
conflict_prefer("Position", "ggplot2") 
```

## some data

```{r anova-loo-2}
my_url="http://ritsokiguess.site/datafiles/jumping.txt"
rats0=read_delim(my_url," ")
rats0
```

turn the groups into numbers

```{r anova-loo-3}
rats0 %>% mutate(group_fct=fct_inorder(group),
              group_no=as.integer(group_fct)) -> rats
rats
```

## fitting a model with different mean for each group

Use before, but add `generated quantities` section. See `anova-loo.stan`:

```{r anova-loo-4}
anova_loo_compiled=stan_model("anova-loo.stan")
```

and then sample from it as usual

```{r anova-loo-5}
anova_data=list(n_obs=30, n_group=3, density=rats$density, group_no=rats$group_no)
anova_loo_samples=sampling(anova_loo_compiled, data=anova_data)
```

## fitting a model with *same* mean for each group

We want to compare this model with a "null" model. In this case, the null would be that all the groups have the *same* mean. To fit that, go back to our Stan code, and replace `mu` as a vector with `mu` as a scalar. This is in `anova_loo_null.stan`.

```{r anova-loo-6}
anova_loo_null_compiled <- stan_model("anova_loo_null.stan")
```

and then sample from *that*, using same data we used before (I didn't change the `data` section):

```{r anova-loo-7}
anova_loo_null_samples=sampling(anova_loo_null_compiled, data=anova_data)
```

check that it worked:

```{r anova-loo-8}
anova_loo_samples
```

```{r anova-loo-9}
anova_loo_null_samples
```


## compare the fits of the two models, thus:

```{r anova-loo-10}
log_lik_a=extract_log_lik(anova_loo_samples, merge_chains = F)
log_lik_0=extract_log_lik(anova_loo_null_samples, merge_chains = F)
r_eff_a=relative_eff(log_lik_a)
r_eff_0=relative_eff(log_lik_0)

```

and then

```{r anova-loo-11}
(loo_a <- loo(log_lik_a, r_eff=r_eff_a))
```

```{r anova-loo-12}
(loo_0 <- loo(log_lik_0, r_eff=r_eff_0))
```

The model with all three group means estimated has a better (lower) looic, but that's what you would expect (since it is a bigger model). 

Compare:

```{r anova-loo-13}
(v=compare(loo_a, loo_0))
```

The negative sign means that the first model fits better, and so should be preferred. Is there a "substantial" difference? Divide by SE:

```{r anova-loo-14}
v[1]/v[2]
```

More than two SEs different, so the model with different means for each group is "substantially" better than having the same mean for each group.
