
## Analysis of covariance


* ANOVA: explanatory variables categorical (divide data into groups)

* traditionally, analysis of covariance has categorical $x$'s plus one numerical $x$ ("covariate") to be adjusted for.

* `lm` handles this too.

* Simple example: two treatments (drugs) (`a` and `b`), with before and after scores. 


* Does knowing before score and/or treatment help to predict after score?

* Is after score different by treatment/before score?





## Data

Treatment, before, after: 


\scriptsize
```
a 5 20
a 10 23
a 12 30
a 9 25
a 23 34
a 21 40
a 14 27
a 18 38
a 6 24
a 13 31
b 7 19
b 12 26
b 27 33
b 24 35
b 18 30
b 22 31
b 26 34
b 21 28
b 14 23
b 9 22
```
\normalsize


## Packages
`tidyverse` and `broom`:
```{r, eval=FALSE }
library(tidyverse)
library(broom)
```

## Read in data

```{r, message=F}
url <- "http://www.utsc.utoronto.ca/~butler/d29/ancova.txt"
# url <- "ancova.txt"
prepost <- read_delim(url, " ")
prepost # randomly chosen rows
```



## Making a plot
```{r ancova-plot<}
ggplot(prepost, aes(x = before, y = after, colour = drug)) +
  geom_point()
```



## Comments 

* As before score goes up, after score goes up.

* Red points (drug A) generally above blue points (drug B), for
comparable before score.

* Suggests before score effect *and* drug effect.


## The means
```{r }
prepost %>%
  group_by(drug) %>%
  summarize(
    before_mean = mean(before),
    after_mean = mean(after)
  )
```
 


* Mean "after" score slightly higher for treatment A.

* Mean "before" score much higher for treatment B.

* Greater *improvement* on treatment A. 



## Testing for interaction
```{r }
prepost.1 <- lm(after ~ before * drug, data = prepost)
anova(prepost.1)
drop1(prepost.1, test = "F")
```
 
```{r}
summary(prepost.1)
```


* Interaction not significant. Will remove later.


## Predictions, with interaction included

Make combinations of before score and drug:
```{r }
new <- crossing(
  before = c(5, 15, 25),
  drug = c("a", "b")
)
new
```

   

## Do predictions: 

```{r size="small"}
pred <- predict(prepost.1, new)
pred
preds <- cbind(new, pred)
# preds <- bind_cols(new, pred = pred)
preds
```

 



## Making a plot with lines for each `drug`
```{r AJHSA}
g <- ggplot(prepost,
  aes(x = before, y = after, colour = drug)) +
  geom_point() + geom_line(data = preds, aes(y = pred))
```
 
* Here, final line:


  *   joins points by lines \emph{for different data
set} (`preds` rather than `prepost`),

  *   *different $y$* (`pred` rather than `after`),

  * but same $x$ (`x=before` inherited from first `aes`).

* Last line could (more easily) be 

```{r eval=F}
geom_smooth(method = "lm", se = F)
```

which would work here, but not for later plot.






## The plot

* Lines almost parallel, but not quite.
 
* Non-parallelism (interaction) not significant:

```{r nachwazzo, fig.height=3.5, echo=F}
g
```
 

 

 

## Taking out interaction
\small
```{r }
prepost.2 <- update(prepost.1, . ~ . - before:drug)
anova(prepost.2)
drop1(prepost.2, test = "F")
```
 
\normalsize


* Take out non-significant interaction.

* `before` and `drug` strongly significant.

* Do predictions again and plot them.



## Predicted values again (no-interaction model)
```{r }
new
pred <- predict(prepost.2, new)
preds <- bind_cols(new, pred = pred)
preds
```

 
Each increase of 10 in before score results in 8.3 in predicted after
score, *the same for both drugs*.


## Making a plot, again
```{r }
g <- ggplot(
  prepost,
  aes(x = before, y = after, colour = drug)
) +
  geom_point() +
  geom_line(data = preds, aes(y = pred))
```
 
Exactly same as before, but using new predictions.


## The no-interaction plot of predicted values
```{r cabazzo,fig.height=3}
g
```
 
Lines now *parallel*. No-interaction model forces them
to have the same slope. 



## Different look at model output


* `anova(prepost.2)` tests for significant effect of
before score and of drug, but doesn't help with interpretation.

* `summary(prepost.2)` views as regression with slopes:

\scriptsize
```{r }
summary(prepost.2)
```

\normalsize       




## Understanding those slopes

\footnotesize
```{r}
tidy(prepost.2)
```

\normalsize



* `before` ordinary numerical variable; `drug`
categorical. 

* `lm` uses first category `druga` as baseline.

* Intercept is prediction of after score for before score 0 and
*drug A*.

* `before` slope is predicted change in after score when
before score increases by 1 (usual slope)

* Slope for `drugb` is *change* in predicted after
score for being on drug B rather than drug A. Same for *any*
before score (no interaction).



## Summary


* ANCOVA model: fits different regression line for each group,
predicting response from covariate.

* ANCOVA model with interaction between factor and covariate
allows different slopes for each line.

* Sometimes those lines can cross over!

* If interaction not significant, take out. Lines then parallel.

* With parallel lines, groups have consistent effect regardless
of value of covariate.

