
## Discriminant analysis


* ANOVA and MANOVA: predict a (counted/measured) response from group membership.

* Discriminant analysis: predict group membership based on counted/measured variables.

* Covers same ground as logistic regression (and its variations), but emphasis on classifying observed data into correct groups.

* Does so by searching for linear combination of original variables that best separates data into groups (canonical variables).

* Assumption here that groups are known (for data we have). If trying to "best separate" data into unknown groups, see *cluster analysis*.

* Examples: revisit seed yield and weight data, peanut data,
professions/activities data; remote-sensing data.



## Packages
```{r eval=F}
library(MASS)
library(tidyverse)
library(ggrepel)
library(ggbiplot)
```

   

`ggrepel` allows labelling points on a plot so they don't
overwrite each other.


## About `select`


* Both `dplyr` (in `tidyverse`) and `MASS`
have a function called `select`, and \emph{they do
different things}.

* How do you know which `select` is going to get called? 

* With `library`, the one loaded *last* is visible,
and others are not.

* Thus we can access the `select` in `dplyr` but
not the one in `MASS`. If we wanted that one, we'd have to
say `MASS::select`.

* I loaded `MASS` before
`tidyverse`. If I had done it the other way around, the
`tidyverse` `select`, which I want to use, would have
been the invisible one.  

- Alternative: load `conflicted` package. Any time you load two packages containing functions with same name, you get error and have to choose between them. 

## Example 1: seed yields and weights
\small
```{r message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/manova1.txt"
hilo <- read_delim(my_url, " ")
g <- ggplot(hilo, aes(
  x = yield, y = weight,
  colour = fertilizer
)) + geom_point(size = 4)
```
\normalsize
 
```{r, include=F, message=F}
ggsave("berzani.pdf", g, device="pdf")
```
 

\begin{minipage}[t]{0.38\linewidth}
\vspace{0.1\textheight}
Recall data from MANOVA: needed a multivariate analysis to find
difference in seed yield and weight based on whether they were high
or low fertilizer.
\end{minipage}\hfill
\begin{minipage}[t][][b]{0.55\textwidth}
\includegraphics[width=0.9\textwidth]{berzani}
\end{minipage}

## Basic discriminant analysis

```{r }
hilo.1 <- lda(fertilizer ~ yield + weight, data = hilo)
```

 



* Uses `lda` from package MASS.

* "Predicting" group membership from measured variables.




## Output

\small
```{r}
hilo.1
```
\normalsize
 



## Things to take from output


* Group means: high-fertilizer plants have (slightly) higher
mean yield and weight than low-fertilizer plants.

* "Coefficients of linear discriminants": \texttt{LD1,
LD2,}\ldots are scores constructed from observed variables that
best separate the groups.


* For any plant, get LD1 score by taking $-0.76$ times yield
plus $-1.25$ times weight, add up, standardize.

- the LD1 coefficients are like slopes: 
  - if yield higher, LD1 score for a plant lower
  - if weight higher, LD1 score for a plant lower

* High-fertilizer plants have higher yield and weight, thus
low (negative) LD1 score. Low-fertilizer plants have low yield and
weight, thus high (positive) LD1 score.

* One LD1 score for each observation. Plot with actual groups.




## How many linear discriminants?


* Smaller of these:

  * Number of variables

  * Number of groups *minus 1*


* Seed yield and weight: 2 variables, 2 groups,
$\min(2,2-1)=1$. 



## Getting LD scores
Feed output from LDA into `predict`:

```{r }
hilo.pred <- predict(hilo.1)
```

 

Component `x` contains LD score(s), here in descending order:

\footnotesize
```{r}
d <- cbind(hilo, hilo.pred$x) %>% arrange(desc(LD1))
d
```
\normalsize
 
High fertilizer have yield and weight high, negative LD1 scores.


## Plotting LD1 scores
With one LD score, plot against (true) groups, eg. boxplot: 
```{r fig.height=3.4}
ggplot(d, aes(x = fertilizer, y = LD1)) + geom_boxplot()
```

   


## Potentially misleading




```{r}
hilo.1$scaling
```


* These are like regression slopes: 
change in LD1 score for 1-unit change in variables. 

## But\ldots

* One-unit change in variables might not be comparable: 

```{r}
hilo %>% select(-fertilizer) %>% 
   summarize(across(everything(), ~quantile(., c(0.25, 0.75))))
```

* Here, IQRs both 2.2, *identical*, so 1-unit change in each variable
means same thing.



## What else is in `hilo.pred?`

\small
```{r}
names(hilo.pred)
```
\normalsize
     



* `class`: predicted fertilizer level (based on values of
`yield` and `weight`).

* `posterior`: predicted probability of being low or high
fertilizer given `yield` and `weight`.
 


## Predictions and predicted groups
\ldots based on `yield` and `weight`: 

\footnotesize
```{r}
cbind(hilo, predicted = hilo.pred$class)
table(obs = hilo$fertilizer, pred = hilo.pred$class)
```
\normalsize
 


## Understanding the predicted groups


* Each predicted fertilizer level is exactly same as observed
one (perfect prediction).

* Table shows no errors: all values on top-left to bottom-right
diagonal. 



## Posterior probabilities
show how clear-cut the classification decisions were:

\small
```{r}
pp <- round(hilo.pred$posterior, 4)
d <- cbind(hilo, hilo.pred$x, pp)
d
```
\normalsize

Only obs.\ 7 has any doubt: `yield` low for a high-fertilizer,
but high `weight` makes up for it.
 


## Example 2: the peanuts 

\scriptsize
```{r message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/peanuts.txt"
peanuts <- read_delim(my_url, " ")
peanuts
```
\normalsize
 

- Recall: `location` and `variety` both significant in
MANOVA. Make combo of them (over):


## Location-variety combos

\footnotesize
```{r combos}
peanuts %>%
   unite(combo, c(variety, location)) -> peanuts.combo
peanuts.combo
```

\normalsize

## Discriminant analysis

\small
```{r}
peanuts.1 <- lda(combo ~ y + smk + w, data = peanuts.combo)
peanuts.1$scaling
peanuts.1$svd
```
\normalsize
   



* Now 3 LDs (3 variables, 6 groups, $\min(3,6-1)=3$).

## Comments

* First: relationship of LDs to original variables. Look for
coeffs far from zero: here,


   *   high `LD1` mainly high `w`
or low `y`.

   * high `LD2` mainly high `w`.


* `svd` values show relative importance of LDs:
`LD1` much more important than `LD2`.



## Group means by variable
```{r }
peanuts.1$means
```


* `5_2` clearly smallest on `y`, `smk`, near
smallest on `w`

* `8_2` clearly biggest on `smk`, `w`, also
largest on `y`

* `8_1` large on `w`, small on `y`.


## The predictions and misclassification
```{r }
peanuts.pred <- predict(peanuts.1)
table(
  obs = peanuts.combo$combo,
  pred = peanuts.pred$class
)
```

Actually classified very well. Only one `6_2` classified as a
`5_1`, rest all correct.


## Posterior probabilities

\scriptsize
```{r}
pp <- round(peanuts.pred$posterior, 2)
peanuts.combo %>%
  select(-c(y, smk, w)) %>%
  cbind(., pred = peanuts.pred$class, pp)
```
\normalsize
   

*Some* doubt about which combo each plant belongs in, but not too
much. The one misclassified plant was a close call.


## Discriminant scores, again


* How are discriminant scores related to original variables?

* Construct data frame with original data and discriminant
scores side by side:

\footnotesize
```{r}
peanuts.1$scaling
lds <- round(peanuts.pred$x, 2)
mm <- with(peanuts.combo,
           data.frame(combo, y, smk, w, lds))
```
\normalsize
   

* LD1 positive if `w` large and/or `y` small.

* LD2 positive if `w` large.    





## Discriminant scores for data

\footnotesize
```{r}
mm
```
\normalsize
   


* Obs.\ 5 and 6 have most negative `LD1`: large `y`,
small `w`.

* Obs.\ 4 has most negative `LD2`: small `w`.



## Predict typical LD1 scores
First and third quartiles for three response variables:
```{r }
peanuts %>%
   select(y:w) %>%
summarize(across(everything(), ~quantile(., c(0.25, 0.75)))) -> quartiles
quartiles
new <- with(quartiles, crossing(y, smk, w))
```

   


## The combinations
```{r }
new
pp <- predict(peanuts.1, new)
```

   


## Predicted typical LD1 scores

\footnotesize
```{r}
cbind(new, pp$x) %>% arrange(LD1)
```
\normalsize
   




* Very negative LD1 score with large `y` and small
`w`

* `smk` doesn't contribute much to LD1

* Very positive LD1 score with small `y` and large
`w`.

* Same as we saw from Coefficients of Linear Discriminants.


## Plot LD1 vs.\ LD2, labelling by combo
```{r fig.height=3.5}
g <- ggplot(mm, aes(x = LD1, y = LD2, colour = combo, 
                    label = combo)) + geom_point() +
  geom_text_repel() + guides(colour = F)
g
```

   


## "Bi-plot" from `ggbiplot`

```{r fig.height=3.3}
ggbiplot(peanuts.1,
  groups = factor(peanuts.combo$combo)
)
```


## Installing `ggbiplot`


* `ggbiplot` not on CRAN, so usual
`install.packages` will not work.

* Install package `devtools` first (once):
```{r eval=F}
install.packages("devtools")
```

     

* Then install `ggbiplot` (once):
```{r eval=F}
library(devtools)
install_github("vqv/ggbiplot")
```

     


## Cross-validation


* So far, have predicted group membership from same data used to
form the groups --- dishonest!

* Better: *cross-validation*: form groups from all
observations *except one*, then predict group membership for
that left-out observation.

* No longer cheating!

* Illustrate with peanuts data again.



## Misclassifications


* Fitting and prediction all in one go:

\small
```{r }
peanuts.cv <- lda(combo ~ y + smk + w,
  data = peanuts.combo, CV = T)
table(obs = peanuts.combo$combo,
      pred = peanuts.cv$class)
```
\normalsize
   
* Some more misclassification this time.




## Repeat of LD plot
```{r graziani,fig.height=3.7}
g
```

   


## Posterior probabilities

\footnotesize
```{r}
pp <- round(peanuts.cv$posterior, 3)
data.frame(
  obs = peanuts.combo$combo,
  pred = peanuts.cv$class, pp
)
```
\normalsize
   


## Why more misclassification?


* When predicting group membership for one observation, only
uses the *other one* in that group.

* So if two in a pair are far apart, or if two groups overlap,
great potential for misclassification.

* Groups `5_1` and `6_2` overlap.

* `5_2` closest to `8_1`s looks more like an
`8_1` than a `5_2` (other one far away).

* `8_1`s relatively far apart and close to other things,
so one appears to be a `5_2` and the other an `8_2`.


## Example 3: professions and leisure activities


* 15 individuals from three different professions (politicians,
administrators and belly dancers) each participate in four
different leisure activities: reading, dancing, TV watching and
skiing. After each activity they rate it on a 0--10 scale.

* Some of the data:

\normalsize
```
bellydancer 7 10 6 5
bellydancer 8 9 5 7
bellydancer 5 10 5 8
politician 5 5 5 6
politician 4 5 6 5
admin 4 2 2 5
admin 7 1 2 4
admin 6 3 3 3
```
\normalsize

## Questions

* How can we best use the scores on the activities to predict a person's profession?

* Or, what combination(s) of scores best separate data into profession groups?




## Discriminant analysis

\small
```{r message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/profile.txt"
active <- read_delim(my_url, " ")
active.1 <- lda(job ~ reading + dance + tv + ski, data = active)
active.1$svd
active.1$scaling
```
\normalsize
   

* Two discriminants, first fair bit more important than second.

* `LD1` depends (negatively) most on `dance`, a bit
on `tv`.

* `LD2` depends mostly on `tv`.



## Misclassification
```{r }
active.pred <- predict(active.1)
table(obs = active$job, pred = active.pred$class)
```

   

Everyone correctly classified.


## Plotting LDs 

\small
```{r fig.height=3}
mm <- data.frame(job = active$job, active.pred$x, person = 1:15)
g <- ggplot(mm, aes(x = LD1, y = LD2, colour = job, 
                    label = job)) + 
  geom_point() + geom_text_repel() + guides(colour = F)
g
```
\normalsize
   


## Biplot 
```{r fig.height=4.3}
ggbiplot(active.1, groups = active$job)
```

   


## Comments on plot


* Groups well separated: bellydancers top left, administrators
top right, politicians lower middle.

* Bellydancers most negative on `LD1`: like dancing most.

* Administrators most positive on `LD1`: like dancing least.

* Politicians most negative on `LD2`: like TV-watching most.



## Plotting individual `persons`
Make `label` be identifier of person. Now need legend:

```{r fig.height=2.8}
ggplot(mm, aes(x = LD1, y = LD2,  colour = job, 
               label = person)) + 
  geom_point() + geom_text_repel()
```

   


## Posterior probabilities

\scriptsize
```{r}
pp <- round(active.pred$posterior, 3)
data.frame(obs = active$job, pred = active.pred$class, pp)
```
\normalsize
   
Not much doubt.


## Cross-validating the jobs-activities data
Recall: no need for `predict`. Just pull out `class` and
make a table:  
```{r }
active.cv <- lda(job ~ reading + dance + tv + ski,
  data = active, CV = T
)
table(obs = active$job, pred = active.cv$class)
```

   

This time one of the bellydancers was classified as a politician.


## and look at the posterior probabilities
picking out the ones where things are not certain:

\footnotesize
```{r}
pp <- round(active.cv$posterior, 3)
data.frame(obs = active$job, pred = active.cv$class, pp) %>%
  mutate(max = pmax(admin, bellydancer, politician)) %>%
  filter(max < 0.9995)
```
\normalsize


* Bellydancer was "definitely" a politician!

* One of the administrators might have been a politician too.


## Why did things get misclassified?

```{r include=F, message=F}
ggsave("nesta.pdf", g, "pdf")
```


\begin{minipage}[t]{0.3\linewidth}

\begin{itemize}

\item Go back to plot of discriminant scores:

\item one bellydancer much closer to the politicians,

\item one administrator a bit closer to the politicians.
\end{itemize}
\end{minipage}\hfill
\begin{minipage}[t][][b]{0.68\linewidth}

\includegraphics[width=0.9\textwidth]{nesta}
       
\end{minipage}

## Example 4: remote-sensing data


* View 38 crops from air, measure 4 variables `x1-x4`.

* Go back and record what each crop was.

* Can we use the 4 variables to distinguish crops?



## Reading in
\small
```{r }
my_url <- 
   "http://www.utsc.utoronto.ca/~butler/d29/remote-sensing.txt"
crops <- read_table(my_url)
```
\normalsize
   

## Starting off: number of LDs
```{r }
crops.lda <- lda(crop ~ x1 + x2 + x3 + x4, data = crops)
crops.lda$svd
```

 



* 4 LDs (four variables, six groups).

* 1st one important, maybe 2nd as well.


## Connecting original variables and LDs
\small
```{r }
crops.lda$means
round(crops.lda$scaling, 3)
```
\normalsize
   



* Links groups to original variables to LDs.



## `LD1` and `LD2`
```{r }
round(crops.lda$scaling, 3)
```



* `LD1` mostly `x1` (minus), so clover low on
`LD1`, corn high.

* `LD2` `x3` (minus), `x2` (plus), so
sugarbeets should be high on `LD2`.



## Predictions


* Thus:

\footnotesize
```{r}
crops.pred <- predict(crops.lda)
table(obs = crops$crop, pred = crops.pred$class)
```
\normalsize
   

* Not very good, eg. only 6 of 11 `Clover` classified correctly.

* Set up for plot:
```{r }
mm <- data.frame(crop = crops$crop, crops.pred$x)
```

   


## Plotting the LDs 
```{r piacentini,fig.height=3.4}
ggplot(mm, aes(x = LD1, y = LD2, colour = crop)) +
  geom_point()
```

   


## Biplot
```{r fig.height=3}
ggbiplot(crops.lda, groups = crops$crop)
```

   


## Comments


* Corn high on LD1 (right).

* Clover all over the place, but mostly low on LD1 (left).

* Sugarbeets tend to be high on LD2.

* Cotton tends to be low on LD2.

* Very mixed up.



## Try removing Clover


* the `dplyr` way:
```{r }
crops %>% filter(crop != "Clover") -> crops2
crops2.lda <- lda(crop ~ x1 + x2 + x3 + x4, data = crops2)
```

   


* LDs for `crops2` will be different from before.

* Concentrate on plot and posterior probs.

```{r }
crops2.pred <- predict(crops2.lda)
mm <- data.frame(crop = crops2$crop, crops2.pred$x)
```

   


## `lda output`
Different from before:

\footnotesize
```{r}
crops2.lda$means
crops2.lda$svd
crops2.lda$scaling
```
\normalsize


## Plot

A bit more clustered:
```{r nedved,fig.height=4.5}
ggplot(mm, aes(x = LD1, y = LD2, colour = crop)) +
  geom_point()
```

   



## Biplot
```{r fig.height=2.9}
ggbiplot(crops2.lda, groups = crops2$crop)
```

   


## Quality of classification

\small
```{r}
table(obs = crops2$crop, pred = crops2.pred$class)
```
\normalsize
   

Better.


## Posterior probs (some)
```{r echo=F}
options(width = 60)
```

\footnotesize  
```{r }
post <- round(crops2.pred$posterior, 3)
data.frame(obs = crops2$crop, pred = crops2.pred$class, post) %>%
  filter(obs != pred)
```
\normalsize

## Comments

* These were the misclassified ones, but the posterior probability
of being correct was not usually too low. 

* The correctly-classified ones are not very clear-cut either.




## MANOVA
Began discriminant analysis as a followup to MANOVA. Do our variables
significantly separate the crops (excluding Clover)?

```{r }
response <- with(crops2, cbind(x1, x2, x3, x4))
crops2.manova <- manova(response ~ crop, data = crops2)
summary(crops2.manova)
```

 

Yes, at least one of the crops differs (in means) from the others. So
it is worth doing this analysis.

We did this the wrong way around, though!


## The right way around


* *First*, do a MANOVA to see whether any of the groups
differ significantly on any of the variables.

* *If the MANOVA is significant*, do a discriminant
analysis in the hopes of understanding how the groups are different.

* For remote-sensing data (without Clover):


* LD1 a fair bit more important than LD2 (definitely ignore LD3).

* LD1 depends mostly on `x1`, on which Cotton was high
and Corn was low. 


* Discriminant analysis in MANOVA plays the same kind of role
that Tukey does in ANOVA.


   



