\section{Introduction to `ggplot`}
\frame{\sectionpage}

## Base graphics and `ggplot`


* When R began life (in 1993), graphics terminals were new, and
was a big rush to use them to display statistical plots.

* Since then, graphics have been "bolted on" with no clear
overarching plan.

* This means that each kind of graph uses different ideas.

* Hadley Wickham used idea of "grammar of graphics" to develop
`ggplot`, a set of graphing tools with consistent ideas.

* See
[link](http://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf).

* In package `ggplot2`, thus first:
```{r }
library(tidyverse)
```

     



## Ideas behind `ggplot`


* Separates *what to plot* from *how to plot it*.

* What to plot, for example:
\begin{description}
\item[x] variable to put on the $x$-axis
\item[y] variable to put on the $y$-axis
\item[colour] colour to make the points (eg. colouring by a
categorical variable)
\item[group] group points together (eg. which ones to join  by lines)
\end{description}

* How to plot, such as:
\begin{description}[geom\_histogram]
\item[geom\_histogram] histogram
\item[geom\_boxplot] boxplot
\item[geom\_point] (scatter plot of) points
\item[geom\_line] join (possibly grouped) points by lines.
\item[facet\_grid] draw separate graphs of subsets of data and put
them side by side.
\end{description}



## Degree of Reading Power data


* Some children were randomly assigned to learn to read via a
new method (labelled `t` for "treatment") or via the
standard method (labelled `c` for "control"). Each child
was given a reading test at the end. How do the test scores
compare for the two groups?

* Read in and show some of the data:
```{r }
drp <- read.table("drp.txt", header = T)
head(drp)
```

     



## Histogram of all reading scores
Histogram has only an $x$-scale (reading score), so do this:

```{r fig.height=3.5}
ggplot(drp, aes(x = score)) +
  geom_histogram()
```

 


## Better histogram
Set number of bins to 10:
```{r fig.height=4.5}
ggplot(drp, aes(x = score)) + geom_histogram(bins = 10)
```

   


## Histograms by group, side by side
Add a `facet_grid`, thus:

```{r fig.height=4.4}
ggplot(drp, aes(x = score)) + geom_histogram(bins = 10) +
  facet_grid(group ~ .)
```

   


## Comments


* Control group scores (top) more spread out (they go higher
and lower).

* Not sure how means/medians compare.

* Above/below histograms make for easier comparison.

* In `facet_grid`, variable before squiggle denotes
up/down, variable after squiggle denotes left/right (like
`y~x` in usual model formula).

* If nothing to go before or after squiggle, use `.`


* Here, no left-right graphs, so right side of squiggle is `.`




## Boxplot of reading scores by group
Boxplot has $y$ scale (measured variable) *and* $x$ scale
(groups), so specify both:
```{r fig.height=4}
ggplot(drp, aes(x = group, y = score)) + geom_boxplot()
```

   


## Boxplot of a single group
To get a boxplot of all the reading scores together, specify a
"dummy" group for `x`:
```{r cateloupe,fig.height=4}
ggplot(drp, aes(x = factor(1), y = score)) + geom_boxplot()
```

   


## Scatterplot
For this, use analysis of covariance data from later:

```{r }
prepost <- read.table("ancova.txt", header = T)
str(prepost)
```

 



* 
20 subjects were randomized to one of two drugs `a` and
`b`.

* Each subject had their before and after scores measured on
some test.

* Want a scatterplot of after score against before score
labelled by drug.



## Basic scatterplot
```{r fig.height=5}
ggplot(prepost, aes(x = before, y = after)) + geom_point()
```

   


## Coloured by `drug`
Add `colour=` to the *aesthetic*:

```{r fig.height=4.3}
ggplot(prepost, aes(x = before, y = after, colour = drug)) +
  geom_point()
```

 


## Adding a smooth trend
```{r fig.height=4}
ggplot(prepost, aes(x = before, y = after)) +
  geom_point() + geom_smooth()
```

   


## A smooth trend for each drug
```{r fig.height=3.5}
ggplot(prepost, aes(x = before, y = after, colour = drug)) +
  geom_point() + geom_smooth()
```

   


## A regression line

```{r fig.height=4.5}
ggplot(prepost, aes(x = before, y = after)) +
  geom_point() + geom_smooth(method = "lm")
```

   


## A regression line for each drug

```{r kalloni,fig.height=4.5}
ggplot(prepost, aes(x = before, y = after, colour = drug)) +
  geom_point() + geom_smooth(method = "lm")
```

   

## Think of before-after data as repeated measures
Reorganize:
```{r }
prepost %>%
  mutate(subject = row_number()) %>%
  pivot_longer(before:after, names_to="time", 
               values_to="score", names_ptypes = list(time=factor())
  ) -> prepost.long
prepost.long %>% sample_n(8) # 8 random rows
```

   

*One* column of scores, with column `time` saying whether
was before or after.


## A "spaghetti plot"
```{r fig.height=4.5}
ggplot(prepost.long, aes(
  x = time, y = score, colour = drug,
  group = subject
)) + geom_point() + geom_line()
```

   


## Exponential growth data


* I made up these data:
```{r }
x <- 0:5
y <- c(2.0, 4.1, 8.2, 15.8, 31.6, 65.0)
grow <- data.frame(x, y)
grow
```

   


* Each $y$-value is approximately twice as big as the previous:
exponential growth.



## Scatter plot with line (bad)
```{r fig.height=4.5}
ggplot(grow, aes(x = x, y = y)) +
  geom_point() + geom_smooth(method = "lm")
```

   


## Use log scale for $y$-axis

```{r fig.height=4}
ggplot(grow, aes(x = x, y = y)) +
  geom_point() + coord_trans(y = "log10")
```

   

On this plot, trend is straight.



