
## Logistic regression


* When response variable is measured/counted, regression can work well.

* But what if response is yes/no, lived/died, success/failure?

* Model *probability* of success.

* Probability must be between 0 and 1; need method that ensures this.

* *Logistic regression* does this. In R, is a
*generalized linear model* with binomial "family": 
```{r, eval=F}
glm(y ~ x, family="binomial")
```


* Begin with simplest case.



## Packages
```{r, eval=F}
library(MASS)
library(tidyverse)
library(broom)
library(nnet)
library(conflicted)
```

   


## The rats, part 1


* Rats given dose of some poison; either live or die:

\small
```
dose status
0 lived
1 died
2 lived
3 lived
4 died
5 died
```

\normalsize

## Read in: 

```{r}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/rat.txt"
rats <- read_delim(my_url, " ")
rats
```
 


## Basic logistic regression


* Make response into a factor first:

\small
```{r}
rats2 <- rats %>% mutate(status = factor(status))
```
\normalsize
   


* then fit model:

\small
```{r, error=T}
status.1 <- glm(status ~ dose, family = "binomial", data = rats2)
```
\normalsize
   

   


## Output

\scriptsize
```{r}
summary(status.1)
```
\normalsize


## Interpreting the output


* Like (multiple) regression, get
tests of significance of individual $x$'s

* Here not significant (only 6 observations).

* "Slope" for dose is negative, meaning that as dose increases, probability of event modelled (survival) decreases.





## Output part 2: predicted survival probs
```{r }
p <- predict(status.1, type = "response")
cbind(rats, p)
```
 

## The rats, more


* More realistic: more rats at each dose (say 10).

* Listing each rat on one line makes a big data file.

* Use format below: dose, number of survivals, number of deaths.

```

dose lived died
0    10    0
1     7    3 
2     6    4 
3     4    6 
4     2    8 
5     1    9  

```


* 6 lines of data correspond to 60 actual rats.

* Saved in `rat2.txt`.



## These data

\footnotesize
```{r}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/rat2.txt"
rat2 <- read_delim(my_url, " ")
rat2
```
\normalsize


## Create response matrix:

- Each row contains *multiple* observations.
- Create *two-column* response:
  - \#survivals in first column, 
  - \#deaths in second.


\footnotesize
```{r }
response <- with(rat2, cbind(lived, died))
response
```
\normalsize

- Response is R `matrix`:

\footnotesize
```{r }
class(response)
```
\normalsize

     
## Fit logistic regression

- using response you just made:

```{r}
rat2.1 <- glm(response ~ dose,
  family = "binomial",
  data = rat2
)
```




## Output

\scriptsize
```{r}
summary(rat2.1)
```
\normalsize


## Predicted survival probs
```{r }
p <- predict(rat2.1, type = "response")
cbind(rat2, p)
```
 


## Comments



* Significant effect of dose. 

* Effect of larger dose is to *decrease* survival probability
("slope" negative; also see in decreasing predictions.)


## Multiple logistic regression


* With more than one $x$, works much like multiple regression.

* Example: study of patients with blood poisoning severe enough to warrant surgery. Relate survival to other potential risk factors.

* Variables, 1=present, 0=absent:


  * survival (death from sepsis=1), response
  * shock
  * malnutrition
  * alcoholism
  * age (as numerical variable)
  * bowel infarction


* See what relates to death.



## Read in data
```{r size="footnotesize"}
my_url <- 
  "http://www.utsc.utoronto.ca/~butler/d29/sepsis.txt"
sepsis <- read_delim(my_url, " ")
```
 


## The data
```{r size="footnotesize"}
sepsis
```

   


## Fit model
```{r }
sepsis.1 <- glm(death ~ shock + malnut + alcohol + age +
  bowelinf,
family = "binomial",
data = sepsis
)
```

   


## Output part 1
```{r size="footnotesize"}
tidy(sepsis.1)
```
 



* All P-values fairly small

* but `malnut` not significant: remove.



## Removing `malnut`
```{r size="footnotesize"}
sepsis.2 <- update(sepsis.1, . ~ . - malnut)
tidy(sepsis.2)
```
 



* Everything significant now.



## Comments


* Most of the original $x$'s helped predict death. Only `malnut` seemed not to add anything.

* Removed `malnut` and tried again.

* Everything remaining is significant (though `bowelinf`
actually became *less* significant).

* All coefficients are *positive*, so having any of the risk
factors (or being older)
*increases* risk of death.  




## Predictions from model without "malnut"


* A few chosen at random:

\normalsize
```{r size="footnotesize"}
sepsis.pred <- predict(sepsis.2, type = "response")
d <- data.frame(sepsis, sepsis.pred)
myrows <- c(4, 1, 2, 11, 32)
slice(d, myrows)
```
\normalsize

## Comments 

* Survival chances pretty good if no risk factors, though decreasing with age.

* Having more than one risk factor reduces survival chances dramatically.

* Usually good job of predicting survival; sometimes death predicted to survive.



## Assessing proportionality of odds for age


* An assumption we made is that log-odds of survival depends
linearly on age.

* Hard to get your head around, but 
basic idea is that survival chances go continuously up (or down)
with age, instead of (for example) going up and then down.

* In this case, seems reasonable, but should check:


## Residuals vs.\ age
```{r virtusentella,fig.height=3.4, warning=F}
ggplot(augment(sepsis.2), aes(x = age, y = .resid)) +
  geom_point()
```
 
## Comments

* No apparent problems overall.

* Confusing "line" across: no risk factors, survived. 



## Probability and odds

* For probability $p$, odds is $p/(1-p)$:


  \begin{tabular}{rrrl}
      \hline
      Prob.\ & Odds & log-odds & in words\\
      \hline
      0.5 & $0.5/0.5=1/1=1.00$ & $0.00$ &  ``even money''\\
      0.1 & $0.1/0.9=1/9=0.11$ & $-2.20$ & ``9 to 1''\\
      0.4 & $0.4/0.6=1/1.5=0.67$ & $-0.41$ & ``1.5 to 1''\\
      0.8 & $0.8/0.2=4/1=4.00$ & $1.39$ & ``4 to 1 on''\\
      \hline
    \end{tabular}


* Gamblers use odds: if you win at 9 to 1 odds, get original
stake back plus 9 times the stake.

* Probability has to be between 0 and 1

* Odds between 0 and infinity

* *Log*-odds can be anything: any log-odds corresponds to
valid probability.



## Odds ratio


* Suppose 90 of 100 men drank wine last week, but only 20 of 100 women.

* Prob of man drinking wine $90/100=0.9$, woman $20/100=0.2$.

* Odds of man drinking wine $0.9/0.1=9$, woman $0.2/0.8=0.25$.

* Ratio of odds is $9/0.25=36$.

* Way of quantifying difference between men and women: ``odds of
drinking wine 36 times larger for males than females''. 



## Sepsis data again


* Recall prediction of probability of death from risk factors:
```{r size="small"}
sepsis.2.tidy <- tidy(sepsis.2)
sepsis.2.tidy
```

     


* Slopes in column `estimate`.



## Multiplying the odds


* Can interpret slopes by taking "exp" of them. We ignore intercept.

```{r expo}
sepsis.2.tidy %>% 
  mutate(exp_coeff=exp(estimate)) %>% 
  select(term, exp_coeff)
```

## Interpretation

\small
```{r ref.label="expo", echo=F}
```
\normalsize


* These say ``how much do you *multiply* odds of death by
for increase of 1 in corresponding risk factor?'' Or, what is odds
ratio for that factor being 1 (present) vs.\ 0 (absent)?

* Eg.\ being alcoholic vs.\ not increases odds of death by 24 times

* One year older multiplies odds by about 1.1 times. Over 40 years,
about  $1.09^{40}=31$ times. 



## Odds ratio and relative risk


* **Relative risk** is ratio of probabilities.

* Above: 90 of 100 men (0.9) drank wine, 20 of 100 women (0.2).

* Relative risk 0.9/0.2=4.5. (odds ratio was 36).

* When probabilities small, relative risk and odds ratio similar.

* Eg.\ prob of man having disease 0.02, woman 0.01.

* Relative risk $0.02/0.01=2$.

## Odds ratio vs.\ relative risk

- Odds for men and for women:

```{r }
(od1 <- 0.02 / 0.98) # men
(od2 <- 0.01 / 0.99) # women
```

-  Odds ratio 

```{r }
od1 / od2
```


- Very close to relative risk of 2.


## More than 2 response categories


* With 2 response categories, model the probability of one, and prob of other is one minus that. So doesn't matter which category you model.

* With more than 2 categories, have to think more carefully about the categories: are they


* *ordered*: you can put them in a natural order (like low, medium, high)

* *nominal*: ordering the categories doesn't make sense (like red, green, blue).


* R handles both kinds of response; learn how.



## Ordinal response: the miners


* 
Model probability of being in given category *or lower*.

* Example: coal-miners often suffer disease pneumoconiosis. Likelihood of disease believed to be greater 
among miners who have worked longer. 

* Severity of disease measured on categorical scale: none,
moderate, 3 severe.

## Miners data

* Data are frequencies:

```
Exposure None Moderate Severe
5.8       98      0       0
15.0      51      2       1
21.5      34      6       3
27.5      35      5       8
33.5      32     10       9
39.5      23      7       8
46.0      12      6      10
51.5       4      2       5
```




## Reading the data

Data in aligned columns with more than one space between, so: 

\small
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/miners-tab.txt"
freqs <- read_table(my_url)
```
\normalsize


## The data
```{r }
freqs
```

   


## Tidying and row proportions
```{r }
freqs %>%
  gather(Severity, Freq, None:Severe) %>%
  group_by(Exposure) %>%
  mutate(proportion = Freq / sum(Freq)) -> miners
```

   


## Result

\small
```{r }
miners
```
\normalsize
     


## Plot proportions against exposure

\small
```{r fig.height=3.5, message=F}
ggplot(miners, aes(x = Exposure, y = proportion,
                   colour = Severity)) + 
  geom_point() + geom_smooth(se = F)
```
\normalsize


## Reminder of data setup
\footnotesize
```{r }
miners
```

   
\normalsize

## Creating an ordered factor


* Problem: on plot, `Severity` categories in \emph{wrong
order}. 

* *In the data frame*, categories in *correct* order.

* Package `forcats` (in `tidyverse`) has functions for creating factors to specifications.

* `fct_inorder` takes levels *in order they appear in data*: 
```{r }
miners %>%
  mutate(sev_ord = fct_inorder(Severity)) -> miners
```

     


* To check:
```{r }
levels(miners$sev_ord)
```

   


## New data frame

\small
```{r }
miners
```
\normalsize
 

## Improved plot
```{r fig.height=3.25, message=FALSE}
ggplot(miners, aes(x = Exposure, y = proportion,
                   colour = sev_ord)) + 
  geom_point() + geom_smooth(se = F)
```

   

## Fitting ordered logistic model

Use function `polr` from package `MASS`. Like `glm`.
```{r }
sev.1 <- polr(sev_ord ~ Exposure,
  weights = Freq,
  data = miners
)
```
 


## Output: not very illuminating
\scriptsize
```{r }
summary(sev.1)
```
\normalsize
   

## Does exposure have an effect?
Fit model without `Exposure`, and compare
using `anova`. Note `1` for model with just intercept:

```{r echo=F}
w <- getOption("width")
options(width = w - 20)
```

 

\small
```{r}
sev.0 <- polr(sev_ord ~ 1, weights = Freq, data = miners)
anova(sev.0, sev.1)
```
\normalsize

Exposure definitely has effect on severity of disease. 


## Another way


* What (if anything) can we drop from model with `exposure`?
```{r }
drop1(sev.1, test = "Chisq")
```

     


* Nothing. Exposure definitely has effect.



## Predicted probabilities

Make new data frame out of all the exposure values (from original data
frame), and predict from that:

```{r echo=F}
options(width = 70)
```

 
```{r size="small"}
sev.new <- tibble(Exposure = freqs$Exposure)
pr <- predict(sev.1, sev.new, type = "p")
miners.pred <- cbind(sev.new, pr)
miners.pred
```
 

## Comments


* Model appears to match data: as exposure goes up, prob of None
goes down, Severe goes up (sharply for high exposure).

* Like original data frame, this one nice to look at but
*not tidy*. We want to make graph, so tidy it.

* Also want the severity values in right order.

* Usual `gather`, plus a bit:
```{r size="small"}
miners.pred %>%
  gather(Severity, probability, -Exposure) %>%
  mutate(sev_ord = fct_inorder(Severity)) -> preds
```

     



## Some of the gathered predictions

\footnotesize
```{r}
preds %>% slice(1:15)
```
\normalsize
   


## Plotting predicted and observed proportions


* Plot:
  *  predicted probabilities, lines (shown) joining points (not shown)

  * data, just the points. 


* Unfamiliar process: data from two *different* data frames:

\small
```{r }
g <- ggplot(preds, aes(
  x = Exposure, y = probability,
  colour = sev_ord
)) + geom_line() +
  geom_point(data = miners, aes(y = proportion))
```
\normalsize
   
* Idea: final `geom_point` uses data in `miners`
rather than `preds`, $y$-variable for plot is `proportion`
from that data frame, but $x$-coordinate is `Exposure`, as it
was before, and `colour` is `Severity` as before. The
final `geom_point` "inherits" from the first `aes`
as needed.




## The plot: data match model
```{r fig.height=3.8}
g
```

## Unordered responses


* With unordered (nominal) responses, can use *generalized logit*.

* Example: 735 people, record age and sex (male 0, female 1), which of 3 brands of some product preferred.

* Data in `mlogit.csv` separated by commas (so
`read_csv` will work):
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/mlogit.csv"
brandpref <- read_csv(my_url)
```
 




## The data
```{r }
brandpref
```

   


## Bashing into shape, and fitting model


* `sex` and `brand` not meaningful as numbers, so
turn into factors:
```{r }
brandpref <- brandpref %>%
  mutate(sex = factor(sex)) %>%
  mutate(brand = factor(brand))
```
 

* We use `multinom` from package `nnet`. Works
like `polr`.

```{r }
brands.1 <- multinom(brand ~ age + sex, data = brandpref)
```
 


## Can we drop anything?


* Unfortunately `drop1` seems not to work:
```{r, error=TRUE}
drop1(brands.1, test = "Chisq", trace = 0)
```

   

* so fall back on fitting model without what you want to test, and
comparing using `anova`.    



## Do age/sex help predict brand? 1/2

Fit models without each of age and sex:
```{r }
brands.2 <- multinom(brand ~ age, data = brandpref)
brands.3 <- multinom(brand ~ sex, data = brandpref)
```
 


## Do age/sex help predict brand? 2/2

\scriptsize
```{r}
anova(brands.2, brands.1)
anova(brands.3, brands.1)
```
\normalsize


## Do age/sex help predict brand? 3/3


* `age` definitely significant (second `anova`)

* `sex` seems significant also (first `anova`)

* Keep both.




## Another way to build model


* Start from model with everything and run `step`:

\footnotesize
```{r }
step(brands.1, trace = 0)
```
\normalsize
     

* Final model contains both `age` and `sex` so neither
could be removed.



## Predictions: all possible combinations

Create data frame with various age and sex:

\footnotesize
```{r}
ages <- c(24, 28, 32, 35, 38)
sexes <- factor(0:1)
new <- crossing(age = ages, sex = sexes)
new
```
\normalsize



## Making predictions
```{r }
p <- predict(brands.1, new, type = "probs")
probs <- cbind(new, p)
```

or

```{r}
p %>% as_tibble() %>% 
  bind_cols(new) -> probs
```

   


## The predictions

\small
```{r }
probs
```
\normalsize
   


* Young males (`sex=0`) prefer brand 1, 
but older males prefer brand 3.

* Females similar, but like brand 1 less and
brand 2 more.



## Making a plot


* Plot fitted probability against age, distinguishing brand by
colour and gender by plotting symbol.

* Also join points by lines, and distinguish lines by gender. 

* I thought about facetting, but this seems to come out clearer.

* First need tidy data frame, by familiar process:
```{r }
probs %>%
  gather(brand, probability, -(age:sex)) -> probs.long
```

## The tidy data (random sample of rows)

\small
```{r}
probs.long %>% sample_n(10)
```
\normalsize

## The plot
```{r fig.height=2.8}
ggplot(probs.long, aes(
  x = age, y = probability,
  colour = brand, shape = sex
)) +
  geom_point() + geom_line(aes(linetype = sex))
```

   

## Digesting the plot


* Brand vs.\ age: younger people (of both genders) prefer brand
1, but older people (of both genders) prefer brand 3. (Explains
significant age effect.)

* Brand vs.\ sex: females (dashed) like brand 1 less than males
(solid), like brand 2 more (for all ages). 

* Not much brand difference between genders (solid and dashed
lines of same colours close), but enough to be significant.

* Model didn't include interaction, so modelled effect of gender
on brand same for each age, modelled effect of age same for each
gender. 


## Alternative data format

Summarize all people of same brand preference, same sex, same age on one line of data file with frequency on end:



```
1 0 24 1
1 0 26 2
1 0 27 4
1 0 28 4
1 0 29 7
1 0 30 3
...
```

Whole data set in 65 lines not 735! But how?


## Getting alternative data format
```{r }
brandpref %>%
  group_by(age, sex, brand) %>%
  summarize(Freq = n()) %>%
  ungroup() -> b
b %>% slice(1:6)
```

   

## Fitting models, almost the same


* Just have to remember `weights` to incorporate
frequencies.

* Otherwise `multinom` assumes you have just 1 obs
on each line!

* Again turn (numerical) `sex` and `brand` into factors:

\footnotesize
```{r, results="hide"}
b %>%
  mutate(sex = factor(sex)) %>%
  mutate(brand = factor(brand)) -> bf
b.1 <- multinom(brand ~ age + sex, data = bf, weights = Freq)
b.2 <- multinom(brand ~ age, data = bf, weights = Freq)
```
\normalsize



## P-value for `sex` identical

\footnotesize
```{r}
anova(b.2, b.1)
```
\normalsize

Same P-value as before, so we haven't changed anything important.


## Including data on plot


* Everyone's age given as whole
number, so maybe not too many different ages with sensible amount
of data at each:

\scriptsize
```{r}
b %>%
  group_by(age) %>%
  summarize(total = sum(Freq))
```
\normalsize


## Comments and next


* Not great (especially at low end), but live with it.

* Need proportions of frequencies in each brand for each
age-gender combination. Mimic what we did for miners:
```{r spal-b}
b %>%
  group_by(age, sex) %>%
  mutate(proportion = Freq / sum(Freq)) -> brands
```

     



## Checking proportions for age 32

\small
```{r }
brands %>% filter(age == 32)
```
\normalsize
   



* First three proportions (males) add up to 1.

* Last three proportions (females) add up to 1.

* So looks like proportions of right thing.



## Attempting plot


* Take code from previous plot and:


* remove `geom_point` for fitted values

* add `geom_point` with correct `data=` and
`aes` to plot data.

```{r fig.height=3.2}
g <- ggplot(probs.long, aes(
  x = age, y = probability,
  colour = brand, shape = sex
)) +
  geom_line(aes(linetype = sex)) +
  geom_point(data = brands, aes(y = proportion))
```

     


* Data seem to correspond more or less to fitted curves:



## The plot
```{r fig.height=3.7}
g
```

   


## But\ldots


* Some of the plotted points based on a lot of people, and some
only a few.

* Idea: make the *size* of plotted point bigger if point
based on a lot of people (in `Freq`).

* Hope that larger points then closer to predictions.

* Code:

\footnotesize
```{r }
g <- ggplot(probs.long, aes(
  x = age, y = probability,
  colour = brand, shape = sex
)) +
  geom_line(aes(linetype = sex)) +
  geom_point(
    data = brands,
    aes(y = proportion, size = Freq)
  )
```
\normalsize
     



## The plot
```{r fig.height=3.7}
g
```

   


## Trying interaction between age and gender

```{r echo=F}
options(width = 60)
```

   
\scriptsize
```{r }
b.4 <- update(b.1, . ~ . + age:sex)
anova(b.1, b.4)
```

   
\normalsize

* No evidence that effect of age on brand preference differs for
the two genders.
