
## Multivariate analysis of variance


* Standard ANOVA has just one response variable.

* What if you have more than one response?

* Try an ANOVA on each response separately.

* But might miss some kinds of interesting dependence between the responses that distinguish the groups.



## Packages
```{r, eval=F }
library(car)
library(tidyverse)
```

   


## Small example


* Measure yield and seed weight of plants grown under 2 conditions: low and high amounts of fertilizer.

* Data (fertilizer, yield, seed weight):
```{r }
url <- "http://www.utsc.utoronto.ca/~butler/d29/manova1.txt"
hilo <- read_delim(url, " ")
```
 

* 2 responses, yield and seed weight.



## The data
```{r }
hilo
```

   


## Boxplot for yield for each fertilizer group
```{r ferto,size="small",fig.height=3}
ggplot(hilo, aes(x = fertilizer, y = yield)) + geom_boxplot()
```
 
Yields overlap for fertilizer groups.


## Boxplot for weight for each fertilizer group
```{r casteldisangro,size="small",fig.height=3}
ggplot(hilo, aes(x = fertilizer, y = weight)) + geom_boxplot()
```
 

Weights overlap for fertilizer groups.


## ANOVAs for yield and weight

\small
```{r }
hilo.y <- aov(yield ~ fertilizer, data = hilo)
summary(hilo.y)
hilo.w <- aov(weight ~ fertilizer, data = hilo)
summary(hilo.w)
```
 
\normalsize

Neither response depends significantly on fertilizer. But\ldots


## Plotting both responses at once

- Have two response variables (not more), so can plot the
response variables against *each other*, labelling points by
which fertilizer group they're from.

\footnotesize

- First, create data frame with points $(31,14)$ and $(38,10)$ (why? Later):

```{r size="footnotesize"}
d <- tribble(
  ~line_x, ~line_y,
  31, 14,
  38, 10
)
```

- Then plot data as points, and add line through points in `d`:

```{r }
g <- ggplot(hilo, aes(x = yield, y = weight,
                      colour = fertilizer)) + geom_point() +
  geom_line(data = d,
            aes(x = line_x, y = line_y, colour = NULL))
```

\normalsize

## The plot
```{r charlecombe, echo=F, fig.height=4}
g
```

## Comments

- Graph construction:
  - Joining points in `d` by line.
  - `geom_line` 
inherits `colour` from `aes` in `ggplot`.
  - Data frame `d` has
no `fertilizer` (previous `colour`), so have to unset.

- Results:
  * High-fertilizer plants have both yield and weight high.

  * True even though no sig difference in yield or weight individually.

  * Drew line separating highs from lows on plot.



## MANOVA finds multivariate differences


* Is difference found by diagonal line significant? MANOVA finds out.

\footnotesize

```{r, echo=FALSE}
options(width = 60)
```


```{r}
response <- with(hilo, cbind(yield, weight))
hilo.1 <- manova(response ~ fertilizer, data = hilo)
summary(hilo.1)
```

\normalsize 

* Yes! Difference between groups is *diagonally*, not just up/down
(weight) or left-right (yield). The *yield-weight combination* matters.




## Strategy



* Create new response variable by gluing together columns of
responses, using `cbind`.

* Use `manova` with new response, looks like `lm` otherwise.

* With more than 2 responses, cannot draw graph. What then?

* If MANOVA test significant, cannot use Tukey. What then?

* Use *discriminant analysis* (of which more later).




## Another way to do MANOVA
Install (once) and load package `car`:
```{r eval=F}
library(car)
```
 


## Another way\ldots
  
```{r size="footnotesize"}
hilo.2.lm <- lm(response ~ fertilizer, data = hilo)
hilo.2 <- Manova(hilo.2.lm)
hilo.2
```

   


* Same result as small-m `manova`.

* `Manova` will also do *repeated measures*, coming up
later.




## Another example: peanuts


*  Three different varieties
of peanuts (mysteriously, 5, 6 and 8) planted in two different
locations.

* Three response variables: `y`, `smk` and
`w`.

```{r size="footnotesize"}
u <- "http://www.utsc.utoronto.ca/~butler/d29/peanuts.txt"
peanuts.orig <- read_delim(u, " ")
```
 
## The data

\small
```{r}
peanuts.orig
```

\normalsize

## Setup for analysis

```{r }
peanuts <- peanuts.orig %>%
  mutate(
    location = factor(location),
    variety = factor(variety)
  )
response <- with(peanuts, cbind(y, smk, w))
head(response)
```
 


## Analysis (using `Manova)`

\small
```{r}
peanuts.1 <- lm(response ~ location * variety, data = peanuts)
peanuts.2 <- Manova(peanuts.1)
peanuts.2
```
\normalsize 
 
## Comments


* Interaction not quite significant, but main effects are.

* Combined response variable `(y,smk,w)` definitely depends
on location and on variety

* Weak dependence of `(y,smk,w)` on the location-variety *combination.*

* Understanding that dependence beyond our scope right now.



