
## Multivariate analysis of variance


* Standard ANOVA has just one response variable.

* What if you have more than one response?

* Try an ANOVA on each response separately.

* But might miss some kinds of interesting dependence between the responses that distinguish the groups.



## Packages
```{r bManova-1, results='hide', message=FALSE}
library(car) # may need to install first
library(tidyverse)
library(MVTests) # also may need to install
```

   


## Small example


* Measure yield and seed weight of plants grown under 2 conditions: low and high amounts of fertilizer.

* Data (fertilizer, yield, seed weight):
```{r bManova-2 }
url <- "http://ritsokiguess.site/datafiles/manova1.txt"
hilo <- read_delim(url, " ")
```
 

* 2 responses, yield and seed weight.



## The data
```{r bManova-3 }
hilo
```

   


## Boxplot for yield for each fertilizer group
```{r ferto,size="small",fig.height=3}
ggplot(hilo, aes(x = fertilizer, y = yield)) + geom_boxplot()
```
 
Yields overlap for fertilizer groups.


## Boxplot for weight for each fertilizer group
```{r casteldisangro,size="small",fig.height=3}
ggplot(hilo, aes(x = fertilizer, y = weight)) + geom_boxplot()
```
 

Weights overlap for fertilizer groups.


## ANOVAs for yield and weight

\small
```{r bManova-4 }
hilo.y <- aov(yield ~ fertilizer, data = hilo)
summary(hilo.y)
hilo.w <- aov(weight ~ fertilizer, data = hilo)
summary(hilo.w)
```
 
\normalsize

Neither response depends significantly on fertilizer. But\ldots


## Plotting both responses at once

- Have two response variables (not more), so can plot the
response variables against *each other*, labelling points by
which fertilizer group they're from.

\footnotesize

- First, create data frame with points $(31,14)$ and $(38,10)$ (why? Later):

```{r bManova-5, size="footnotesize"}
d <- tribble(
  ~line_x, ~line_y,
  31, 14,
  38, 10
)
```

- Then plot data as points, and add line through points in `d`:

```{r bManova-6 }
ggplot(hilo, aes(x = yield, y = weight,
                      colour = fertilizer)) + geom_point() +
  geom_line(data = d,
            aes(x = line_x, y = line_y, colour = NULL)) -> g
```

\normalsize

## The plot
```{r charlecombe, echo=F, fig.height=4}
g
```

## Comments

- Graph construction:
  - Joining points in `d` by line.
  - `geom_line` 
inherits `colour` from `aes` in `ggplot`.
  - Data frame `d` has
no `fertilizer` (previous `colour`), so have to unset.

- Results:
  * High-fertilizer plants have both yield and weight high.

  * True even though no sig difference in yield or weight individually.

  * Drew line separating highs from lows on plot.



## MANOVA finds multivariate differences


* Is difference found by diagonal line significant? MANOVA finds out.

\footnotesize

```{r bManova-7, echo=FALSE}
options(width = 60)
```


```{r bManova-8}
response <- with(hilo, cbind(yield, weight))
hilo.1 <- manova(response ~ fertilizer, data = hilo)
summary(hilo.1)
```

\normalsize 

* Yes! Difference between groups is *diagonally*, not just up/down
(weight) or left-right (yield). The *yield-weight combination* matters.




## Strategy



* Create new response variable by gluing together columns of
responses, using `cbind`.

* Use `manova` with new response, looks like `lm` otherwise.

* With more than 2 responses, cannot draw graph. What then?

* If MANOVA test significant, cannot use Tukey. What then?

* Use *discriminant analysis* (of which more later).




## Another way to do MANOVA

using `Manova` from package `car`:

```{r, include=FALSE}
w <- getOption("width")
options(width = 132)
```


\tiny
```{r bManova-10}
hilo.2.lm <- lm(response ~ fertilizer, data = hilo)
hilo.2 <- Manova(hilo.2.lm)
summary(hilo.2)
```
\normalsize

```{r, include=FALSE}
options(width = w)
```


## Comments

* Same result as small-m `manova`.

* `Manova` will also do *repeated measures*, coming up
later.

## Assumptions

- normality of each response variable within each treatment group
  - this is actually *multivariate* normality, with correlations

- equal spreads: each response variable has same variances and correlations (with other response variables) within each treatment group. Here:
  - yield has same spread for low and high fertilizer
  - weight has same spread for low and high fertilizer
  - correlation between yield and weight is same for low and high fertilizer
- test equal spread using Box's $M$ test
  - a certain amount of unequalness is OK, so only a concern if P-value from $M$-test is very small (eg. less than 0.001).
  
## Assumptions for yield-weight data

For normal quantile plots, need "extra-long" with all the data values in one column:

```{r}
hilo %>% 
  pivot_longer(-fertilizer, names_to = "xname", 
               values_to = "xvalue") %>% 
  ggplot(aes(sample = xvalue)) + stat_qq() + 
    stat_qq_line() +
    facet_grid(xname ~ fertilizer, scales = "free") -> g
```

There are only four observations per response variable - treatment group combination, so graphs are not very informative (over):

## The plots

```{r, fig.height=4}
g
```

## Box M test

- Make sure package `MVTests` loaded first. 
- inputs:
  - the response matrix (or, equivalently, the response-variable columns from your dataframe)
  - the column with the grouping variable in it (most easily gotten with `$`).

```{r}
hilo %>% select(yield, weight) -> numeric_values
summary(BoxM(numeric_values, hilo$fertilizer))
```

- No problem at all with unequal spreads.

## Another example: peanuts


*  Three different varieties
of peanuts (mysteriously, 5, 6 and 8) planted in two different
locations.

* Three response variables: `y`, `smk` and
`w`.

```{r bManova-11, size="footnotesize"}
u <- "http://ritsokiguess.site/datafiles/peanuts.txt"
peanuts.orig <- read_delim(u, " ")
```
 
## The data

\small
```{r bManova-12}
peanuts.orig
```

\normalsize

## Setup for analysis

```{r bManova-13 }
peanuts.orig %>%
  mutate(
    location = factor(location),
    variety = factor(variety)
  ) -> peanuts
response <- with(peanuts, cbind(y, smk, w))
head(response)
```



## Analysis (using `Manova`)

```{r, include=FALSE}
w <- getOption("width")
options(width = 100)
```


\tiny
```{r bManova-14}
peanuts.1 <- lm(response ~ location * variety, data = peanuts)
peanuts.2 <- Manova(peanuts.1)
summary(peanuts.2)
```
\normalsize 
 
```{r, include=FALSE}
options(width = w)
```
 
 
## Comments


* Interaction not quite significant, but main effects are.

* Combined response variable `(y,smk,w)` definitely depends
on location and on variety

* Weak dependence of `(y,smk,w)` on the location-variety *combination.*

* Understanding that dependence beyond our scope right now.

## Normal quantile plots

Setup for normal quantile plots

```{r}
peanuts %>% 
  pivot_longer(y:w, names_to = "xname", 
               values_to = "xvals") -> d
```

Make the plots (save to show on next pages)

```{r}
ggplot(d, aes(sample = xvals)) + stat_qq() + 
  stat_qq_line() +
  facet_grid(xname ~ location, scales = "free") -> g1

ggplot(d, aes(sample = xvals)) + stat_qq() + 
  stat_qq_line() +
  facet_grid(xname ~ variety, scales = "free") -> g2
```

## The plot for location

```{r, fig.height=4}
g1
```

## The plot for variety

```{r, fig.height=4}
g2
```


## Comments

- this time there are only six observations per location  and four per variety, so normality is still difficult to be confident about

- `y` at location 1 seems to be the worst for normality (long tails / outliers), and maybe `y` at location 2 is skewed left, but the others are not bad

- there is some evidence of unequal spread (slopes of lines), but is it bad enough to worry about? (Box M-test, over).


## Box's M tests

- One for location, one for variety:

```{r}
summary(BoxM(response, peanuts$location))
summary(BoxM(response, peanuts$variety))
```

- Neither of these P-values is low enough to worry about. (Remember, the P-value here has to be *really* small to indicate a problem.)


