
## Multi-way frequency analysis


* A study of gender and eyewear-wearing finds the following frequencies:
\begin{tabular}{lrrr}
\hline
Gender & Contacts & Glasses & None \\
\hline
Female & 121 & 32 & 129 \\
Male & 42 & 37 & 85\\
\hline
\end{tabular}

* Is there association between eyewear and gender?

* Normally answer this with chisquare test (based on observed and expected frequencies from null hypothesis of no association).

* Two categorical variables and a frequency.

* We assess in way that generalizes to more categorical variables.




##  The data file

\footnotesize
```
gender contacts glasses none
female 121      32      129
male   42       37      85
```
\normalsize


* This is *not tidy!*

* Two variables are gender and *eyewear*, and those numbers
all frequencies.

\footnotesize
```{r message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/eyewear.txt"
(eyewear <- read_delim(my_url, " "))
```
\normalsize
   


##  Tidying the data
```{r size="footnotesize"}
eyewear %>%
  pivot_longer(contacts:none, names_to="eyewear", 
               values_to="frequency") -> eyes
eyes
```

## Making tidy data back into a table

- use `pivot_longer`
- or this (we use it again later):

```{r}
xt <- xtabs(frequency ~ gender + eyewear, data = eyes)
xt
```

   

##  Modelling


* Predict frequency from other factors and combos. 
- `glm`
with `poisson` family.

```{r }
eyes.1 <- glm(frequency ~ gender * eyewear,
  data = eyes,
  family = "poisson"
)
summary(eyes.1)
```



* Called **log-linear model**.



##  What can we get rid of?

\small    
```{r }
drop1(eyes.1, test = "Chisq")
```
 
nothing!


##  Conclusions


* `drop1` says what we can remove at this
step. Significant = must stay.

* Cannot remove anything.

* Frequency depends on `gender-wear`
*combination*, cannot be simplified further.

* Gender and eyewear are *associated*.

* Stop here.


##  `prop.table`
Original table:

\footnotesize
```{r }
xt
```
\normalsize
 

Calculate eg. row proportions like this:

\small
```{r }
prop.table(xt, margin = 1)
```
\normalsize

## Comments

* `margin` says what to make add to 1.

* More females wear contacts and more males wear glasses.




##  No association


* Suppose table had been as shown below:

\scriptsize
```{r, message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/eyewear2.txt"
eyewear2 <- read_table(my_url)
eyewear2
eyes2 <- eyewear2 %>% pivot_longer(contacts:none, names_to = "eyewear", values_to = "frequency")
eyes2
xt2 <- xtabs(frequency ~ gender + eyewear, data = eyes2)
xt2
prop.table(xt2, margin = 1)
```
\normalsize
   
## Comments

* Females and males wear contacts and glasses *in same proportions*
   - though more females and more contact-wearers. 
- No
*association* between gender and eyewear.



##  Analysis for revised data
```{r}
eyes.2 <- glm(frequency ~ gender * eyewear,
  data = eyes2,
  family = "poisson"
)
drop1(eyes.2, test = "Chisq")
```

   

No longer any association. Take out interaction.


##  No interaction
\small
```{r }
eyes.3 <- update(eyes.2, . ~ . - gender:eyewear)
drop1(eyes.3, test = "Chisq")
```
\normalsize
   


* More females (gender effect)

* more contact-wearers (eyewear
effect)

* no association (no interaction). 



##  Chest pain, being overweight and being a smoker


* In a hospital emergency department, 176 subjects who attended
for acute chest pain took part in a study.

* Each subject had a
normal or abnormal electrocardiogram reading (ECG), were
overweight (as judged by BMI) or not, and were a smoker or
not.

* How are these three variables related, or not?



## The data
In modelling-friendly format:

```

ecg bmi smoke count
abnormal overweight yes 47
abnormal overweight no 10
abnormal normalweight yes 8 
abnormal normalweight no 6
normal overweight yes 25 
normal overweight no 15 
normal normalweight yes 35
normal normalweight no 30

```



##  First step
\small
```{r, message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/ecg.txt"
chest <- read_delim(my_url, " ")
chest
chest.1 <- glm(count ~ ecg * bmi * smoke,
  data = chest,
  family = "poisson"
)
drop1(chest.1, test = "Chisq")
```
\normalsize
   

That 3-way interaction comes out.


##  Removing the 3-way interaction
\small
```{r}
chest.2 <- update(chest.1, . ~ . - ecg:bmi:smoke)
drop1(chest.2, test = "Chisq")
```
\normalsize
   

At $\alpha=0.05$, `bmi:smoke` comes out.


##  Removing `bmi:smoke`
\small
```{r}
chest.3 <- update(chest.2, . ~ . - bmi:smoke)
drop1(chest.3, test = "Chisq")
```
\normalsize
   

`ecg:smoke` has become significant. So we have to stop.


## Understanding the final model


* Thinking of `ecg` as "response" that might depend on
anything else.

* What is associated with `ecg`? Both `bmi` on its
own and `smoke` on its own, but *not* the combination
of both.

* `ecg:bmi` table:
```{r }
xt <- xtabs(count ~ ecg + bmi, data = chest)
xt
```

```{r}
prop.table(xt, margin = 2)
```
     


* Most normal weight people have a normal ECG, but a
majority of overweight people have an *abnormal* ECG. That is,
knowing about BMI says something about likely ECG.



## `ecg:smoke`


* `ecg:smoke` table:
```{r }
xt <- xtabs(count ~ ecg + smoke, data = chest)
xt
```

```{r}
prop.table(xt, margin = 2)
```
   


* Most nonsmokers have a normal ECG, but smokers are about 50--50
normal and abnormal ECG.

* Don't look at `smoke:bmi` table since not significant. 

```{r}
xtabs(count ~ bmi + smoke, data = chest)
```


## Simpson's paradox: the airlines example
\begin{tabular}{|l|rr|rr|}
\hline
& \multicolumn{2}{|c|}{Alaska Airlines} & 
\multicolumn{2}{|c|}{America West}\\
Airport & On time & Delayed & On time & Delayed\\
\hline
Los Angeles & 497 & 62 & 694 & 117\\
Phoenix & 221 & 12 & 4840 & 415\\
San Diego & 212 & 20 & 383 & 65\\
San Francisco & 503 & 102 & 320 & 129 \\
Seattle & 1841 & 305 & 201 & 61\\
\hline
Total & 3274 & 501 & 6438 & 787\\
\hline
\end{tabular}
\vspace{2ex}
Use `status` as variable name for "on time/delayed".


* Alaska: 13.3\% flights delayed ($501/(3274+501)$).

* America West: 10.9\% ($787/(6438+787)$).

* America West more punctual, right?



##  Arranging the data


* Can only have single thing in columns, so we have to construct
column names like this:
\small

```

airport    aa_ontime aa_delayed aw_ontime aw_delayed
LosAngeles   497          62       694        117
Phoenix      221          12      4840        415
SanDiego     212          20       383         65
SanFrancisco 503         102       320        129
Seattle     1841         305       201         61

```

\normalsize

- Read in:

```{r, message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/airlines.txt"
airlines <- read_table2(my_url)
airlines
```

## Tidying 

the modern way

```{r}
airlines %>% 
   pivot_longer(-airport, names_to = c("airline", "status"), names_sep = "_", values_to = "freq") -> punctual
```


* Some tidying gets us the right layout, with frequencies all in
one column and the airline and delayed/on time status separated out: 
```{r message=F, size="small"}
airlines %>%
  gather(line.status, freq, contains("_")) %>%
  separate(line.status, c("airline", "status")) -> punctual
punctual
```

- See how this works by running it one line at a time.
     


##  The data frame `punctual`

\scriptsize
```{r echo=F}
punctual
```
\normalsize
   


##  Proportions delayed by airline


* Two-step process: get appropriate subtable:

\small
```{r }
xt <- xtabs(freq ~ airline + status, data = punctual)
xt
```
\normalsize
     


* and then calculate appropriate proportions:

\small
```{r }
prop.table(xt, margin = 1)
```
\normalsize

* More of Alaska Airlines' flights delayed (13.3\% vs.\ 10.9\%).



##  Proportion delayed by airport, for each airline

\footnotesize
```{r}
xt <- xtabs(freq ~ airline + status + airport, data = punctual)
xp <- prop.table(xt, margin = c(1, 3))
xp
ftable(xp,
  row.vars = c("airport", "airline"),
  col.vars = "status"
)
```
\normalsize
   

##  Simpson's Paradox

\begin{tabular}{|l|rr|}
\hline
Airport & Alaska & America West\\
\hline
Los Angeles & 11.4 & 14.4\\
Phoenix & 5.2 & 7.9\\
San Diego & 8.6 & 14.5\\
San Francisco & 16.9 & 28.7\\
Seattle & 14.2 & 23.2 \\
\hline
Total & 13.3 & 10.9 \\
\hline
\end{tabular}


* America West more punctual overall,

* but worse at *every single* airport!

* How is that possible?

* Log-linear analysis sheds some light.



## Model 1 and output
```{r}
punctual.1 <- glm(freq ~ airport * airline * status,
  data = punctual, family = "poisson"
)
drop1(punctual.1, test = "Chisq")
```
 


##  Remove 3-way interaction

\footnotesize
```{r}
punctual.2 <- update(punctual.1, ~ . - airport:airline:status)
drop1(punctual.2, test = "Chisq")
```
\normalsize

Stop here.


## Understanding the significance


* `airline:status`:

```{r }
xt <- xtabs(freq ~ airline + status, data = punctual)
prop.table(xt, margin = 1)
```

     


* More of Alaska Airlines' flights delayed overall.

* Saw this before.



## Understanding the significance (2)


* `airport:status`:
```{r }
xt <- xtabs(freq ~ airport + status, data = punctual)
xt
prop.table(xt, margin = 1)
```



* Flights into San Francisco (and maybe Seattle) are often late, and
flights into Phoenix are usually on time.

* Considerable variation among airports.



## Understanding the significance (3)


* `airport:airline`:
```{r }
xt <- xtabs(freq ~ airport + airline, data = punctual)
xt
prop.table(xt, margin = 2)
```



* What fraction of each airline's flights are to each airport.

* Most of Alaska Airlines' flights to Seattle and San Francisco.

* Most of America West's flights to Phoenix.



##  The resolution


* Most of America West's flights to Phoenix, where it is easy to
be on time.

* Most of Alaska Airlines' flights to San Francisco and Seattle,
where it is difficult to be on time.

* Overall comparison looks bad for Alaska because of this.

* But, *comparing like with like*, if you compare each
airline's performance *to the same airport*, Alaska does better.

* Aggregating over the very different airports was a (big)
mistake: that was the cause of the Simpson's paradox.

* Alaska Airlines is *more* punctual when you do the proper comparison.


##  Ovarian cancer: a four-way table


* Retrospective study of ovarian cancer done in 1973.

* Information about 299 women operated on for ovarian cancer 10 years previously.

* Recorded:


   * stage of cancer (early or advanced)

   * type of operation (radical or limited)

   * X-ray treatment received (yes or no)

   * 10-year survival (yes or no)


* Survival looks like response (suggests logistic
regression).

* Log-linear model finds any associations at all.


##  The data

after tidying:

\scriptsize

```

stage operation xray survival freq
early radical no no 10
early radical no yes 41
early radical yes no 17
early radical yes yes 64
early limited no no 1
early limited no yes 13
early limited yes no 3
early limited yes yes 9
advanced radical no no 38
advanced radical no yes 6
advanced radical yes no 64
advanced radical yes yes 11
advanced limited no no 3
advanced limited no yes 1
advanced limited yes no 13
advanced limited yes yes 5


```

\normalsize

## Reading in data

\small
```{r, message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/cancer.txt"
cancer <- read_delim(my_url, " ")
cancer
```
\normalsize


##  Model 1

hopefully looking familiar by now:
```{r}
cancer.1 <- glm(freq ~ stage * operation * xray * survival,
  data = cancer, family = "poisson"
)
```
 


##  Output 1

See what we can remove:

\scriptsize
```{r}
drop1(cancer.1, test = "Chisq")
```
\normalsize

Non-significant interaction can come out.

##  Model 2

\scriptsize
```{r}
cancer.2 <- update(cancer.1, . ~ . - stage:operation:xray:survival)
drop1(cancer.2, test = "Chisq")
```
\normalsize

Least significant term is `stage:xray:survival`: remove.

##  Take out `stage:xray:survival`

\scriptsize
```{r}
cancer.3 <- update(cancer.2, . ~ . - stage:xray:survival)
drop1(cancer.3, test = "Chisq")
```
\normalsize
   

`operation:xray:survival` comes out next.


##  Remove `operation:xray:survival`

\scriptsize
```{r}
cancer.4 <- update(cancer.3, . ~ . - operation:xray:survival)
drop1(cancer.4, test = "Chisq")
```
\normalsize
   


##  Comments


* `stage:operation:xray` has now become significant, so
won't remove that.

* Shows value of removing terms one at a time.

* There are no higher-order interactions containing both
`xray` and `survival`, so now we get to test (and
remove) `xray:survival`.



##  Remove `xray:survival`

\scriptsize
```{r}
cancer.5 <- update(cancer.4, . ~ . - xray:survival)
drop1(cancer.5, test = "Chisq")
```
\normalsize
   


##  Remove `stage:operation:survival`

\scriptsize
```{r}
cancer.6 <- update(cancer.5, . ~ . - stage:operation:survival)
drop1(cancer.6, test = "Chisq")
```
\normalsize
   


##  Last step?
Remove `operation:survival`.  

\footnotesize
```{r size="footnotesize"}
cancer.7 <- update(cancer.6, . ~ . - operation:survival)
drop1(cancer.7, test = "Chisq")
```
\normalsize
   
Finally done!


## Conclusions 


* What matters is things associated with `survival` (`survival` is
"response").

* Only significant such term is `stage:survival`:
```{r }
xt <- xtabs(freq ~ stage + survival, data = cancer)
xt
prop.table(xt, margin = 1)
```

     

* Most people in early stage of cancer survived, and most people
in advanced stage did not survive.

* This true *regardless* of type of operation or whether or
not X-ray treatment was received. These things have no impact on
survival. 



##  What about that other interaction?

\footnotesize
```{r }
xt <- xtabs(freq ~ operation + xray + stage, data = cancer)
ftable(prop.table(xt, margin = 3))
```
\normalsize
   



* Out of the people at each stage of cancer (since
`margin=3` and `stage` was listed 3rd).

* The association is between `stage` and `xray`
*only for those who had the limited operation*.

* For those who
had the radical operation, there was no association between
`stage` and `xray`.

* This is of less interest than associations with
`survival`. 



##  General procedure



* Start with "complete model" including all possible interactions.

* `drop1` gives highest-order interaction(s) remaining, remove least non-significant.

* Repeat as necessary until everything significant.

* Look at subtables of significant interactions.

* Main effects not usually very interesting.

* Interactions with "response" usually of most interest: show association with response.


   




