---
title: "STAC32 all slides"
author: "Ken Butler"
header-includes:
   - \usepackage{multicol}
output: 
  beamer_presentation:
    latex_engine: lualatex
    slide_level: 2
    df_print: kable
    theme: "AnnArbor"
    colortheme: "dove"
urlcolor: blue
---



```{r child="preliminaries.Rmd", echo=F}
```

# Course outline

```{r, child="outline_c32.Rmd"}
```

# Running R

```{r, child="running_R.Rmd"}
```

# Reading data files

```{r, child="readfile_R.Rmd"}
```

# Making graphs

```{r, child="graphs_R.Rmd"}
```

# Numerical summaries

```{r, child="numsum_R.Rmd"}
```

# Inference 1: one- and two-sample t

```{r, child="inference_1_R.Rmd"}
```

# The bootstrap for assessing sampling distributions (for extra insight)

```{r, child="bootstrap_R.Rmd"}
```

# Inference 2: power

```{r, child="inference_2_R.Rmd"}
```

# Inference 3: the sign test

```{r, child="inference_3_R.Rmd"}
```

# Inference 4: matched pairs, normal quantile plot

```{r, child="inference_4_R.Rmd"}
```

# Inference 5: Mood's median test, analysis of variance

```{r, child="inference_5_R.Rmd"}
```

# Writing reports

```{r, child="reports_R.Rmd"}
```

# Tidying data

```{r, child="tidying_R.Rmd"}
```

# Choosing rows and columns

```{r, child="choosing_R.Rmd"}
```

# Case study: the windmill data

```{r, child="windmill.Rmd"}
```

# Case study: the asphalt data

```{r, child="asphalt.Rmd"}
```

# Regression with categorical data

```{r, child="with_categ_R.Rmd"}
```

# Functions

```{r, child="functions.Rmd"}
```
