## Doing things with data frames
Let’s go back to our Australian athletes: 

```{r read-athletes, echo=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/ais.txt"
athletes <- read_tsv(my_url)
```

\footnotesize

```{r}
athletes
```
\normalsize

## Choosing a column

```{r}
athletes %>% select(Sport)
```

## Choosing several columns

```{r}
athletes %>% select(Sport, Hg, BMI)
```

## Choosing consecutive columns

```{r}
athletes %>% select(Sex:WCC)
```

## Choosing all-but some columns

```{r}
athletes %>% select(-(RCC:LBM))
```

## Select-helpers
Other ways to select columns: those whose name:

- `starts_with` something
- `ends_with` something
- `contains` something
- `matches` a “regular expression”
- `everything()` select all the columns

## Columns whose names begin with S 

```{r}
athletes %>% select(starts_with("S"))
```

## Columns whose names end with C

either uppercase or lowercase:

```{r}
athletes %>% select(ends_with("c"))
```

## Case-sensitive

This works with any of the select-helpers:

```{r}
athletes %>% select(ends_with("C", ignore.case=F))
```


## Column names containing letter R

```{r}
athletes %>% select(contains("r"))
```

## Exactly two characters, ending with T

In regular expression terms, this is `^.t$`:

- `^` means “start of text”
- `.` means “exactly one character, but could be anything”
- `$` means “end of text”.
```{r}
athletes %>% select(matches("^.t$"))
```

## Choosing columns by property

- Use `where` as with summarizing several columns
- eg, to choose text columns:

```{r}
athletes %>% select(where(is.character))
```


## Choosing rows by number 

```{r}
athletes %>% slice(16:25)
```




## Non-consecutive rows 

```{r}
athletes %>% 
  slice(10,13,17,42)
```

## A random sample of rows

```{r}
athletes %>% slice_sample(n=8)
```

## Rows for which something is true

\footnotesize
```{r}
athletes %>% filter(Sport == "Tennis")
```
\normalsize

## More complicated selections

```{r}
athletes %>% filter(Sport == "Tennis", RCC < 5)
```

## Another way to do "and"

```{r}
athletes %>% filter(Sport == "Tennis") %>% 
  filter(RCC < 5)
```


## Either/Or

```{r}
athletes %>% filter(Sport == "Tennis" | RCC > 5)
```

## Sorting into order

```{r}
athletes %>% arrange(RCC)
```

## Breaking ties by another variable

```{r}
athletes %>% arrange(RCC, BMI)
```

## Descending order

```{r}
athletes %>% arrange(desc(BMI))
```

## “The top ones”


```{r}
athletes %>%
  arrange(desc(Wt)) %>%
  slice(1:7) %>%
  select(Sport, Wt)
```

## Another way

```{r}
athletes %>% 
  slice_max(order_by = Wt, n=7) %>% 
  select(Sport, Wt)
```


## Create new variables from old ones

```{r new-from-old}
athletes %>%
  mutate(wt_lb = Wt * 2.2) %>%
  select(Sport, Sex, Wt, wt_lb) %>% 
  arrange(Wt)
```

## Turning the result into a number
Output is always data frame unless you explicitly turn it into something
else, eg. the weight of the heaviest athlete, as a number:

```{r to-number}
athletes %>% arrange(desc(Wt)) %>% pluck("Wt", 1)
```

Or the 20 heaviest weights in descending order:


```{r}
athletes %>%
  arrange(desc(Wt)) %>%
  slice(1:20) %>%
  pluck("Wt")
```

## Another way to do the last one

```{r}
athletes %>%
  arrange(desc(Wt)) %>%
  slice(1:20) %>%
  pull("Wt")
```

`pull` grabs the column you name *as a vector* (of whatever it contains).

## To find the mean height of the women athletes
Two ways:

\small
```{r}
athletes %>% group_by(Sex) %>% summarize(m = mean(Ht))
```

```{r}
athletes %>%
  filter(Sex == "female") %>%
  summarize(m = mean(Ht))
```

\normalsize

## Summary of data selection/arrangement "verbs"

 \begin{tabular}{lp{0.7\textwidth}}
    Verb & Purpose\\
    \hline
    \texttt{select} & Choose columns\\
    \texttt{print} & Display non-default \# of rows/columns \\
    \texttt{slice} & Choose rows by number\\
    \texttt{sample\_n} & Choose random rows\\ 
    \texttt{filter} & Choose rows satisfying conditions \\
    \texttt{arrange}& Sort in order by column(s) \\
    \texttt{mutate} & Create new variables\\
    \texttt{group\_by} & Create groups to summarize by\\
    \texttt{summarize} & Calculate summary statistics (by groups if defined)\\
    \texttt{pluck} & Extract items from data frame\\
    \texttt{pull} & Extract a single column from a data frame as a vector\\
    \hline
  \end{tabular}
  
  
## Looking things up in another data frame

```{r, echo=FALSE}
tb3 <- read_rds("tb3.rds")
```


Recall the tuberculosis data set, tidied: 
```{r}
tb3
```

What are actual names of those countries in `iso2`?

## Actual country names
Found actual country names to go with those abbreviations, in spreadsheet: 

\footnotesize
```{r}
my_url <- 
  "http://www.utsc.utoronto.ca/~butler/c32/ISOCountryCodes081507.xlsx"
```

\normalsize

Note trick for reading in `.xlsx` from URL:

```{r country-codes}
f <- tempfile()
download.file(my_url, f)
country_names <- read_excel(f)
```

- set up temporary file
- download spreadsheet to there
- read it from temporary file (which is "local")


## The country names

```{r}
country_names
```

## Looking up country codes
Matching a variable in one data frame to one in another is called a **join**
(database terminology):

```{r}
tb3 %>% left_join(country_names, by = c("iso2" = "Code_UC"))
```

## Total cases by country

```{r}
options(dplyr.summarise.inform=FALSE)
```



```{r}
tb3 %>%
  group_by(iso2) %>%
  summarize(cases = sum(freq)) %>%
  left_join(country_names, by = c("iso2" = "Code_UC")) %>%
  select(Country, cases)
```

## or even sorted in order

```{r}
tb3 %>%
  group_by(iso2) %>%
  summarize(cases = sum(freq)) %>%
  left_join(country_names, by = c("iso2" = "Code_UC")) %>%
  select(Country, cases) %>%
  arrange(desc(cases))
```

## Comments

- This is probably not quite right because of:
  - the 1994-1995 thing
  - there is at least one country in `tb3` that was not in `country_names` (the NA above). Which?
  
\footnotesize  
```{r}
tb3 %>%
  anti_join(country_names, by = c("iso2" = "Code_UC")) %>%
  distinct(iso2)
```
\normalsize


