---
title: "Drawing graphs"
---

## Our data
- To illustrate making graphs, we need some data.
- Data on 202 male and female athletes at the Australian Institute of
Sport.
- Variables:
  - categorical: Sex of athlete, sport they play
  - quantitative: height (cm), weight (kg), lean body mass, red and white
  blood cell counts, haematocrit and haemoglobin (blood), ferritin
  concentration, body mass index, percent body fat.
- Values separated by tabs (which impacts reading in).

## Packages for this section

```{r graphs-R-1}
library(tidyverse)
```


## Reading data into R
- Use `read_tsv` (“tab-separated values”), like `read_csv`.
- Data in `ais.txt`:

\scriptsize
```{r graphs-R-2}
my_url <- "http://ritsokiguess.site/datafiles/ais.txt"
athletes <- read_tsv(my_url)
```
\normalsize

## The data (some)

```{r}
#| echo: false
options(width = 72)
```


```{r graphs-R-3}
athletes
```


## Types of graph {.smaller}

Depends on number and type of variables:

|   Categorical | Quantitative | Graph |
|----:|----:|:--------------------|
    |1 | 0 | bar chart |
    |0 | 1 | histogram|
    |2 | 0 | grouped bar charts |
    |1 | 1 | side-by-side boxplots|
    |0 |2 | scatterplot|
    |2 | 1 | grouped boxplots|
    |1 | 2 | scatterplot with points identified by group (eg.\ by colour)|

With more (categorical) variables, might want *separate plots by groups*. This is called
`facetting` in R.


## `ggplot`

- R has a standard graphing procedure ggplot, that we use for all our
graphs.
- Use in different ways to get precise graph we want.
- Let’s start with bar chart of the sports played by the athletes.

## Bar chart

```{r graphs-R-4, fig.height=3.9}
ggplot(athletes, aes(x = Sport)) + geom_bar()
```


## Histogram of body mass index

```{r graphs-R-5, fig.height=3.9}
ggplot(athletes, aes(x = BMI)) + geom_histogram(bins = 10)
```

## Which sports are played by males and females?

Grouped bar chart:

```{r graphs-R-6, fig.height=3.15}
ggplot(athletes, aes(x = Sport, fill = Sex)) +
  geom_bar(position = "dodge")
```

## BMI by gender

```{r graphs-R-7, fig.height=4}
ggplot(athletes, aes(x = Sex, y = BMI)) + geom_boxplot()
```

## Height vs. weight

Scatterplot:

```{r graphs-R-8, fig.height=3.4}
ggplot(athletes, aes(x = Ht, y = Wt)) + geom_point()
```

## With regression line

```{r graphs-R-9, fig.height=3.6}
ggplot(athletes, aes(x = Ht, y = Wt)) +
  geom_point() + geom_smooth(method = "lm")
```

## BMI by sport and gender

```{r graphs-R-10, fig.height=3.6}
ggplot(athletes, aes(x = Sport, y = BMI, colour = Sex)) +
  geom_boxplot()
```

## Height and weight by gender

```{r graphs-R-11, fig.height=3.6}
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sex)) +
  geom_point()
```

## Height by weight by gender for each sport, with facets

```{r graphs-R-12, fig.height=3.6}
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sex)) +
  geom_point() + facet_wrap(~Sport)
```

## Filling each facet

Default uses same scale for each facet. To use different scales for each
facet, this:

\small
```{r graphs-R-13, fig.height=3.2}
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sex)) +
  geom_point() + facet_wrap(~Sport, scales = "free")
```
\normalsize

