## Our data
- To illustrate making graphs, we need some data.
- Data on 202 male and female athletes at the Australian Institute of
Sport.
- Variables:
  - categorical: Sex of athlete, sport they play
  - quantitative: height (cm), weight (kg), lean body mass, red and white
  blood cell counts, haematocrit and haemoglobin (blood), ferritin
  concentration, body mass index, percent body fat.
- Values separated by *tabs* (which impacts reading in).

## Packages for this section

```{r graphs-R-post-1, eval=F}
library(tidyverse)
```


## Reading data into R
- Use `read_tsv` (“tab-separated values”), like `read_csv`.
- Data in `ais.txt`:

\scriptsize
```{r graphs-R-post-2}
my_url <- "http://ritsokiguess.site/datafiles/ais.txt"
athletes <- read_tsv(my_url)
```
\normalsize

read_table?

```{r graphs-R-post-3}
read_table(my_url)
```


## The data (some)

\footnotesize
```{r graphs-R-post-4}
athletes
```
\normalsize

## Types of graph

Depends on number and type of variables:

  \begin{tabular}{ccp{0.5\textwidth}}
    Categorical & Quantitative & Graph\\
    \hline
    1 & 0 & bar chart\\
    0 & 1 & histogram\\
    2 & 0 & grouped bar charts\\
    1 & 1 & side-by-side boxplots\\
    0 & 2 & scatterplot\\
    2 & 1 & grouped boxplots\\
    1 & 2 & scatterplot with points identified by group (eg.\ by colour)\\
    \hline
  \end{tabular}

With more variables, might want *separate plots by groups*. This is called
`facetting` in R.


## `ggplot`

- R has a standard graphing procedure `ggplot`, that we use for all our
graphs.
- Use in different ways to get precise graph we want.
- Let’s start with bar chart of the sports played by the athletes.

## Bar chart

```{r graphs-R-post-5, fig.height=3.9}
ggplot(athletes, aes(x = Sport)) + geom_bar()
```


## Histogram of body mass index

```{r graphs-R-post-6, fig.height=3.9}
ggplot(athletes, aes(x = BMI)) + geom_histogram(bins = 30)
```

```{r graphs-R-post-7, fig.height=3.9}
ggplot(athletes, aes(x = BMI)) + geom_histogram(bins = 10)
```

## Which sports are played by males and females?

Grouped bar chart:

```{r graphs-R-post-8, fig.height=3.15}
ggplot(athletes, aes(x = Sport, fill = Sex)) +
  geom_bar(position = "dodge")
```

## BMI by gender

```{r graphs-R-post-9, fig.height=4}
ggplot(athletes, aes(x = Sex, y = BMI)) + geom_boxplot()
```

## Height vs. weight

Scatterplot:

```{r graphs-R-post-10, fig.height=3.4}
ggplot(athletes, aes(x = Ht, y = Wt)) + geom_point()
```

## With regression line

```{r graphs-R-post-11, fig.height=3.6}
ggplot(athletes, aes(x = Ht, y = Wt)) +
  geom_point() + geom_smooth(method = "lm")
```

## BMI by sport and gender

```{r graphs-R-post-12, fig.height=3.6}
ggplot(athletes, aes(x = Sport, y = BMI, colour = Sex)) +
  geom_boxplot()
```

## Height and weight by gender

```{r graphs-R-post-13, fig.height=3.6}
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sex)) +
  geom_point()
```


or

```{r graphs-R-post-14}
ggplot(athletes, aes(x = Ht, y = Wt)) +
  geom_point() + facet_wrap(~Sex)
```


## Height by weight for each sport and gender, with facets

```{r graphs-R-post-15, fig.height=3.6}
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sex)) +
  geom_point() + facet_wrap(~Sport)
```

## Filling each facet

Default uses same scale for each facet. To use different scales for each
facet, this:

\small
```{r graphs-R-post-16, fig.height=3.2}
ggplot(athletes, aes(x = Ht, y = Wt, colour = Sex)) +
  geom_point() + facet_wrap(~Sport, scales = "free")
```
\normalsize
