---
title: "logistic regression"
author: "Ken"
date: "1/14/2019"
output: html_document
---

```{r}
library(tidyverse)
library(broom)
```

other rat data

```{r}
my_url="notes/rat2.txt"
rat2=read_delim(my_url," ")
rat2
```

same 5 doses, but more rats (10 at each dose), summarized onto one line per dose (#lived, #died)

To model, create two-column response, thing you model (living) in 1st:

```{r}
response=with(rat2,cbind(lived,died))
class(response)
```

now fit model:

```{r}
rat2.1=glm(response~dose,family="binomial",
  data=rat2)
summary(rat2.1)
```

there is now a significant effect of dose. Negative slope: prob of living goes *down* as dose increases.

Predictions:

```{r}
p=predict(rat2.1,type="response")
cbind(rat2,p)
```

graph with CIs again

## sepsis data

slide 105

```{r}
my_url="sepsis.txt"
sepsis=read_delim(my_url," ")
sepsis
```

Predict from everything:

```{r}
sepsis.1=glm(death~shock+malnut+alcohol+age+
              bowelinf,family="binomial",
	      data=sepsis)
summary(sepsis.1)
```

or

```{r}
tidy(sepsis.1)
```


Take out `malnut`:

```{r}
sepsis.2=update(sepsis.1,.~.-malnut)
tidy(sepsis.2)
```

Nothing else can be removed.

Predictions:

```{r}
sepsis.pred=predict(sepsis.2,type="response")
d=data.frame(sepsis,sepsis.pred)
myrows=c(4,1,2,11,32) ; slice(d,myrows)
```




## ----virtusentella,fig.height=3.4, warning=F-----------------------------
ggplot(augment(sepsis.2),aes(x=age,y=.resid))+
  geom_point()

## ----size="small"--------------------------------------------------------
sepsis.2.tidy=tidy(sepsis.2)
sepsis.2.tidy

## ----size="small"--------------------------------------------------------
cc=exp(sepsis.2.tidy$estimate)
data.frame(sepsis.2.tidy$term,expcoeff=round(cc,2))
