---
title: "Drawing maps with R"
format: revealjs
execute:
  echo: true
editor: 
  markdown: 
    wrap: 72
---


## Making maps in R

- Spatial data comes with locations (perhaps with information about those locations).
- A good way to draw spatial data is on a map.
- The leaflet package is the easiest way to draw maps in R.
- Install these two packages, with two familiar ones:

\small
```{r mapping-1}
#| message = FALSE
library(tmaptools)
library(leaflet)
library(tidyverse)
library(conflicted)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::arrange)
```
\normalsize

## Hockey league map

The Ontario hockey divisions (the last example for cluster analysis) came with a very bad map. Can we do better?

- reload the Ontario road distances

\small
```{r mapping-2}
my_url <-
  "http://ritsokiguess.site/datafiles/ontario-road-distances.csv"
# my_url <- "ontario-road-distances.csv"
ontario <- read_csv(my_url)
```
\normalsize

## Ontario road distances (some)

```{r}
#| echo: false
options(width = 72)
```


```{r mapping-3}
ontario
```


## Grab the places

- and append province ("ON") for reasons shortly to become clear:

\small
```{r mapping-4}
tibble(place = ontario$place) %>% 
  mutate(prov = "ON") %>% 
  unite(place1, c(place, prov), sep = " ") -> ontario2
ontario2
```
\normalsize

## Geocode 1/2

- find their latitudes and longitudes ("geocode"; slow). 
- Save the geocoded places.

```{r mapping-5}
#| eval = FALSE
ontario2 %>% 
  rowwise() %>% 
  mutate(ll = list(geocode_OSM(place1))) -> d
```

```{r}
#| echo = FALSE
# write_rds(d, "d.rds")
d <- read_rds("d.rds")
```

\small
```{r}
d
```
\normalsize

## Geocode 2/2

Untangle the lats and longs:

```{r mapping-6}
d %>% 
  unnest_wider(ll) %>% 
  unnest_wider(coords) -> ontario3
ontario3
```

## Make map

- finally:


```{r mapping-7}
leaflet(data = ontario3) %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~x, lat = ~y) 
```
 



## Cluster analysis revisited

```{r mapping-8}
ontario %>% select(-1) %>% as.dist() -> ontario.d
ontario.hc <- hclust(ontario.d, method = "ward.D")
```

## Seven clusters:

```{r mapping-9, fig.height=4 }
plot(ontario.hc)
rect.hclust(ontario.hc, 7)
```

## Get the clusters

\footnotesize
```{r mapping-10}
tibble(place = ontario$place, cluster = cutree(ontario.hc, 7)) -> clusters
clusters %>% arrange(cluster)
```

\normalsize

## Combine clusters

- combine clusters 6 and 7 with 4 ("north")
- combine clusters 2 and 3 ("east")
- make named divisions

```{r mapping-11}
clusters %>% 
  mutate(division = fct_collapse(factor(cluster),
                                "north" = c("4", "6", "7"),
                                "east" = c("2", "3"),
                                "west" = "5",
                                "central" = "1")) %>% 
  arrange(division) -> divisions
```

## The divisions

```{r mapping-12}
divisions
```

## Take "ON" off of `ontario3`

\footnotesize
```{r mapping-13}
ontario3 %>% 
  mutate(place = str_replace(place1, " ON$", "")) -> ontario3
ontario3
```
\normalsize

## Add the divisions, matching by place

- and draw map

```{r mapping-14}
pal <- colorFactor("Set1", divisions$division)

ontario3 %>% left_join(divisions) %>% 
  select(place, x, y, division) %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~x, lat = ~y, 
                   color = ~pal(division)) 
```


## Original seven clusters

The same idea gets a map of the original seven clusters:

```{r mapping-16}
pal <- colorFactor("Set1", divisions$cluster)
ontario3 %>% left_join(divisions) %>% 
  select(place, x, y, cluster) %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(lng = ~x, lat = ~y, 
                   color = ~pal(cluster))
```
