## Summarizing data in R
- Have seen `summary` (5-number summary of each column). But what
if we want:
  - a summary or two of just one column
  - a count of observations in each category of a categorical variable
  - summaries by group
  - a different summary of all columns (eg. SD)
- To do this, meet pipe operator `%>%`. This takes input data frame,
does something to it, and outputs result. (Learn: `Ctrl-Shift-M`.) 
- Output from a pipe can be used as input to something else, so can
have a sequence of pipes.
- Summaries include: mean, median, min, max, sd, IQR, quantile (for
obtaining quartiles or any percentile), n (for counting observations).
- Use our Australian athletes data again.

## Packages for this section

```{r numsum-R-post-1, eval=F}
library(tidyverse)
```

```{r numsum-R-post-2, echo=F, message=F}
my_url <- "http://ritsokiguess.site/datafiles/ais.txt"
athletes <- read_tsv(my_url)
athletes
```

## Summarizing one column
- Mean height: 

```{r numsum-R-post-3}
athletes %>% summarize(m=mean(Ht))
athletes %>% summarise(m=mean(Ht))
```

(these both work)

or to get mean and SD of BMI:

```{r numsum-R-post-4}
athletes %>% summarize(mean_bmi=mean(BMI), sd_bmi=sd(BMI))
```

## Quartiles

- `quantile` calculates percentiles ("fractiles"), so we want the 25th and 75th
percentiles:

```{r numsum-R-post-5}
v <- c(0.25, 0.75)
athletes %>% summarize(q=quantile(Wt, v), prob=v)
```


## Creating new columns

- These weights are in kilograms. Maybe we want to summarize the
weights in pounds.
- Convert kg to lb by multiplying by 2.2.
- Create new column and summarize that: 

```{r numsum-R-post-6}
v
athletes %>% mutate(wt_lb=Wt*2.2) %>%
  summarize(Q_lb=quantile(wt_lb, v), prob=v)
```

## Counting how many
for example, number of athletes in
each sport: 

```{r numsum-R-post-7, echo=FALSE}
options(dplyr.summarise.inform=FALSE) 
```


```{r numsum-R-post-8}
athletes %>% count(Sport)
```

## Counting how many, variation 2:

Another way (which will make sense
in a moment):

\small
```{r numsum-R-post-9}
athletes %>% group_by(Sport) %>%
  summarize(count=n())
```
\normalsize

## Summaries by group
- Might want separate summaries for each “group”, eg. mean and SD
of height for males and females. Strategy is `group_by` (to define the
groups) and then `summarize`: 

```{r numsum-R-post-10}
athletes %>% group_by(Sex) %>% 
  summarize(m=mean(Ht), s=sd(Ht))
```

## Count plus stats

If you want number of observations per group plus some stats, you need to go the `n()` way:

```{r numsum-R-post-11}
athletes %>% group_by(Sex) %>% 
  summarize(n=n(), m=mean(Ht), s=sd(Ht))
```

- This explains second variation on counting within group: “within each
sport, how many athletes were there?”

## quantiles by group

```{r numsum-R-post-12}
v
athletes %>%
  group_by(Sex) %>% 
  summarize(Q_wt=quantile(Wt, v), prob=v)
```


## Summarizing several columns

- Standard deviation of each (numeric) column: 

\small
```{r numsum-R-post-13}
athletes %>% summarize(across(where(is.numeric), ~sd(.)))
```
\normalsize

- Median and IQR of all columns whose name starts with H: 

\small
```{r numsum-R-post-14}
athletes %>% summarize(across(starts_with("H"),
                       list(med=~median(.), iqr=~IQR(.))))
```
\normalsize

## by group

```{r numsum-R-post-15}
athletes %>% 
  group_by(Sex) %>% 
  summarize(across(starts_with("H"), 
                   list(med=~median(.), 
                        iqr=~IQR(.))))
```
