---
title: "STAC33: Introduction to Applied Statistics"
author: "Lecture notes"
header-includes:
   - \usepackage{multicol}
output: 
  beamer_presentation:
    latex_engine: lualatex
    slide_level: 2
    df_print: kable
    theme: "AnnArbor"
    colortheme: "dove"
---



```{r slides-c33-1, child="preliminaries.Rmd", echo=F}
```

# Course outline 

```{r slides-c33-2, child="outline_c33.Rmd"}
```


# Running R

```{r slides-c33-3, child="running_R.Rmd"}
```

# Reading in data

```{r slides-c33-4, child="readfile_R.Rmd"}
```

# Making graphs

```{r slides-c33-5, child="graphs_R.Rmd"}
```

# Numerical summaries: more detailed

```{r slides-c33-6, child="numsum_R.Rmd"}
```

# Statistical Inference

```{r slides-c33-7, child="inference_R.Rmd"}
```

# Writing reports

```{r slides-c33-8, child="reports_R.Rmd"}
```


# Tidying and organizing data

```{r slides-c33-9, child="tidying_R.Rmd"}
```

# Case study 1: the windmill data

```{r slides-c33-10, child="windmill.Rmd"}
```


# Case study 3: Asphalt

```{r slides-c33-11, child="asphalt.Rmd"}
```



# Regression with categorical variables

```{r slides-c33-12, child="with_categ_R.Rmd"}
```

# Functions

```{r slides-c33-13, child="functions.Rmd"}
```

# Dates and times

```{r slides-c33-14, child="dates_and_times.Rmd"}
```

# Miscellanea

```{r slides-c33-15, child="miscellanea.Rmd"}
```


# Vector and matrix algebra (for C33)

```{r slides-c33-16, child="vector_matrix.Rmd"}
```

# The bootstrap

```{r slides-c33-17, child="bootstrap.Rmd"}
```


# Bayesian statistics with `rstan` 

```{r slides-c33-18, child="rstan.Rmd"}
```


# the end of the notes

