## Packages for this section

```{r with-categ-R-post-1, eval=F}
library(tidyverse)
library(broom)
```


## The pigs revisited

```{r with-categ-R-post-2, echo=FALSE}
options(dplyr.summarise.inform = FALSE)
```

- Recall pig feed data, after we tidied it: 

```{r with-categ-R-post-3, message=F}
my_url <- "http://ritsokiguess.site/datafiles/pigs2.txt"
pigs <- read_delim(my_url, " ")
pigs 
```

## Summaries

```{r with-categ-R-post-4}
pigs %>%
  group_by(feed) %>%
  summarize(n = n(), mean_wt = mean(weight), 
            sd_wt = sd(weight))
```

## Running through `aov` and `lm`
- What happens if we run this through `lm` rather than `aov`? 
- Recall `aov` first: 

```{r with-categ-R-post-5}
pigs.1 <- aov(weight ~ feed, data = pigs)
summary(pigs.1)
```

## and now `lm`

\footnotesize
```{r with-categ-R-post-6}
pigs.2 <- lm(weight ~ feed, data = pigs)
summary(pigs.2)
tidy(pigs.2)
glance(pigs.2)
```
\normalsize

## Understanding those slopes
- Get one slope for each category of categorical variable feed, except
for first.
- feed1 treated as “baseline”, others measured relative to that.
- Thus prediction for feed 1 is intercept, 60.62 (mean weight for feed 1).
- Prediction for feed 2 is 60.62 + 8.68 = 69.30 (mean weight for feed 2).
- Or, mean weight for feed 2 is 8.68 bigger than for feed 1.
- Mean weight for feed 3 is 33.48 bigger than for feed 1.
- Slopes can be negative, if mean for a feed had been smaller than for
feed 1.

## Reproducing the ANOVA
- Pass the fitted model object into `anova`:

\footnotesize
```{r with-categ-R-post-7}
anova(pigs.2)
```
\normalsize

- Same as before.
- But no Tukey this way:

\footnotesize
```{r with-categ-R-post-8, error=TRUE}
TukeyHSD(pigs.2)
```
\normalsize

## The crickets
- Male crickets rub their wings together to produce a chirping sound.
- Rate of chirping, called “pulse rate”, depends on species and possibly
on temperature.
- Sample of crickets of two species’ pulse rates measured; temperature
also recorded.
- Does pulse rate differ for species, especially when temperature
accounted for?

## The crickets data
Read the data:

\small
```{r with-categ-R-post-9, message=F}
my_url <- "http://ritsokiguess.site/datafiles/crickets2.csv"
crickets <- read_csv(my_url)
crickets 
```
\normalsize
## Fit model with `lm` 

```{r with-categ-R-post-10}
crickets.1 <- lm(pulse_rate ~ temperature + species, 
                 data = crickets)
summary(crickets.1)
```

Can I remove anything? No:

```{r with-categ-R-post-11}
drop1(crickets.1, test = "F")
```

`drop1` is right thing to use in a regression with categorical (explanatory) variables in it: "can I remove this categorical variable *as a whole*?"

## The summary 

```{r with-categ-R-post-12}
glance(crickets.1)
tidy(crickets.1)
```

## Conclusions
- Slope for temperature says that increasing temperature by 1 degree
increases pulse rate by 3.6 (same for both species)
- Slope for `speciesniveus` says that pulse rate for niveus about 10
lower than that for exclamationis at same temperature (latter species
is baseline).
- R-squared of almost 0.99 is very high, so that the prediction of pulse
rate from species and temperature is very good.

## To end with a graph

- Two quantitative variables and one categorical: scatterplot with categories distinguished by colour.
- This graph seems to need a title, which I define first. 

\small

```{r with-categ-R-post-13}
t1 <- "Pulse rate against temperature for two species of crickets"
t2 <- "Temperature in degrees Celsius"
```

\normalsize

```{r spectemp, fig.keep="none", warning=F, message=F}
ggplot(crickets, aes(x = temperature, y = pulse_rate,
  colour = species)) +
  geom_point() + geom_smooth(method = "lm", se = F) +
  ggtitle(t1, t2)
```

## The graph

```{r thingy, ref.label="spectemp", echo=FALSE, message=FALSE}
```


interaction

```{r with-categ-R-post-14}
crickets.2 <- lm(pulse_rate ~ temperature * species, 
                 data = crickets)
summary(crickets.2)
```

```{r with-categ-R-post-15}
drop1(crickets.2, test = "F")
```

```{r with-categ-R-post-16}
crickets.3 <- update(crickets.2, .~. - temperature:species)
drop1(crickets.3, test="F")
```

