
## Repeated measures by profile analysis 


* More than one response *measurement* for each subject. Might be


* measurements of the same thing at different times

* measurements of different but related things


* Generalization of matched pairs ("matched triples", etc.).

* Variation: each subject does several different treatments at different times (called *crossover design*).

* Expect measurements on same subject to be correlated, so
assumptions of independence will fail.

* Called *repeated measures*. Different approaches, but *profile analysis* uses `Manova` (set up right way).

* Another approach uses *mixed models* (random effects).



## Packages
```{r, eval=F}
library(car)
library(tidyverse)
```

   

## Example: histamine in dogs


* 8 dogs take part in experiment.

* Dogs randomized to one of 2 different drugs.

* Response: log of blood concentration of histamine 0, 1, 3 and 5 minutes after taking drug. (Repeated measures.)

* Data in `dogs.txt`, column-aligned.




## Read in data
```{r }
my_url <- "http://www.utsc.utoronto.ca/~butler/d29/dogs.txt"
dogs <- read_table(my_url)
```

   


## Setting things up
```{r size="small"}
dogs
response <- with(dogs, cbind(lh0, lh1, lh3, lh5))
dogs.1 <- lm(response ~ drug, data = dogs)
```
 


## The repeated measures MANOVA

Get list of response variable names; we call them `times`. Save
in data frame. 

```{r, echo=FALSE}
options(width = 70)
```


\footnotesize 
```{r, error=TRUE}
times <- colnames(response)
times.df <- data.frame(times=factor(times))
dogs.2 <- Manova(dogs.1,
  idata = times.df,
  idesign = ~times
)
dogs.2
```
 
\normalsize



## Wide and long format

- Interaction significant. Pattern of response over time different
for the two drugs.


* Want to investigate interaction.

## The wrong shape

* But data frame has several observations per line ("wide format"):

\scriptsize
```{r }
dogs %>% slice(1:6)
```
 
\normalsize

* Plotting works with data in "long format":
one response per line.

* The responses are log-histamine at different times, labelled
`lh`-something. Call them all `lh` and put them in
one column, with the time they belong to labelled.


## Running `pivot_longer`, try 1

\footnotesize

```{r size="footnotesize"}
dogs %>% pivot_longer(starts_with("lh"), 
                      names_to = "time", values_to = "lh") 
```

\normalsize

   


## Getting the times
Not quite right: for the times, we want just the numbers, not the
letters `lh` every time. Want new variable
containing just number in `time`:
`parse_number`. 

\footnotesize

```{r}
dogs %>%
  pivot_longer(starts_with("lh"), 
               names_to = "timex", values_to = "lh") %>% 
  mutate(time = parse_number(timex)) 
```

\normalsize



## What I did differently


* I realized that `pivot_longer` was going to produce something
like `lh1`, which I needed to do something further with, so
this time I gave it a temporary name `timex`.

* This enabled me to use the name `time` for the actual
numeric time.

* This works now, so next save into a new data frame `dogs.long`.



## Saving the pipelined results
```{r }
dogs %>%
  pivot_longer(starts_with("lh"), 
               names_to = "timex", values_to = "lh") %>% 
  mutate(time = parse_number(timex)) -> dogs.long
```

 

This says:



* Take data frame dogs, and then:

* Combine the columns `lh0` through `lh5` into one
column called `lh`, with the column that each `lh`
value originally came from labelled by `timex`, and then:

* Pull out numeric values in `timex`, saving in `time` and then:

* save the result in a data frame `dogs.long`.



## Interaction plot

\small
```{r fig.height=4}
ggplot(dogs.long, aes(x = time, y = lh, 
                      colour = drug, group = drug)) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun = mean, geom = "line")
```
\normalsize
   

## Comments


* Plot mean `lh` value at each time, joining points on same
drug by lines.

* drugs same at time 0

* after that, Trimethaphan higher than Morphine.

* Effect of drug not consistent over time: significant interaction.



## Take out time zero


* Lines on interaction plot would then be parallel, and so interaction should
no longer be significant.

* Go back to original "wide" `dogs` data frame.

```{r size="footnotesize", error=TRUE}
response <- with(dogs, cbind(lh1, lh3, lh5)) # excl time 0
dogs.1 <- lm(response ~ drug, data = dogs)
times <- colnames(response)
times.df <- data.frame(times=factor(times))
dogs.2 <- Manova(dogs.1,
  idata = times.df,
  idesign = ~times
)
```
 


## Results and comments

\footnotesize
```{r, error=TRUE}
dogs.2
```
 
\normalsize




* Correct: interaction no longer significant.

* Significant effect of time.

* Drug effect not quite significant (some variety among dogs
within drug).



## Is the non-significant drug effect reasonable?


* Plot *actual data*: `lh` against `days`,
labelling observations by drug: "spaghetti plot".

* Uses long data frame (confusing, yes I know):

* Plot (time,lh) points coloured  by drug

* and connecting measurements for each *dog* by lines.

* This time, we want `group=dog` (want the measurements for each
*dog* joined by lines), but `colour=drug`:
```{r platanias}
g <- ggplot(dogs.long, aes(
  x = time, y = lh,
  colour = drug, group = dog
)) +
  geom_point() + geom_line()
```

   



## The spaghetti plot
```{r hoverla,fig.height=5}
g
```

   


## Comments


* For each dog over time, there is a strong increase and gradual
decrease in log-histamine. This
explains the significant time effect.

* The pattern is more or less the same for each dog, regardless
of drug. This explains the non-significant interaction.

* Most of the trimethaphan dogs (blue) have higher log-histamine
throughout (time 1 and after), and some of the morphine dogs have
lower.

* *But* two of the morphine dogs have log-histamine
profiles like the trimethaphan dogs. This ambiguity is probably
why the `drug` effect is not quite significant.


## Mixed models

- Another way to fit repeated measures
- Subjects (on whom repeated measures taken) are *random sample of all possible subjects* (random effects)
- Times and treatments are *the only ones we care about* (fixed effects)
- Use package `lme4` function `lmer` (like `lm` in some ways)
- Uses long-format "tidy" data

## Fitting the model

```{r, message=FALSE}
library(lme4)
# dogs.long
dogs.2=lmer(lh~drug*time+(1|dog),data=dogs.long)
```

- note specification of random effect: each dog has "random intercept" that moves log-histamine up or down for that dog over all times


## What can we drop?

- using `drop1`:

```{r}
drop1(dogs.2,test="Chisq")
```

- Interaction again not significant, but P-value smaller than before

## Re-fit without interaction

```{r}
dogs.3=update(dogs.2,.~.-drug:time)
drop1(dogs.3,test="Chisq")
```

- This time neither drug nor (surprisingly) time is significant.
- MANOVA and `lmer` methods won't agree, but both valid ways to approach problem.


## The exercise data


* 30 people took part in an exercise study.

* Each subject was
randomly assigned to one of two diets ("low fat" or ``non-low
fat'') and to one of three exercise programs ("at rest",
"walking", "running").

* There are $2\times3 = 6$ experimental treatments, and thus
each one is replicated $30/6=5$ times.

* Nothing unusual so far.

* However, each subject had their pulse rate measured at three
different times (1, 15 and 30 minutes after starting their
exercise), so have repeated measures.



## Reading the data
Separated by *tabs*:  
```{r }
url <- "http://www.utsc.utoronto.ca/~butler/d29/exercise.txt"
exercise.long <- read_tsv(url)
```

   


## The data

\footnotesize
```{r}
exercise.long %>% slice(1:8)
```
\normalsize
   



* This is "long format", which is usually what we want.

* But for repeated measures analysis, we want *wide* format!

* `pivot_wider`.



## Making wide format


* `pivot_wider` needs: a column that is
going to be split, and the column to make the values out of:

\footnotesize
```{r}
exercise.long %>% pivot_wider(names_from=time, 
                              values_from=pulse) -> exercise.wide
exercise.wide %>% sample_n(5)
```
\normalsize


* Normally `pivot_longer` \texttt{min01, min15,
min30} into one column called `pulse` labelled by the
number of minutes. But `Manova` needs it the other way.



## Setting up the repeated-measures analysis


* Make a response variable consisting of `min01, min15, min30`:

\small
```{r }
response <- with(exercise.wide, cbind(min01, min15, min30))
```
\normalsize


* Predict that from `diet` and `exertype` and
interaction using `lm`:

\small
```{r }
exercise.1 <- lm(response ~ diet * exertype,
  data = exercise.wide
)
```
\normalsize
   



* Run this through `Manova`: 

\small
```{r, error=TRUE}
times <- colnames(response)
times.df <- data.frame(times=factor(times))
exercise.2 <- Manova(exercise.1, 
                     idata = times.df, 
                     idesign = ~times)
```
\normalsize
   



## Results

```{r, echo=F}
options(width=70)
```

\scriptsize
```{r }
exercise.2
```
\normalsize



* Three-way interaction significant, so cannot remove anything.

* Pulse rate depends on diet and exercise type *combination*,
and *that* is different for each time.



## Making some graphs


* Three-way  interactions are difficult to understand. To make
an attempt, look at some graphs.

* Plot time trace of pulse rates for each individual, joined by
lines, and make *separate* plots for each
`diet-exertype` combo.

* `ggplot` again. Using *long* data frame:

```{r }
g <- ggplot(exercise.long, aes(
  x = time, y = pulse,
  group = id
)) + geom_point() + geom_line() +
  facet_grid(diet ~ exertype)
```

 


* `facet_grid(diet~exertype)`: do a separate plot for each
combination of diet and exercise type, with diets going down the
page and exercise types going across. (Graphs are usually landscape,
so have the factor `exertype` with more levels going across.)



## The graph(s)
```{r fig.height=5}
g
```

   


## Comments on graphs


* For subjects who were at rest, no change in pulse rate over
time, for both diet groups.

* For walking subjects, not much change in pulse rates over
time. Maybe a small increase on average between 1 and 15 minutes.

* For both running groups, an overall increase in pulse rate
over time, but the increase is stronger for the `lowfat`
group.

* No consistent effect of diet over all exercise groups.

* No consistent effect of exercise type over both diet groups.

* No consistent effect of time over all diet-exercise type combos.



## "Simple effects" of diet for the subjects who ran


* Looks as if there is only any substantial time effect for the
runners. For them, does diet have an effect?

* Pull out only the runners from the wide data:
```{r }
exercise.wide %>%
  filter(exertype == "running") -> runners.wide
```

     

* Create response variable and do MANOVA. Some of this looks like
before, but I have different data now:

\footnotesize
```{r}
response <- with(runners.wide, cbind(min01, min15, min30))
runners.1 <- lm(response ~ diet, data = runners.wide)
times <- colnames(response)
times.df <- data.frame(times=factor(times))
runners.2 <- Manova(runners.1,
  idata = times.df,
  idesign = ~times
)
```
\normalsize
   



## Results



\scriptsize
```{r }
runners.2
```
\normalsize
   
text under 

* The `diet` by `time` interaction is still
significant (at $\alpha=0.05$): the effect of time on pulse rates is different for
the two diets.

* At $\alpha=0.01$, the interaction is not significant, and then
we have only two (very) significant main effects of `diet`
and `time`. 



## How is the effect of diet different over time?


* Table of means. Only I need long data for this, so make it (in
a pipeline):
```{r }
runners.wide %>%
  pivot_longer(starts_with("min"), 
               names_to = "time", values_to = "pulse") %>%
  group_by(time, diet) %>%
  summarize(
    mean = mean(pulse),
    sd = sd(pulse)
  ) -> summ
```

 


* Result of `summarize` is data frame, so can save it (and
do more with it if needed).



## Understanding diet-time interaction


* The summary:

\footnotesize
```{r}
summ
```
\normalsize
   

* Pulse rates at any given time higher for `lowfat` (diet
effect),

* Pulse rates increase over time of exercise (time effect),

* but the *amount by which pulse rate higher* for a diet depends on
time: `diet` by `time` interaction.


## Interaction plot


* We went to trouble of finding means by group, so making
interaction plot is now mainly easy:
```{r fig.height=2.7}
ggplot(summ, aes(x = time, y = mean, colour = diet,
                 group = diet)) + geom_point() + geom_line()
```

   
## Comment on interaction plot

* The lines are not parallel, so there is interaction between diet
and time for the runners.
- The effect of time on pulse rate is different for the two diets, even though all the subjects here were running.



   



